---
title: "{{PROJECT_TITLE}}"
author:
  - {{AUTHOR_NAME}} {{AUTHOR_LAST}}:
      correspondence: true
      email: {{AUTHOR_EMAIL}}
      orcid: {{AUTHOR_ORCID}}
      institute:
        - affil1
institute:
  - affil1: {{INSTITUTION}}
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    keep_tex: false
    toc: false
    number_sections: true
  bookdown::html_document2:
    toc: true
    toc_float: true
    number_sections: true
  bookdown::word_document2:
    toc: false
    number_sections: true
bibliography: references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
abstract: |
  Brief summary of research question, methods, and key findings.
keywords: "reproducible research, data analysis, R"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../figures/",
  fig.align = "center",
  fig.retina = 2,
  out.width = "100%"
)

# Load required packages
library(here)
library(knitr)
library(dplyr)
library(ggplot2)

# Source custom functions
source(here("R", "utils.R"))

# Set seed for reproducibility
set.seed(42)
```

# Introduction

Describe the research question, background, and motivation.

- What problem are you addressing?
- Why is it important?
- What gap in knowledge does this fill?

Research compendia enable reproducible computational research by packaging
data, code, and documentation together [@Marwick2018].

# Methods

## Data

Describe your data sources, collection methods, and any preprocessing steps.

```{r data-summary}
# Example: Load and summarize data
# data <- read.csv(here("analysis", "data", "derived_data", "cleaned_data.csv"))
#
# kable(
#   head(data),
#   caption = "Summary of dataset"
# )
```

## Analysis

Describe your analytical approach.

```{r analysis}
# Example analysis code
# model <- lm(outcome ~ predictor, data = data)
# summary(model)
```

# Results

Present your findings with appropriate visualizations and tables.

```{r figure-example, fig.cap="Example figure demonstrating results"}
# Example plot
# ggplot(data, aes(x = predictor, y = outcome)) +
#   geom_point() +
#   geom_smooth(method = "lm") +
#   theme_minimal() +
#   labs(
#     x = "Predictor Variable",
#     y = "Outcome Variable"
#   )
```

## Key Findings

Summarize the main results:

1. First major finding
2. Second major finding
3. Third major finding

# Discussion

Interpret your results in the context of existing literature.

- What do these findings mean?
- How do they relate to previous work?
- What are the limitations?
- What are the implications?

# Conclusion

Summarize the key takeaways and suggest future directions.

# Computational Environment

This analysis was conducted in the following computational environment:

```{r session-info}
sessionInfo()
```

# Data and Code Availability

All data, code, and materials necessary to reproduce these analyses are available
at: https://github.com/{{GITHUB_USER}}/{{PROJECT_NAME}}

# References

::: {#refs}
:::

# Acknowledgments

Acknowledge funding sources, collaborators, and contributors.
