---
title: "Getting Started with {{PACKAGE_NAME}}"
author: "{{AUTHOR_NAME}} {{AUTHOR_LAST}}"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with {{PACKAGE_NAME}}}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)

# Load the package
library({{PACKAGE_NAME}})
```

# Introduction

Welcome to the {{PACKAGE_NAME}} package! This vignette provides a comprehensive introduction to the package's main functionality and demonstrates how to use its key features.

## What is {{PACKAGE_NAME}}?

{{PACKAGE_NAME}} is an R package designed to [describe the main purpose and goals of your package]. It provides tools for [list main capabilities].

Key features include:

- **Feature 1**: [Brief description]
- **Feature 2**: [Brief description]
- **Feature 3**: [Brief description]
- **Feature 4**: [Brief description]

## Installation

You can install {{PACKAGE_NAME}} from GitHub using the `remotes` package:

```{r installation, eval = FALSE}
# Install from GitHub
remotes::install_github("{{GITHUB_ACCOUNT}}/{{PACKAGE_NAME}}")

# Load the package
library({{PACKAGE_NAME}})
```

# Basic Usage

## Getting Started with Core Functions

Let's start with the basic functions provided by {{PACKAGE_NAME}}.

### Hello World Function

The `hello_world()` function demonstrates basic package functionality:

```{r hello-world}
# Basic greeting
hello_world()

# Customize the greeting
hello_world("R User")

# Modify greeting options
hello_world("Developer", exclaim = FALSE)
hello_world("coder", capitalize = TRUE)
```

### Mathematical Operations

The package includes functions for mathematical operations with enhanced validation:

```{r math-operations}
# Basic addition
result1 <- add_numbers(5, 3)
result1

# Vector operations
numbers1 <- c(1, 2, 3)
numbers2 <- c(4, 5, 6)
result2 <- add_numbers(numbers1, numbers2)
result2

# Handling missing values
numbers_with_na <- c(1, NA, 3)
result3 <- add_numbers(numbers_with_na, c(2, 3, 4), na.rm = TRUE)
result3
```

## Working with Data

The package provides tools for data analysis and summarization.

### Data Summarization

```{r data-summary}
# Create sample data
sample_data <- data.frame(
  height = c(170, 165, 180, 175, 160),
  weight = c(70, 65, 85, 78, 58),
  age = c(25, 30, 35, 28, 22),
  category = c("A", "B", "A", "C", "B")
)

# Basic summary
summary_result <- summarize_data(sample_data)
print(summary_result)

# Custom statistics
custom_summary <- summarize_data(
  sample_data,
  columns = c("height", "weight"),
  stats = c("mean", "sd", "n")
)
print(custom_summary)
```

### Data Quality Assessment

```{r data-quality}
# Create data with missing values
messy_data <- data.frame(
  x = c(1, 2, NA, 4, 5),
  y = c(10, NA, 30, 40, 50),
  z = c("a", "b", "c", "d", "e")
)

# Comprehensive summary including missing values
quality_summary <- summarize_data(
  messy_data,
  stats = c("mean", "n", "missing")
)
print(quality_summary)
```

# Advanced Features

## Function Families

{{PACKAGE_NAME}} organizes related functions into families for consistent APIs and behavior.

### Greeting Functions Family

Both `hello_world()` and `goodbye_world()` share the same parameter structure:

```{r greeting-family}
# Consistent API across greeting functions
hello_world("Alice", exclaim = FALSE, capitalize = TRUE)
goodbye_world("Alice", exclaim = FALSE, capitalize = TRUE)
```

### Data Analysis Functions Family

Data analysis functions share common patterns for handling missing values and validation:

```{r data-analysis-family}
# Functions with consistent missing value handling
test_data <- c(1, 2, NA, 4, 5)

# Different functions, consistent na.rm parameter
add_numbers(test_data, c(1, 1, 1, 1, 1), na.rm = TRUE)
```

## Error Handling and Validation

{{PACKAGE_NAME}} provides informative error messages and comprehensive input validation:

```{r error-handling, error = TRUE}
# Demonstrate error handling
try(hello_world(123))  # Invalid input type
try(add_numbers("not", "numbers"))  # Invalid input types
try(summarize_data(c(1, 2, 3)))  # Invalid data structure
```

## Customization Options

Many functions provide customization options for different use cases:

```{r customization}
# Flexible data summarization
data(mtcars)

# Basic summary of all numeric variables
basic_summary <- summarize_data(mtcars)
head(basic_summary, 3)

# High-precision custom summary
precise_summary <- summarize_data(
  mtcars,
  columns = c("mpg", "hp", "wt"),
  stats = c("mean", "median", "sd"),
  round_digits = 4
)
print(precise_summary)
```

# Best Practices

## Reproducible Analysis

When using {{PACKAGE_NAME}} for analysis, follow these best practices:

```{r best-practices}
# Set seed for reproducible results
set.seed(123)

# Create reproducible sample data
n <- 100
analysis_data <- data.frame(
  variable1 = rnorm(n, mean = 50, sd = 10),
  variable2 = rnorm(n, mean = 100, sd = 15),
  group = sample(c("A", "B", "C"), n, replace = TRUE)
)

# Document your analysis parameters
analysis_params <- list(
  seed = 123,
  n_observations = n,
  analysis_date = Sys.Date()
)

# Perform analysis
results <- summarize_data(analysis_data, round_digits = 3)
print(results)
```

## Error-Safe Workflows

Use the package's validation features to create robust analysis workflows:

```{r robust-workflows}
# Function to safely analyze data
safe_analysis <- function(data, required_columns) {
  # Check data structure
  if (!is.data.frame(data)) {
    stop("Input must be a data.frame")
  }

  # Check required columns exist
  missing_cols <- setdiff(required_columns, names(data))
  if (length(missing_cols) > 0) {
    stop("Missing required columns: ", paste(missing_cols, collapse = ", "))
  }

  # Perform analysis using package functions
  results <- summarize_data(
    data,
    columns = required_columns,
    stats = c("mean", "sd", "n", "missing")
  )

  return(results)
}

# Use the safe analysis function
safe_results <- safe_analysis(
  analysis_data,
  required_columns = c("variable1", "variable2")
)
print(safe_results)
```

# Package Integration

## Working with Other Packages

{{PACKAGE_NAME}} is designed to work well with other R packages commonly used for data analysis:

```{r integration, message = FALSE}
# Integration with dplyr (if available)
if (requireNamespace("dplyr", quietly = TRUE)) {
  library(dplyr)

  # Use with dplyr workflows
  processed_data <- analysis_data %>%
    filter(variable1 > 45) %>%
    select(variable1, variable2)

  # Analyze processed data
  integrated_results <- summarize_data(processed_data)
  print(integrated_results)
}
```

## Extending Package Functionality

You can build upon {{PACKAGE_NAME}} functions for custom analyses:

```{r extension}
# Custom function building on package functionality
enhanced_summary <- function(data, ...) {
  # Get basic summary
  basic <- summarize_data(data, ...)

  # Add custom metrics
  basic$range <- basic$max - basic$min
  basic$cv <- basic$sd / basic$mean  # Coefficient of variation

  return(basic)
}

# Use enhanced function
enhanced_results <- enhanced_summary(analysis_data)
print(enhanced_results)
```

# Getting Help

## Documentation

- Use `help(package = "{{PACKAGE_NAME}}")` to see all available functions
- Use `?function_name` to get help on specific functions
- Use `example(function_name)` to run function examples

## Common Issues

### Installation Issues

If you encounter installation issues:

1. Ensure you have the latest version of R
2. Update your packages: `update.packages()`
3. Check your R configuration: `sessionInfo()`

### Function Errors

If functions return unexpected results:

1. Check input data types and structure
2. Read function documentation carefully
3. Look at function examples: `example(function_name)`

## Support

For additional support:

- **Issues**: Report bugs at https://github.com/{{GITHUB_ACCOUNT}}/{{PACKAGE_NAME}}/issues
- **Documentation**: Visit the package website (if available)
- **Contact**: Email {{AUTHOR_EMAIL}}

# Summary

This vignette introduced the core functionality of {{PACKAGE_NAME}}, including:

- Basic function usage and parameters
- Data analysis and summarization capabilities
- Error handling and validation features
- Best practices for reproducible analysis
- Integration with other R packages

For more advanced usage, see the additional vignettes:

- **Advanced Features**: [Description of advanced vignette]
- **Case Studies**: [Description of case studies vignette]
- **API Reference**: Complete function reference

# Session Information

```{r session-info}
sessionInfo()
```