---
title: "{{MANUSCRIPT_TITLE}}"
author:
  - name: "{{AUTHOR_NAME}} {{AUTHOR_LAST}}"
    email: "{{AUTHOR_EMAIL}}"
    affiliation: "{{AUTHOR_INSTITUTE}}"
    orcid: "{{AUTHOR_ORCID}}"
date: "`r Sys.Date()`"
output:
  bookdown::word_document2:
    reference_docx: "templates/manuscript_template.docx"
    number_sections: true
  bookdown::pdf_document2:
    latex_engine: xelatex
    number_sections: true
  bookdown::html_document2:
    number_sections: true
    toc: true
    toc_float: true
bibliography: references.bib
csl: "templates/journal-style.csl"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

# Load required packages and functions
library({{PACKAGE_NAME}})
library(dplyr)
library(ggplot2)
library(knitr)
library(here)

# Source analysis functions
devtools::load_all()
```

# Abstract

This research compendium presents...

**Keywords:** research reproducibility, computational analysis, academic publishing

# Introduction

This manuscript demonstrates a reproducible research workflow using the {{PACKAGE_NAME}} research compendium.

# Methods

## Data Sources

The analysis uses data stored in the `data/raw_data/` directory. Data processing functions are implemented in the R package component of this research compendium.

```{r data-processing}
# Load and process data using package functions
# processed_data <- process_raw_data()
```

## Statistical Analysis

Statistical analyses are implemented as documented functions in the `R/` directory and can be reproduced using the scripts in `analysis/reproduce/`.

```{r analysis-example}
# Example analysis using package functions
# results <- perform_statistical_analysis(processed_data)
```

# Results

## Primary Findings

```{r results-figure, fig.cap="Main analysis results"}
# Generate figure using package functions
# create_results_figure(results)
```

```{r results-table}
# Generate table using package functions
# kable(create_results_table(results), caption = "Summary statistics")
```

# Discussion

The findings demonstrate the effectiveness of the research compendium approach for reproducible research.

# Conclusion

This research compendium provides a complete framework for reproducible computational research.

# Acknowledgments

We thank the contributors to the research reproducibility community.

# References

::: {#refs}
:::

# Supplementary Materials

See `supplementary.Rmd` for additional analyses and materials.

# Reproducibility Information

This manuscript was generated using the research compendium structure. To reproduce all analyses:

1. Ensure all dependencies are installed: `renv::restore()`
2. Run the reproduction workflow: `source("analysis/reproduce/04_manuscript_render.R")`
3. All figures and tables are generated programmatically from the underlying data and analysis functions.

## Session Information

```{r session-info}
sessionInfo()
```