---
title: "Supplementary Materials: {{MANUSCRIPT_TITLE}}"
author: "{{AUTHOR_NAME}} {{AUTHOR_LAST}}"
date: "`r Sys.Date()`"
output:
  bookdown::word_document2:
    number_sections: true
  bookdown::pdf_document2:
    latex_engine: xelatex
    number_sections: true
  bookdown::html_document2:
    number_sections: true
    toc: true
    toc_float: true
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Load required packages
library({{PACKAGE_NAME}})
library(dplyr)
library(ggplot2)
library(knitr)
library(here)

# Load all package functions
devtools::load_all()
```

# Supplementary Methods

## Data Processing Details

This section provides detailed documentation of data processing steps implemented in the R package functions.

```{r data-processing-details}
# Detailed data processing code
# show_data_processing_steps()
```

## Additional Statistical Analyses

Extended statistical analyses that support the main findings.

```{r additional-analyses}
# Supplementary statistical analyses
# perform_sensitivity_analysis()
```

# Supplementary Results

## Additional Figures

```{r supplementary-figures, fig.cap="Supplementary analysis figures"}
# Generate supplementary figures
# create_supplementary_figures()
```

## Additional Tables

```{r supplementary-tables}
# Generate supplementary tables
# kable(create_supplementary_tables(), caption = "Detailed results")
```

## Model Diagnostics

```{r model-diagnostics, fig.cap="Model diagnostic plots"}
# Generate model diagnostic plots
# create_diagnostic_plots()
```

# Code Availability

All analysis code is available as documented R functions in this research compendium:

- Data processing functions: `R/data_processing.R`
- Statistical analysis functions: `R/statistical_analysis.R`
- Plotting functions: `R/plotting_functions.R`
- Utility functions: `R/utils.R`

## Reproduction Scripts

Complete reproduction workflows are available in `analysis/reproduce/`:

1. `01_data_preparation.R` - Data cleaning and preparation
2. `02_statistical_analysis.R` - Statistical modeling and inference
3. `03_figures_tables.R` - Generate all manuscript outputs
4. `04_manuscript_render.R` - Render final manuscript formats

## Testing

All functions include comprehensive unit tests in `tests/testthat/` to ensure computational reproducibility.

# References