---
title: "ZZCOLLAB Configuration System"
author: "ZZCOLLAB Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{ZZCOLLAB Configuration System}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

ZZCOLLAB's configuration system provides flexible control over Docker environments, package management, and team collaboration through a multi-layered configuration hierarchy and comprehensive CLI flags.

## Key Features

- **Multi-layered configuration**: User, project, system, and built-in defaults
- **Docker profile system**: 14+ specialized environments from 200MB to 3.5GB
- **Dynamic package management**: Add packages via `renv::install()` as needed
- **Two-layer architecture**: Docker environment specification + renv packages
- **CLI and R interfaces**: Work via command line or entirely within R/RStudio

## Configuration Philosophy

ZZCOLLAB follows a "configure once, use everywhere" approach:
- Set defaults once in configuration files
- Override per-project when needed
- Eliminate repetitive flag typing
- Share team configurations via version control

# Configuration Hierarchy

## Priority Order (Highest to Lowest)

ZZCOLLAB loads configuration from multiple sources:

1. **Command-line flags** - Explicit flags override all other settings
2. **Environment variables** - Shell environment variables (e.g., `ZZCOLLAB_TEAM_NAME`)
3. **Project config** - `./zzcollab.yaml` in current directory (team-specific)
4. **User config** - `~/.zzcollab/config.yaml` (personal defaults)
5. **System config** - `/etc/zzcollab/config.yaml` (organization-wide)
6. **Built-in defaults** - Fallback values ensuring functionality

Higher priority sources override lower priority ones, allowing precise control at each level.

## Configuration Locations

### User Configuration (~/.zzcollab/config.yaml)

Personal defaults across all your projects:

```yaml
# ~/.zzcollab/config.yaml
defaults:
  team-name: "myteam"
  github-account: "myusername"
  dockerhub-account: "myusername"
  profile-name: "analysis"
  dotfiles-dir: "~/dotfiles"

# Optional: Custom profile compositions
profiles:
  my-custom:
    base-image: "rocker/r-ver:latest"
    libs-bundle: "minimal"
    pkgs-bundle: "tidyverse"
```

**When to use**: Set your personal preferences once, apply to all future projects.

### Project Configuration (./zzcollab.yaml)

Team-specific settings shared via Git:

```yaml
# ./zzcollab.yaml (in project root)
team:
  name: "genomicslab"
  project: "cancer-study"
  description: "Multi-omics cancer research"

defaults:
  profile-name: "bioinformatics"
  dotfiles-dir: "~/dotfiles"

collaboration:
  github:
    auto-create-repo: true
    default-visibility: "private"
    enable-actions: true
```

**When to use**: Share team standards, ensure all members use identical configurations.

### System Configuration (/etc/zzcollab/config.yaml)

Organization-wide policies (requires admin):

```yaml
# /etc/zzcollab/config.yaml
organization:
  name: "University Research Computing"
  docker-registry: "registry.university.edu"

defaults:
  profile-name: "analysis"

collaboration:
  github:
    default-visibility: "private"  # Enforce private repos
    enable-actions: true
```

**When to use**: IT administrators setting organization-wide policies.

# CLI Configuration Commands

## Essential Configuration Commands

### Initialize Configuration

```bash
# Create user configuration file with interactive prompts
zzcollab --config init

# Create configuration in specific location
zzcollab --config init --config-file ./custom-config.yaml
```

### Set Configuration Values

```bash
# Set team name (typically your GitHub username for solo work)
zzcollab --config set team-name "myteam"

# Set GitHub account
zzcollab --config set github-account "myusername"

# Set Docker Hub account (defaults to team-name if not set)
zzcollab --config set dockerhub-account "myusername"

# Set default Docker profile
zzcollab --config set profile-name "analysis"

# Set dotfiles directory
zzcollab --config set dotfiles-dir "~/dotfiles"
```

### Get Configuration Values

```bash
# Get specific value
zzcollab --config get team-name
# Output: myteam

# Get Docker profile
zzcollab --config get profile-name
# Output: analysis

# Get dotfiles directory
zzcollab --config get dotfiles-dir
# Output: /Users/username/dotfiles
```

### List All Configuration

```bash
# Show all effective configuration
zzcollab --config list

# Show configuration sources (which file provides each value)
zzcollab --config list --verbose

# Output shows:
# team-name: "myteam" (from ~/.zzcollab/config.yaml)
# profile-name: "bioinformatics" (from ./zzcollab.yaml) [OVERRIDING user config]
# dotfiles-dir: "~/dotfiles" (from ~/.zzcollab/config.yaml)
```

### Validate Configuration

```bash
# Check configuration for errors and inconsistencies
zzcollab --config validate

# Validate with detailed output
zzcollab --config validate --verbose
```

## Configuration Workflow Examples

### Solo Researcher Setup

```bash
# One-time configuration
zzcollab --config init
zzcollab --config set team-name "myresearch"
zzcollab --config set github-account "myusername"
zzcollab --config set profile-name "analysis"
zzcollab --config set dotfiles-dir "~/dotfiles"

# Create projects using defaults
mkdir penguin-study && cd penguin-study
zzcollab  # Uses all configured defaults
```

### Team Lead Setup

```bash
# Personal defaults
zzcollab --config set team-name "genomicslab"
zzcollab --config set github-account "genomicslab"
zzcollab --config set profile-name "bioinformatics"

# Create team project with project-specific config
mkdir cancer-study && cd cancer-study
zzcollab -p cancer-study

# Customize project config for team
cat > zzcollab.yaml << 'EOF'
team:
  name: "genomicslab"
  project: "cancer-study"

defaults:
  profile-name: "bioinformatics"

collaboration:
  github:
    auto-create-repo: true
    default-visibility: "private"
EOF

# Build and share team image
make docker-build
make docker-push-team
git add . && git commit -m "Initial team setup" && git push
```

### Team Member Setup

```bash
# Minimal personal config needed
zzcollab --config set team-name "genomicslab"
zzcollab --config set dotfiles-dir "~/dotfiles"

# Join existing project
git clone https://github.com/genomicslab/cancer-study.git
cd cancer-study
zzcollab --use-team-image  # Pull team's Docker image

# Project config (./zzcollab.yaml) provides all team settings
```

# CLI Flags Reference

## Project Initialization Flags

### Team and Project Identity

```bash
# -t, --team-name NAME
# Docker Hub namespace for team images
zzcollab -t myteam -p study1

# -p, --project-name NAME
# Project name (used for Docker image naming)
zzcollab -t myteam -p cancer-analysis

# Combined usage
zzcollab -t genomicslab -p study1
# Creates: genomicslab/study1:latest
```

**Image Naming Convention**:
- Team lead: `genomicslab/study1:latest` (team image)
- Personal: `username/study1-personal:latest` (if customized)

### Docker Profile Selection

```bash
# --profile-name PROFILE
# Use predefined Docker profile
zzcollab --profile-name analysis        # ~1.2GB, tidyverse ecosystem
zzcollab --profile-name minimal         # ~780MB, essential dev tools
zzcollab --profile-name bioinformatics  # ~1.98GB, Bioconductor
zzcollab --profile-name geospatial      # ~1.5GB, GIS tools
zzcollab --profile-name modeling        # ~1.5GB, tidymodels
zzcollab --profile-name publishing      # ~3GB, LaTeX/Quarto
zzcollab --profile-name alpine_minimal  # ~200MB, ultra-lightweight

# List all available profiles
zzcollab --list-profiles
```

**Profile Details**: See "Docker Profile System" section below.

### Custom Docker Composition

For specialized needs, compose custom environments using base images and bundles:

```bash
# -b, --base-image IMAGE
# Specify custom base Docker image
zzcollab -b rocker/r-ver:latest
zzcollab -b rocker/rstudio:4.3.0
zzcollab -b bioconductor/bioconductor_docker:latest

# --libs BUNDLE[,BUNDLE,...]
# System library bundles (apt-get packages)
zzcollab -b rocker/r-ver --libs minimal       # Basic system deps
zzcollab -b rocker/r-ver --libs geospatial    # GDAL, PROJ, GEOS
zzcollab -b rocker/r-ver --libs bioinfo       # Bioinformatics libs
zzcollab -b rocker/r-ver --libs modeling      # ML system deps

# --pkgs BUNDLE[,BUNDLE,...]
# R package bundles (pre-installed in Docker)
zzcollab -b rocker/r-ver --pkgs essential     # renv, devtools, usethis
zzcollab -b rocker/r-ver --pkgs tidyverse     # tidyverse ecosystem
zzcollab -b rocker/r-ver --pkgs modeling      # tidymodels packages
zzcollab -b rocker/r-ver --pkgs bioinfo       # Bioconductor packages

# Combined custom composition
zzcollab -b bioconductor/bioconductor_docker --libs bioinfo --pkgs modeling
# Creates: Bioconductor base + genomics libs + ML packages

# List available bundles
zzcollab --list-libs   # System library bundles
zzcollab --list-pkgs   # R package bundles
```

**Bundle System**:
- **libs**: System dependencies (compiled libraries, command-line tools)
- **pkgs**: R packages (pre-installed in Docker image for fast startup)

### Dotfiles Integration

```bash
# -d, --dotfiles-dir PATH
# Copy personal dotfiles into container
zzcollab -d ~/dotfiles

# Dotfiles directory structure:
# ~/dotfiles/
# ├── zshrc       # Becomes ~/.zshrc in container
# ├── vimrc       # Becomes ~/.vimrc in container
# ├── tmux.conf   # Becomes ~/.tmux.conf in container
# └── gitconfig   # Becomes ~/.gitconfig in container

# Files without leading dots in ~/dotfiles get dots added automatically
```

**Important**: Store dotfiles WITHOUT leading dots in your dotfiles directory for easier version control.

## Team Collaboration Flags

### Using Team Images

```bash
# --use-team-image
# Pull and use existing team Docker image
git clone https://github.com/myteam/study1.git
cd study1
zzcollab --use-team-image
# Pulls: myteam/study1:latest from Docker Hub

# Check for team image
zzcollab --check-team-image
# Reports if team image exists and is accessible
```

**Team Image Workflow**:
1. Team lead commits Dockerfile to repository
2. Team lead optionally pushes image: `make docker-build && make docker-push-team` (for faster team setup)
3. Team members either:
   - Pull pre-built image: `zzcollab --use-team-image` (faster)
   - Build from Dockerfile: `make docker-build` (slower, works without Docker Hub)
4. Everyone has identical Docker environment from same Dockerfile specification

### Configuration File Override

```bash
# --config-file PATH
# Use custom configuration file location
zzcollab --config-file ./team-config.yaml

# --profiles-config PATH
# Use custom profile definitions
zzcollab --profiles-config ./custom-profiles.yaml
```

## Output and Behavior Flags

### Verbosity Control

```bash
# -v, --verbose
# Show detailed output during operations
zzcollab -v -t myteam -p study1

# -q, --quiet
# Suppress non-essential output
zzcollab -q --config list

# --dry-run
# Show what would be done without executing
zzcollab --dry-run -t myteam -p study1
```

### Confirmation Prompts

```bash
# -y, --yes
# Auto-confirm all prompts (useful for scripts)
zzcollab -y -t myteam -p study1

# --no-confirm
# Skip confirmation prompts
zzcollab --no-confirm --config set team-name "newteam"
```

## Advanced Flags

### Docker Build Control

```bash
# --no-cache
# Force rebuild without using Docker layer cache
zzcollab --no-cache -t myteam -p study1

# --platform PLATFORM
# Specify platform for Docker build
zzcollab --platform linux/amd64  # Force AMD64
zzcollab --platform linux/arm64  # Force ARM64

# --build-arg KEY=VALUE
# Pass build arguments to Docker
zzcollab --build-arg R_VERSION=4.3.0
```

### Package Management

```bash
# --skip-renv
# Skip renv initialization
zzcollab --skip-renv -t myteam -p study1

# --renv-version VERSION
# Use specific renv version
zzcollab --renv-version 1.0.3
```

# Docker Profile System

## Understanding Docker Profiles

Docker profiles are pre-configured environments optimized for specific research workflows. Each profile combines:
- **Base Docker image** (e.g., rocker/rstudio, bioconductor)
- **System libraries** (e.g., GDAL for geospatial, JAGS for Bayesian)
- **R packages** (pre-installed for fast container startup)

Profiles provide the Docker environment (Layer 1), while additional packages are added dynamically via `renv::install()` (Layer 2).

## Available Profiles

### Standard Research Profiles

#### minimal (~780MB)
```yaml
base_image: rocker/r-ver:latest
packages:
  - renv
  - devtools
  - usethis
  - here
  - testthat

use_cases:
  - Lightweight development
  - CI/CD pipelines
  - Quick prototyping
  - Resource-constrained environments
```

#### analysis (~1.2GB) [RECOMMENDED]
```yaml
base_image: rocker/rstudio:latest
packages:
  - renv, devtools, usethis, here, testthat
  - tidyverse (dplyr, ggplot2, tidyr, readr, purrr, stringr, forcats, lubridate)
  - knitr, rmarkdown
  - palmerpenguins (example data)

use_cases:
  - General data analysis
  - Exploratory data analysis
  - Team development
  - Teaching and workshops
```

#### modeling (~1.5GB)
```yaml
base_image: rocker/ml:latest
packages:
  - analysis packages plus:
  - tidymodels (parsnip, recipes, workflows, tune, yardstick)
  - xgboost, ranger, glmnet

use_cases:
  - Machine learning projects
  - Predictive modeling
  - Model evaluation and tuning
  - Feature engineering
```

#### publishing (~3GB)
```yaml
base_image: rocker/verse:latest
packages:
  - analysis packages plus:
  - LaTeX (full installation)
  - quarto, bookdown, blogdown
  - tinytex
  - kableExtra, gt, flextable

use_cases:
  - Manuscript writing
  - Book authoring
  - Academic publishing
  - Technical documentation
```

### Specialized Domain Profiles

#### bioinformatics (~1.98GB)
```yaml
base_image: bioconductor/bioconductor_docker:latest
packages:
  - Bioconductor core packages
  - DESeq2, edgeR, limma
  - GenomicRanges, Biostrings
  - SummarizedExperiment

use_cases:
  - Genomics analysis
  - RNA-seq workflows
  - Proteomics studies
  - Systems biology
```

#### geospatial (~1.5GB)
```yaml
base_image: rocker/geospatial:latest
system_libs:
  - GDAL, PROJ, GEOS (pre-compiled)
packages:
  - sf, terra, raster
  - leaflet, mapview, tmap
  - stars, gstat

use_cases:
  - GIS analysis
  - Spatial statistics
  - Climate modeling
  - Ecology and conservation
```

### Lightweight Alpine Profiles

#### alpine_minimal (~200MB)
```yaml
base_image: rocker/r-alpine:latest
packages:
  - renv, devtools, testthat

use_cases:
  - Ultra-fast CI/CD
  - Minimal resource footprint
  - Container orchestration
  - Edge computing
```

#### alpine_analysis (~400MB)
```yaml
base_image: rocker/r-alpine:latest
packages:
  - Essential tidyverse packages
  - Minimal analysis tools

use_cases:
  - Lightweight analysis
  - Fast testing
  - Cloud deployments
  - Resource-constrained servers
```

#### hpc_alpine (~600MB)
```yaml
base_image: rocker/r-alpine:latest
packages:
  - Parallel processing tools
  - HPC-optimized packages

use_cases:
  - High-performance computing
  - Cluster computing
  - Parallel workflows
  - Batch processing
```

## Profile Selection Guide

| Research Domain | Recommended Profile | Alternative Profiles |
|-----------------|---------------------|----------------------|
| Data analysis | analysis | minimal, alpine_analysis |
| Machine learning | modeling | analysis |
| Genomics | bioinformatics | minimal + renv packages |
| GIS/spatial | geospatial | analysis + spatial packages |
| Publishing | publishing | analysis + publishing packages |
| CI/CD testing | alpine_minimal | minimal |
| Teaching | analysis | minimal, publishing |

## Custom Profile Composition

When predefined profiles don't fit, compose custom environments:

### Example: Bayesian Statistics Environment

```bash
# Start with R base, add statistical libraries and modeling packages
zzcollab -b rocker/r-ver:latest \
  --libs modeling \
  --pkgs modeling \
  -t myteam -p bayes-study

# Inside container, add specialized Bayesian packages via renv:
# renv::install(c("rstan", "brms", "rstanarm", "bayesplot", "loo"))
```

### Example: Web Scraping + NLP Environment

```bash
# Base image with web tools and text analysis
zzcollab -b rocker/rstudio:latest \
  --libs minimal \
  --pkgs tidyverse \
  -t myteam -p nlp-scraper

# Add NLP packages dynamically:
# renv::install(c("rvest", "httr", "xml2", "tidytext", "textrecipes", "spacyr"))
```

### Example: Lightweight Reporting

```bash
# Alpine base for minimal footprint, add publishing tools
zzcollab -b rocker/r-alpine:latest \
  --libs alpine \
  --pkgs essential \
  -t myteam -p reports

# Add reporting packages:
# renv::install(c("rmarkdown", "knitr", "gt", "flextable"))
```

# Two-Layer Package Management Architecture

ZZCOLLAB uses a two-layer system for maximum flexibility and reproducibility.

## Layer 1: Docker Profile (Team/Shared)

**What**: Pre-installed R packages in the Docker image
**Control**: `--profile-name`, `--pkgs`, or profile configuration
**Shared**: All team members get identical packages
**Purpose**: Fast container startup, team consistency

```bash
# Team lead creates base environment
zzcollab -t genomicslab -p study --profile-name bioinformatics
make docker-build
make docker-push-team

# All team members get identical Bioconductor environment
```

**Layer 1 packages are in the Docker image**:
- Available immediately when container starts
- No installation time
- Shared across all users of the image
- Fixed until image is rebuilt

## Layer 2: Dynamic renv (Personal/Flexible)

**What**: Packages added via `renv::install()` inside containers
**Control**: Team members install as needed
**Flexible**: Each person can add different packages
**Purpose**: Analysis-specific packages, reproducible dependency tracking

```bash
# Inside container
make docker-zsh

# Install packages as needed
renv::install("tidymodels")
renv::install("pins")
renv::install("vetiver")

# Save to renv.lock
renv::snapshot()

# Exit and commit
exit
git add renv.lock DESCRIPTION
git commit -m "Add ML deployment packages"
git push
```

**Layer 2 packages via renv**:
- Installed on-demand inside containers
- Added dynamically as analysis needs evolve
- Accumulates packages from all team members
- renv.lock is the source of truth for reproducibility

## How the Layers Work Together

### Scenario: Genomics Team with Multiple Analyses

**Team Lead Setup (Layer 1)**:
```bash
# Create bioinformatics base with common packages
zzcollab -t genomicslab -p baseenv --profile-name bioinformatics
make docker-build
make docker-push-team
```

**Team Member Alice (Layer 2 - RNA-seq)**:
```bash
git clone https://github.com/genomicslab/rnaseq-study.git
cd rnaseq-study
zzcollab --use-team-image
make docker-zsh

# Add RNA-seq specific packages
renv::install(c("tximport", "DESeq2", "apeglm", "EnhancedVolcano"))
renv::snapshot()
```

**Team Member Bob (Layer 2 - ChIP-seq)**:
```bash
git clone https://github.com/genomicslab/chipseq-study.git
cd chipseq-study
zzcollab --use-team-image
make docker-zsh

# Add ChIP-seq specific packages
renv::install(c("ChIPseeker", "ChIPpeakAnno", "DiffBind", "csaw"))
renv::snapshot()
```

**Result**:
- Both use **identical Layer 1** (Bioconductor base via team image)
- Each has **different Layer 2** (analysis-specific packages in renv.lock)
- Complete reproducibility via renv.lock (source of truth)

## Key Principle: renv.lock is Source of Truth

Docker images provide foundation and performance, but **renv.lock ensures reproducibility**:

```bash
# Five years later, reproduce Alice's RNA-seq analysis
git clone https://github.com/genomicslab/rnaseq-study.git
cd rnaseq-study

# Can use ANY compatible Docker base image
zzcollab --profile-name minimal  # or analysis, or bioinformatics
make docker-zsh

# renv restores EXACT packages Alice used
renv::restore()  # Installs tximport, DESeq2, apeglm, EnhancedVolcano at exact versions

# Run analysis - identical results
Rscript analysis/scripts/differential_expression.R
```

**Why this works**: renv.lock specifies exact package versions from CRAN/Bioconductor/GitHub snapshots, independent of Docker image.

## Package Selection Strategy

### Choose Docker Profile (Layer 1) Based On:

1. **Research domain**
   - Bioinformatics → `bioinformatics` profile
   - GIS/spatial → `geospatial` profile
   - Machine learning → `modeling` profile
   - General analysis → `analysis` profile

2. **Team consensus**
   - What does everyone need?
   - What's the common foundation?

3. **Performance considerations**
   - Large, stable packages → Docker profile (tidyverse, Bioconductor core)
   - Team-wide dependencies → Docker profile

### Add via renv (Layer 2) Based On:

1. **Analysis-specific needs**
   - Domain-specific packages (tximport, ChIPseeker)
   - Specialized statistical methods

2. **Experimental packages**
   - Newly released packages
   - Development versions from GitHub

3. **Version-specific requirements**
   - Packages requiring exact versions
   - Packages with breaking changes between versions

# Practical Configuration Scenarios

## Scenario 1: Solo Data Scientist

**Goal**: Quick project setup, standard data analysis tools

**Configuration**:
```bash
# One-time setup
zzcollab --config init
zzcollab --config set team-name "datasci"
zzcollab --config set profile-name "analysis"
zzcollab --config set dotfiles-dir "~/dotfiles"

# Create projects effortlessly
mkdir customer-churn && cd customer-churn
zzcollab  # Uses all configured defaults

mkdir sales-forecast && cd sales-forecast
zzcollab  # Consistent environment across projects
```

**Benefits**:
- No repetitive flag typing
- Consistent environment across projects
- Professional reproducibility for portfolio

## Scenario 2: Academic Research Lab

**Goal**: Team collaboration, consistent environments, specialized workflows

**Team Lead Setup**:
```bash
# Personal config
zzcollab --config set team-name "psychlab"
zzcollab --config set github-account "psychlab"

# Create shared project
mkdir attention-study && cd attention-study
zzcollab -p attention-study --profile-name analysis

# Customize for team
cat > zzcollab.yaml << 'EOF'
team:
  name: "psychlab"
  project: "attention-study"

defaults:
  profile-name: "analysis"
  dotfiles-dir: "~/dotfiles"

collaboration:
  github:
    auto-create-repo: true
    default-visibility: "private"
    enable-actions: true
EOF

# Build and share
make docker-build
make docker-push-team
git add . && git commit -m "Team setup" && git push
```

**Team Member Workflow**:
```bash
# Minimal personal setup
zzcollab --config set team-name "psychlab"
zzcollab --config set dotfiles-dir "~/dotfiles"

# Join project
git clone https://github.com/psychlab/attention-study.git
cd attention-study
zzcollab --use-team-image

# Start working immediately
make docker-zsh
```

## Scenario 3: Bioinformatics Core Facility

**Goal**: Multiple specialized environments, organization-wide standards

**System Admin Setup** (`/etc/zzcollab/config.yaml`):
```yaml
organization:
  name: "Bioinformatics Core"
  docker-registry: "registry.core.university.edu"

defaults:
  github-account: "bioinformatics-core"

collaboration:
  github:
    default-visibility: "private"
    enable-actions: true
```

**Core Director Setup**:
```bash
# Base images for different workflows
mkdir biocore-rnaseq && cd biocore-rnaseq
zzcollab -t biocore -p rnaseq-base --profile-name bioinformatics
make docker-build && make docker-push-team

mkdir biocore-chipseq && cd ../biocore-chipseq
zzcollab -t biocore -p chipseq-base --profile-name bioinformatics
make docker-build && make docker-push-team

mkdir biocore-singlecell && cd ../biocore-singlecell
zzcollab -t biocore -p singlecell-base --profile-name bioinformatics
make docker-build && make docker-push-team
```

**Researcher Workflow**:
```bash
# Join workflow-specific project
git clone https://github.com/biocore/rnaseq-study.git
cd rnaseq-study
zzcollab --use-team-image  # Gets appropriate base image

# Add analysis-specific packages
make docker-zsh
renv::install("workflow_specific_package")
renv::snapshot()
```

## Scenario 4: Data Science Consulting Firm

**Goal**: Client projects, quarterly environment updates, reproducible deliverables

**DevOps Team Setup**:
```bash
# Quarterly standardized environment
mkdir q1-2025-base && cd q1-2025-base
zzcollab -t datasci-firm -p q1-2025-base --profile-name modeling

# Document package versions
cat > ENVIRONMENT.md << 'EOF'
# Q1 2025 Standardized Environment

Base: rocker/ml:latest (R 4.4.0)
Profile: modeling

Key packages:
- tidyverse 2.0.0
- tidymodels 1.2.0
- xgboost 1.7.6
- shiny 1.8.0

Last updated: 2025-01-15
EOF

make docker-build
make docker-push-team
git add . && git commit -m "Q1 2025 base environment" && git push
```

**Data Scientist Workflow**:
```bash
# Create client project using quarterly base
mkdir client-churn-prediction && cd client-churn-prediction
zzcollab -t datasci-firm -p client-churn --profile-name modeling

# Team base image provides foundation
make docker-zsh

# Add client-specific packages
renv::install(c("probably", "workflowsets", "themis"))
renv::snapshot()

# All projects use same Q1 2025 foundation
# Client-specific packages in renv.lock
```

# Configuration Validation and Troubleshooting

## Validation Commands

### Check Configuration Integrity

```bash
# Validate all configuration files
zzcollab --config validate

# Expected output:
# ✓ User configuration valid: ~/.zzcollab/config.yaml
# ✓ No project configuration found (not in project directory)
# ✓ No system configuration found (not installed system-wide)
# ✓ Configuration hierarchy is consistent
```

### View Effective Configuration

```bash
# See merged configuration from all sources
zzcollab --config list

# See which file provides each value
zzcollab --config list --verbose

# Output:
# team-name: "genomicslab"
#   Source: ~/.zzcollab/config.yaml (user config)
# profile-name: "bioinformatics"
#   Source: ./zzcollab.yaml (project config) [OVERRIDES user setting "analysis"]
# dotfiles-dir: "~/dotfiles"
#   Source: ~/.zzcollab/config.yaml (user config)
```

### Test Configuration

```bash
# Dry-run project creation
zzcollab --dry-run -t myteam -p test-project --profile-name analysis

# Shows what would be created without executing:
# Would create project: test-project
# Team namespace: myteam
# Docker profile: analysis (~1.2GB)
# Base image: rocker/rstudio:latest
# Docker image name: myteam/test-project:latest
```

## Common Issues and Solutions

### Issue 1: Configuration Hierarchy Confusion

**Problem**: Settings not taking effect as expected

```bash
# Symptom: Set profile-name in user config but project uses different profile
zzcollab --config set profile-name "analysis"
zzcollab --config get profile-name
# Output: bioinformatics  (Expected: analysis)
```

**Solution**: Check configuration sources
```bash
zzcollab --config list --verbose
# Output shows:
# profile-name: "bioinformatics"
#   Source: ./zzcollab.yaml (project config) [OVERRIDES user setting]

# Project config (./zzcollab.yaml) overrides user config
# Remove from project config or understand this is intentional
vim ./zzcollab.yaml  # Remove profile-name to use user default
```

### Issue 2: Docker Platform Compatibility

**Problem**: ARM64 Mac cannot build AMD64-only profiles

```bash
# Error when building geospatial profile:
# Error: rocker/geospatial:latest: no matching manifest for linux/arm64/v8
```

**Solution 1**: Use ARM64-compatible profiles
```bash
# Instead of geospatial, use:
zzcollab --profile-name analysis -t myteam -p study
# Then add spatial packages via renv:
# renv::install(c("sf", "terra", "raster"))
```

**Solution 2**: Force AMD64 platform
```bash
zzcollab --platform linux/amd64 --profile-name geospatial -t myteam -p study
# Runs via emulation (slower but compatible)
```

**Solution 3**: Check profile compatibility
```bash
zzcollab --list-profiles
# Output shows platform compatibility:
# analysis      ARM64 ✓  AMD64 ✓
# geospatial    ARM64 ✗  AMD64 ✓  (AMD64 only)
# alpine_*      ARM64 ✓  AMD64 ✓
```

### Issue 3: Missing Configuration File

**Problem**: Configuration file not found

```bash
# Error: Configuration file not found: ~/.zzcollab/config.yaml
```

**Solution**: Initialize configuration
```bash
zzcollab --config init
# Interactive prompts:
# Team name: myteam
# GitHub account: myusername
# Docker profile (default: analysis):
# Dotfiles directory (default: ~/dotfiles):

# Configuration created at: ~/.zzcollab/config.yaml
```

### Issue 4: Docker Daemon Not Running

**Problem**: Cannot connect to Docker daemon

```bash
# Error: Cannot connect to the Docker daemon at unix:///var/run/docker.sock
```

**Solution**: Start Docker daemon
```bash
# macOS/Windows: Start Docker Desktop app

# Linux: Start Docker service
sudo systemctl start docker
sudo systemctl enable docker  # Auto-start on boot

# Verify Docker is running
docker info
```

### Issue 5: Profile Not Found

**Problem**: Custom profile specified but not defined

```bash
# Error: Profile 'my-custom' not found in configuration
```

**Solution**: Define the profile
```bash
# Edit user configuration
vim ~/.zzcollab/config.yaml

# Add custom profile:
profiles:
  my-custom:
    base-image: "rocker/r-ver:latest"
    libs-bundle: "minimal"
    pkgs-bundle: "tidyverse"
    description: "My custom analysis environment"
```

# R Interface Configuration

ZZCOLLAB provides a complete R interface for configuration management.

## Configuration Functions

```{r config-functions}
library(zzcollab)

# Initialize configuration
init_config()

# Set configuration values
set_config("team_name", "myteam")
set_config("profile_name", "analysis")
set_config("dotfiles_dir", "~/dotfiles")
set_config("github_account", "myusername")

# Get configuration values
team <- get_config("team_name")
profile <- get_config("profile_name")
dotfiles <- get_config("dotfiles_dir")

# List all configuration
all_config <- list_config()
print(all_config)

# Validate configuration
valid <- validate_config()
if (!valid$valid) {
  message("Configuration errors:")
  print(valid$errors)
}
```

## Configuration-Aware Project Functions

```{r config-projects}
# Functions use configuration defaults automatically

# Create project with defaults from config
init_project("my-analysis")
# Uses: team_name, profile_name, dotfiles_dir from config

# Override specific settings
init_project(
  project_name = "special-study",
  profile_name = "bioinformatics"  # Override default profile
  # Other settings from config
)

# Explicit full control
init_project(
  team_name = "genomicslab",
  project_name = "cancer-study",
  profile_name = "bioinformatics",
  dotfiles_path = "~/lab-dotfiles"
)
```

## Configuration Inspection

```{r config-inspect}
# Get effective configuration from all sources
effective <- get_effective_config()
str(effective)

# Check configuration sources
sources <- get_config_sources()
print(sources$team_name)  # Shows which file provides value

# Get default values
defaults <- get_config_defaults()
print(defaults)
```

# Best Practices

## Configuration Strategy by Use Case

### Solo Researcher
- **Use**: User configuration only (`~/.zzcollab/config.yaml`)
- **Profiles**: 1-2 profiles (minimal, analysis)
- **Package management**: Add via renv as needed
- **Version control**: Commit renv.lock, not user config

### Small Team (2-5 people)
- **Use**: Project configuration (`./zzcollab.yaml` in Git)
- **Profiles**: 2-3 profiles for team domains
- **Package management**: Team discusses additions, uses renv
- **Version control**: Commit project config, renv.lock

### Large Organization (10+ people)
- **Use**: System + project configurations
- **Profiles**: Comprehensive profile library
- **Package management**: IT manages base images, teams use renv
- **Version control**: Centralized profile library + project configs

## Security Considerations

### Private Repository Settings

```yaml
# ./zzcollab.yaml
collaboration:
  github:
    default-visibility: "private"  # Always private
    branch-protection: true         # Require PR reviews

repository:
  allow-merge-commits: false       # Clean history
  allow-squash-merging: true      # Allow squash
  delete-head-branches: true      # Auto-cleanup
```

### Container Security

```yaml
development:
  container:
    default-user: "analyst"        # Never run as root
    read-only-volumes: true        # Mount shared volumes read-only
```

### Sensitive Data

- **Never commit**: API keys, passwords, credentials
- **Use**: Environment variables for secrets
- **Store**: Credentials outside Git (e.g., `~/.Renviron`)

## Performance Optimization

### Build Time Reduction

1. **Use appropriate profiles**
   - Development: `minimal` or `analysis`
   - CI/CD: `alpine_minimal`
   - Production: `modeling` or domain-specific

2. **Layer caching**
   - Base image changes → full rebuild
   - Package additions → partial rebuild
   - Use stable base images

3. **Parallel operations**
   - Enable parallel package installation
   - Build profiles concurrently when possible

### Container Size Optimization

```yaml
# Prefer lightweight profiles for CI/CD
profiles:
  alpine_minimal:    # ~200MB
    enabled: true
  analysis:          # ~1.2GB (for development)
    enabled: true
  # Avoid enabling unnecessary heavy profiles
```

## Troubleshooting Strategy

1. **Check configuration hierarchy**
   ```bash
   zzcollab --config list --verbose
   ```

2. **Validate configuration**
   ```bash
   zzcollab --config validate
   ```

3. **Test with dry-run**
   ```bash
   zzcollab --dry-run -t myteam -p test
   ```

4. **Check Docker status**
   ```bash
   docker info
   docker images
   ```

5. **Verbose output**
   ```bash
   zzcollab -v -t myteam -p test --profile-name analysis
   ```

# Summary

ZZCOLLAB's configuration system provides powerful, flexible control over reproducible research environments:

## Key Concepts

1. **Configuration hierarchy**: Command-line → environment → project → user → system → defaults
2. **Docker profiles**: 14+ specialized environments from 200MB to 3.5GB
3. **Two-layer architecture**: Docker profiles (shared) + renv packages (flexible)
4. **CLI flags**: Comprehensive control over all aspects
5. **R interface**: Complete configuration management within R

## Quick Reference

**Initialize configuration**:
```bash
zzcollab --config init
zzcollab --config set team-name "myteam"
zzcollab --config set profile-name "analysis"
```

**Create project**:
```bash
zzcollab -t myteam -p study --profile-name analysis -d ~/dotfiles
```

**Team collaboration**:
```bash
# Team lead:
make docker-build && make docker-push-team

# Team member:
zzcollab --use-team-image
```

**Package management**:
```bash
# Inside container:
renv::install("package")
renv::snapshot()
```

## Further Reading

- `zzcollab --help` - Complete CLI reference
- `help(package = "zzcollab")` - R package documentation
- `vignette("getting-started")` - Comprehensive tutorial
- `docs/VARIANTS.md` - Detailed profile documentation
- `docs/CONFIGURATION.md` - Advanced configuration guide
