---
title: "ZZCOLLAB Configuration System"
author: "ZZCOLLAB Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{ZZCOLLAB Configuration System}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

ZZCOLLAB's configuration system provides flexible control over Docker environments, package management, and team collaboration through a multi-layered configuration hierarchy and comprehensive CLI flags.

## Key Features

- **Multi-layered configuration**: User, project, system, and built-in defaults
- **Docker profile system**: 14+ specialized environments from 200MB to 3.5GB
- **Dynamic package management**: Add packages via `install.packages()` with automatic snapshot/restore
- **Two-layer architecture**: Docker profiles (team/shared) + renv packages (personal/flexible)
- **CLI and R interfaces**: Work via command line or entirely within R/RStudio

## Configuration Philosophy

ZZCOLLAB follows a "configure once, use everywhere" approach:
- Set defaults once in configuration files
- Override per-project when needed
- Eliminate repetitive flag typing
- Share team configurations via version control

# Configuration Hierarchy

## Priority Order (Highest to Lowest)

ZZCOLLAB loads configuration from multiple sources:

1. **Command-line flags** - Explicit flags override all other settings
2. **Environment variables** - Shell environment variables (e.g., `ZZCOLLAB_TEAM_NAME`)
3. **Project config** - `./zzcollab.yaml` in current directory (team-specific)
4. **User config** - `~/.zzcollab/config.yaml` (personal defaults)
5. **System config** - `/etc/zzcollab/config.yaml` (organization-wide)
6. **Built-in defaults** - Fallback values ensuring functionality

Higher priority sources override lower priority ones, allowing precise control at each level.

## Configuration Locations

### User Configuration (~/.zzcollab/config.yaml)

Personal defaults across all your projects:

```yaml
# ~/.zzcollab/config.yaml
defaults:
  team-name: "myteam"
  github-account: "myusername"
  dockerhub-account: "myusername"
  profile-name: "analysis"

# Optional: Custom profile compositions
profiles:
  my-custom:
    base-image: "rocker/r-ver:latest"
    libs-bundle: "minimal"
    pkgs-bundle: "tidyverse"
```

**When to use**: Set your personal preferences once, apply to all future projects.

### Project Configuration (./zzcollab.yaml)

Team-specific settings shared via Git:

```yaml
# ./zzcollab.yaml (in project root)
team:
  name: "genomicslab"
  project: "cancer-study"
  description: "Multi-omics cancer research"

defaults:
  profile-name: "modeling"

collaboration:
  github:
    auto-create-repo: true
    default-visibility: "private"
    enable-actions: true
```

**When to use**: Share team standards, ensure all members use identical configurations.

### System Configuration (/etc/zzcollab/config.yaml)

Organization-wide policies (requires admin):

```yaml
# /etc/zzcollab/config.yaml
organization:
  name: "University Research Computing"
  docker-registry: "registry.university.edu"

defaults:
  profile-name: "analysis"

collaboration:
  github:
    default-visibility: "private"  # Enforce private repos
    enable-actions: true
```

**When to use**: IT administrators setting organization-wide policies.

# CLI Configuration Commands

## Essential Configuration Commands

### Initialize Configuration

```bash
# Create user configuration file with interactive prompts
zzcollab config init
```

### Set Configuration Values

```bash
# Set team name (typically your GitHub username for solo work)
zzcollab config set team-name "myteam"

# Set GitHub account
zzcollab config set github-account "myusername"

# Set Docker Hub account (defaults to team-name if not set)
zzcollab config set dockerhub-account "myusername"

# Set default Docker profile
zzcollab config set profile-name "analysis"
```

### Get Configuration Values

```bash
# Get specific value
zzcollab config get team-name
# Output: myteam

# Get Docker profile
zzcollab config get profile-name
# Output: analysis

```

### List All Configuration

```bash
# Show all effective configuration
zzcollab config list

# Show configuration sources (which file provides each value)
zzcollab config list --verbose

# Output shows:
# team-name: "myteam" (from ~/.zzcollab/config.yaml)
# profile-name: "modeling" (from ./zzcollab.yaml) [OVERRIDING user config]
```

### Validate Configuration

```bash
# Check configuration for errors and inconsistencies
zzcollab config validate

# Validate with detailed output
zzcollab config validate --verbose
```

## Configuration Workflow Examples

### Solo Researcher Setup

```bash
# One-time configuration
zzcollab config init
zzcollab config set team-name "myresearch"
zzcollab config set github-account "myusername"
zzcollab config set profile-name "analysis"

# Create projects using defaults
mkdir penguin-study && cd penguin-study
zzcollab  # Uses all configured defaults
```

### Team Lead Setup

```bash
# Personal defaults
zzcollab config set team-name "genomicslab"
zzcollab config set github-account "genomicslab"
zzcollab config set profile-name "modeling"

# Create team project with project-specific config
mkdir cancer-study && cd cancer-study
zzcollab modeling

# Customize project config for team
cat > zzcollab.yaml << 'EOF'
team:
  name: "genomicslab"
  project: "cancer-study"

defaults:
  profile-name: "modeling"

collaboration:
  github:
    auto-create-repo: true
    default-visibility: "private"
EOF

# Build and share team image
make docker-build
make docker-push-team
git add . && git commit -m "Initial team setup" && git push
```

### Team Member Setup

```bash
# Minimal personal config needed
zzcollab config set team-name "genomicslab"

# Join existing project
git clone https://github.com/genomicslab/cancer-study.git
cd cancer-study
make docker-build  # Build Docker image from project Dockerfile

# Project config (./zzcollab.yaml) provides all team settings
```

# CLI Flags Reference

## Project Initialization Flags

### Team and Project Identity

**Recommended approach** (new):

```bash
# Create project with profile, then add GitHub
mkdir study1 && cd study1
zzc analysis                     # Full setup (init + renv + docker)
zzc github                       # Create private GitHub repo
```

**Team image naming**: record the DockerHub account once; the project name
comes from the working directory.

```bash
# Docker Hub namespace for team images
zzcollab config set dockerhub-account myteam
# Image name becomes: myteam/<project>:latest
```

### Docker Profile Selection

```bash
# Create the project with a profile (run in the project directory)
zzcollab analysis         # ~1.2GB, tidyverse ecosystem
zzcollab minimal          # ~650MB, essential dev tools
zzcollab modeling         # ~1.5GB, tidymodels / ML
zzcollab publishing       # ~3GB, LaTeX/Quarto
zzcollab rstudio          # RStudio Server
zzcollab shiny            # Shiny apps

# Switch profile on the docker step
zzcollab docker --profile publishing

# List all available profiles
zzcollab list
```

**Profile Details**: See "Docker Profile System" section below.

### Custom Docker Composition

For specialized needs, build on a custom base image with the `docker` command:

```bash
# Override the base Docker image
zzcollab docker --base-image rocker/r-ver:latest
zzcollab docker --base-image rocker/rstudio:4.3.0
zzcollab docker --base-image bioconductor/bioconductor_docker:latest
```

Additional R packages are installed inside the container and captured in
`renv.lock`; the package set is recorded there rather than selected by a CLI
flag. System library dependencies are detected automatically.


## Team Collaboration Flags

### Team Docker Workflow

```bash
# Team member joins existing project
git clone https://github.com/myteam/study1.git
cd study1
make docker-build  # Build Docker image from project Dockerfile
make r             # Enter container
```

**Team Image Workflow**:
1. Team lead builds and pushes: `make docker-build && make docker-push-team`
2. Team members clone and build: `git clone ... && make docker-build`
3. Everyone has identical Docker environment (same Dockerfile)

## Output and Behavior Flags

These global flags work with any command:

```bash
# -v / --verbose : detailed output
zzcollab -v docker

# -q / --quiet : errors only
zzcollab -q config list

# -y / --yes (or -Y / --yes-all) : accept defaults (non-interactive)
zzcollab -y analysis

# --no-build : skip the Docker build prompt
zzcollab --no-build docker
```

# Docker Profile System

## Understanding Docker Profiles

Docker profiles are pre-configured environments optimized for specific research workflows. Each profile combines:
- **Base Docker image** (e.g., rocker/rstudio, bioconductor)
- **System libraries** (e.g., GDAL for geospatial, JAGS for Bayesian)
- **R packages** (pre-installed for fast container startup)

Profiles provide the Docker environment (Layer 1), while additional packages are added dynamically via `install.packages()` with automatic snapshot/restore on container exit (Layer 2).

## Available Profiles

### Standard Research Profiles

#### minimal (~780MB)
```yaml
base_image: rocker/r-ver:latest
packages:
  - renv
  - devtools
  - usethis
  - here
  - testthat

use_cases:
  - Lightweight development
  - CI/CD pipelines
  - Quick prototyping
  - Resource-constrained environments
```

#### analysis (~1.2GB) [RECOMMENDED]
```yaml
base_image: rocker/rstudio:latest
packages:
  - renv, devtools, usethis, here, testthat
  - tidyverse (dplyr, ggplot2, tidyr, readr, purrr, stringr, forcats, lubridate)
  - knitr, rmarkdown
  - palmerpenguins (example data)

use_cases:
  - General data analysis
  - Exploratory data analysis
  - Team development
  - Teaching and workshops
```

#### modeling (~1.5GB)
```yaml
base_image: rocker/ml:latest
packages:
  - analysis packages plus:
  - tidymodels (parsnip, recipes, workflows, tune, yardstick)
  - xgboost, ranger, glmnet

use_cases:
  - Machine learning projects
  - Predictive modeling
  - Model evaluation and tuning
  - Feature engineering
```

#### publishing (~3GB)
```yaml
base_image: rocker/verse:latest
packages:
  - analysis packages plus:
  - LaTeX (full installation)
  - quarto, bookdown, blogdown
  - tinytex
  - kableExtra, gt, flextable

use_cases:
  - Manuscript writing
  - Book authoring
  - Academic publishing
  - Technical documentation
```

### Additional Profiles

#### analysis_pdf (~1.5GB)
```yaml
base_image: rocker/tidyverse:latest
packages:
  - analysis packages plus tinytex, pandoc

use_cases:
  - Data analysis with PDF report rendering
```

#### shiny (~1.8GB)
```yaml
base_image: rocker/shiny:latest
packages:
  - shiny, shinydashboard, DT, plotly, bslib

use_cases:
  - Interactive web applications and dashboards
```

#### manuscript-package (~1.5GB)
```yaml
base_image: rocker/verse:latest
packages:
  - R package structure with companion Rmd manuscripts (tinytex)

use_cases:
  - R packages with paper-style reports
```

### Specialized Domains via a Custom Base Image

For domains without a built-in profile (e.g. genomics or geospatial), build on a
domain base image with `zzcollab docker --base-image` and add packages via renv:

```bash
zzcollab docker --base-image bioconductor/bioconductor_docker  # genomics
zzcollab docker --base-image rocker/geospatial                 # GIS / spatial
```

## Profile Selection Guide

| Research Domain | Recommended Profile | Alternative |
|-----------------|---------------------|-------------|
| Data analysis | analysis | minimal |
| Machine learning | modeling | analysis |
| Genomics | analysis + Bioconductor base image | minimal |
| GIS/spatial | analysis + rocker/geospatial base image | minimal |
| Publishing | publishing | manuscript-package |
| Interactive apps | shiny | rstudio |
| Teaching | analysis | minimal, publishing |

## Custom Profile Composition

When predefined profiles don't fit, compose custom environments:

### Example: Bayesian Statistics Environment

```bash
# Start from the modeling profile
mkdir bayes-study && cd bayes-study
zzcollab modeling

# Inside container, add specialized Bayesian packages:
# install.packages(c("rstan", "brms", "rstanarm", "bayesplot", "loo"))
# (Auto-snapshot on exit saves these to renv.lock)
```

### Example: Web Scraping + NLP Environment

```bash
# Start from the analysis profile
mkdir nlp-scraper && cd nlp-scraper
zzcollab analysis

# Add NLP packages dynamically:
# install.packages(c("rvest", "httr", "xml2", "tidytext", "textrecipes", "spacyr"))
# (Auto-snapshot on exit saves these to renv.lock)
```

### Example: Lightweight Reporting

```bash
# Minimal base for a small footprint
mkdir reports && cd reports
zzcollab minimal

# Add reporting packages:
# install.packages(c("rmarkdown", "knitr", "gt", "flextable"))
# (Auto-snapshot on exit saves these to renv.lock)
```

# Two-Layer Package Management Architecture

ZZCOLLAB uses a two-layer system for maximum flexibility and reproducibility.

## Layer 1: Docker Profile (Team/Shared)

**What**: Pre-installed R packages in the Docker image
**Control**: the profile (`zzc <profile>` / `docker --profile`) or a custom base image
**Shared**: All team members get identical packages
**Purpose**: Fast container startup, team consistency

```bash
# Team lead creates base environment
mkdir study && cd study
zzc analysis                     # Or: zzc modeling for a machine-learning profile
zzc github
make docker-build
zzc dockerhub                    # Push for team members

# All team members get identical environment
```

**Layer 1 packages are in the Docker image**:
- Available immediately when container starts
- No installation time
- Shared across all users of the image
- Fixed until image is rebuilt

## Layer 2: Dynamic renv (Personal/Flexible)

**What**: Packages added via `install.packages()` inside containers with automatic snapshot on exit
**Control**: Team members install as needed
**Flexible**: Each person can add different packages
**Purpose**: Analysis-specific packages, reproducible dependency tracking

```bash
# Inside container
make r

# Install packages as needed (standard R command)
install.packages("tidymodels")
install.packages("pins")
install.packages("vetiver")

# When you exit R, packages are automatically saved to renv.lock (via .Last() in .Rprofile)
q()

# Back on host - commit the updated renv.lock
git add renv.lock DESCRIPTION
git commit -m "Add ML deployment packages"
git push
```

**Layer 2 packages via renv**:
- Installed on-demand inside containers
- Added dynamically as analysis needs evolve
- Accumulates packages from all team members
- renv.lock is the source of truth for reproducibility

## How the Layers Work Together

### Scenario: Genomics Team with Multiple Analyses

**Team Lead Setup (Layer 1)**:
```bash
# Create a base environment with common packages
mkdir baseenv && cd baseenv
zzc analysis                     # Or another profile
zzc github
make docker-build
zzc dockerhub                    # Push for team members
```

**Team Member Alice (Layer 2 - RNA-seq)**:
```bash
git clone https://github.com/genomicslab/rnaseq-study.git
cd rnaseq-study
make docker-build                # Or: docker pull genomicslab/rnaseq-study:latest
make r

# Add RNA-seq specific packages
install.packages(c("tximport", "DESeq2", "apeglm", "EnhancedVolcano"))
# Auto-snapshot on exit captures these packages in renv.lock
q()  # Exit R - snapshot happens automatically
```

**Team Member Bob (Layer 2 - ChIP-seq)**:
```bash
git clone https://github.com/genomicslab/chipseq-study.git
cd chipseq-study
make docker-build
make r

# Add ChIP-seq specific packages
install.packages(c("ChIPseeker", "ChIPpeakAnno", "DiffBind", "csaw"))
# Auto-snapshot on exit captures these packages in renv.lock
q()  # Exit R - snapshot happens automatically
```

**Result**:
- Both use **identical Layer 1** (Bioconductor base via team image)
- Each has **different Layer 2** (analysis-specific packages in renv.lock)
- Complete reproducibility via renv.lock (source of truth)

## Key Principle: renv.lock is Source of Truth

Docker images provide foundation and performance, but **renv.lock ensures reproducibility**:

```bash
# Five years later, reproduce Alice's RNA-seq analysis
git clone https://github.com/genomicslab/rnaseq-study.git
cd rnaseq-study

# Can use ANY compatible Docker base image
zzcollab --profile minimal  # or analysis, or modeling
make r

# renv automatically restores EXACT packages Alice used on R startup
# Auto-restore installs: tximport, DESeq2, apeglm, EnhancedVolcano at exact versions
# (No manual renv::restore() call needed - happens automatically)

# Run analysis - identical results
Rscript analysis/scripts/differential_expression.R
```

**Why this works**: renv.lock specifies exact package versions from CRAN/Bioconductor/GitHub snapshots, independent of Docker image.

## Package Selection Strategy

### Choose Docker Profile (Layer 1) Based On:

1. **Research domain**
   - Genomics → `analysis` + a Bioconductor base image
   - GIS/spatial → `analysis` + the `rocker/geospatial` base image
   - Machine learning → `modeling` profile
   - General analysis → `analysis` profile

2. **Team consensus**
   - What does everyone need?
   - What's the common foundation?

3. **Performance considerations**
   - Large, stable packages → Docker profile (tidyverse, Bioconductor core)
   - Team-wide dependencies → Docker profile

### Add via renv (Layer 2) Based On:

1. **Analysis-specific needs**
   - Domain-specific packages (tximport, ChIPseeker)
   - Specialized statistical methods

2. **Experimental packages**
   - Newly released packages
   - Development versions from GitHub

3. **Version-specific requirements**
   - Packages requiring exact versions
   - Packages with breaking changes between versions

# Practical Configuration Scenarios

## Scenario 1: Solo Data Scientist

**Goal**: Quick project setup, standard data analysis tools

**Configuration**:
```bash
# One-time setup
zzcollab config init
zzcollab config set team-name "datasci"
zzcollab config set profile-name "analysis"

# Create projects with configured defaults
mkdir customer-churn && cd customer-churn
zzcollab  # Uses all configured defaults

mkdir sales-forecast && cd sales-forecast
zzcollab  # Consistent environment across projects
```

**Benefits**:
- No repetitive flag typing
- Consistent environment across projects
- Professional reproducibility for portfolio

## Scenario 2: Academic Research Lab

**Goal**: Team collaboration, consistent environments, specialized workflows

**Team Lead Setup**:
```bash
# Personal config
zzcollab config set team-name "psychlab"
zzcollab config set github-account "psychlab"

# Create shared project
mkdir attention-study && cd attention-study
zzc analysis                     # Full setup (init + renv + docker)
zzc github                       # Create private GitHub repo

# Customize for team
cat > zzcollab.yaml << 'EOF'
team:
  name: "psychlab"
  project: "attention-study"

defaults:
  profile-name: "analysis"

collaboration:
  github:
    auto-create-repo: true
    default-visibility: "private"
    enable-actions: true
EOF

# Build and share
make docker-build
make docker-push-team
git add . && git commit -m "Team setup" && git push
```

**Team Member Workflow**:
```bash
# Minimal personal setup
zzcollab config set team-name "psychlab"

# Join project
git clone https://github.com/psychlab/attention-study.git
cd attention-study
make docker-build

# Start working immediately
make r
```

## Scenario 3: Bioinformatics Core Facility

**Goal**: Multiple specialized environments, organization-wide standards

**System Admin Setup** (`/etc/zzcollab/config.yaml`):
```yaml
organization:
  name: "Bioinformatics Core"
  docker-registry: "registry.core.university.edu"

defaults:
  github-account: "bioinformatics-core"

collaboration:
  github:
    default-visibility: "private"
    enable-actions: true
```

**Core Director Setup**:
```bash
# Base images for different workflows
mkdir biocore-rnaseq && cd biocore-rnaseq
zzc modeling                     # Full setup; add genomics packages via renv
zzc github
make docker-build && make docker-push-team

cd .. && mkdir biocore-chipseq && cd biocore-chipseq
zzc modeling
zzc github
make docker-build && make docker-push-team

cd .. && mkdir biocore-singlecell && cd biocore-singlecell
zzc modeling
zzc github
make docker-build && make docker-push-team
```

**Researcher Workflow**:
```bash
# Join workflow-specific project
git clone https://github.com/biocore/rnaseq-study.git
cd rnaseq-study
make docker-build  # Build from project Dockerfile

# Add analysis-specific packages
make r
install.packages("workflow_specific_package")
# Auto-snapshot on exit captures the package
q()  # Exit R - snapshot happens automatically
```

## Scenario 4: Data Science Consulting Firm

**Goal**: Client projects, quarterly environment updates, reproducible deliverables

**DevOps Team Setup**:
```bash
# Quarterly standardized environment
mkdir q1-2025-base && cd q1-2025-base
zzc modeling                     # Full setup with modeling profile
zzc github

# Document package versions
cat > ENVIRONMENT.md << 'EOF'
# Q1 2025 Standardized Environment

Base: rocker/ml:latest (R 4.4.0)
Profile: modeling

Key packages:
- tidyverse 2.0.0
- tidymodels 1.2.0
- xgboost 1.7.6
- shiny 1.8.0

Last updated: 2025-01-15
EOF

make docker-build
make docker-push-team
git add . && git commit -m "Q1 2025 base environment" && git push
```

**Data Scientist Workflow**:
```bash
# Create client project using quarterly base
mkdir client-churn-prediction && cd client-churn-prediction
zzc modeling                     # Full setup with modeling profile
zzc github

# Team base image provides foundation
make r

# Add client-specific packages
install.packages(c("probably", "workflowsets", "themis"))
# Auto-snapshot on exit captures these packages
q()  # Exit R - snapshot happens automatically

# All projects use same Q1 2025 foundation
# Client-specific packages in renv.lock
```

# Configuration Validation and Troubleshooting

## Validation Commands

### Check Configuration Integrity

```bash
# Validate all configuration files
zzcollab config validate

# Expected output:
# ✓ User configuration valid: ~/.zzcollab/config.yaml
# ✓ No project configuration found (not in project directory)
# ✓ No system configuration found (not installed system-wide)
# ✓ Configuration hierarchy is consistent
```

### View Effective Configuration

```bash
# See merged configuration from all sources
zzcollab config list

# See which file provides each value
zzcollab config list --verbose

# Output:
# team-name: "genomicslab"
#   Source: ~/.zzcollab/config.yaml (user config)
# profile-name: "modeling"
#   Source: ./zzcollab.yaml (project config) [OVERRIDES user setting "analysis"]
```

### Test Configuration

```bash
# Validate the effective configuration before creating a project
zzcollab config validate

# Review the merged configuration and its sources
zzcollab config list
```

## Common Issues and Solutions

### Issue 1: Configuration Hierarchy Confusion

**Problem**: Settings not taking effect as expected

```bash
# Symptom: Set profile-name in user config but project uses different profile
zzcollab config set profile-name "analysis"
zzcollab config get profile-name
# Output: modeling  (Expected: analysis)
```

**Solution**: Check configuration sources
```bash
zzcollab config list --verbose
# Output shows:
# profile-name: "modeling"
#   Source: ./zzcollab.yaml (project config) [OVERRIDES user setting]

# Project config (./zzcollab.yaml) overrides user config
# Remove from project config or understand this is intentional
vim ./zzcollab.yaml  # Remove profile-name to use user default
```

### Issue 2: Docker Platform Compatibility

**Problem**: An ARM64 Mac cannot build an AMD64-only base image.

```bash
# Error when building on the rocker/geospatial base image:
# Error: rocker/geospatial:latest: no matching manifest for linux/arm64/v8
```

**Solution 1**: Use an ARM64-compatible base and add packages via renv
```bash
zzcollab analysis
# Then add spatial packages inside the container:
# install.packages(c("sf", "terra", "raster"))
# (Auto-snapshot on exit saves these to renv.lock)
```

**Solution 2**: Force the AMD64 platform (runs under emulation)
```bash
export DOCKER_DEFAULT_PLATFORM=linux/amd64
zzcollab docker --base-image rocker/geospatial
# Slower under emulation, but compatible
```

### Issue 3: Missing Configuration File

**Problem**: Configuration file not found

```bash
# Error: Configuration file not found: ~/.zzcollab/config.yaml
```

**Solution**: Initialize configuration
```bash
zzcollab config init
# Interactive prompts:
# Team name: myteam
# GitHub account: myusername
# Docker profile (default: analysis):

# Configuration created at: ~/.zzcollab/config.yaml
```

### Issue 4: Docker Daemon Not Running

**Problem**: Cannot connect to Docker daemon

```bash
# Error: Cannot connect to the Docker daemon at unix:///var/run/docker.sock
```

**Solution**: Start Docker daemon
```bash
# macOS/Windows: Start Docker Desktop app

# Linux: Start Docker service
sudo systemctl start docker
sudo systemctl enable docker  # Auto-start on boot

# Verify Docker is running
docker info
```

### Issue 5: Profile Not Found

**Problem**: Custom profile specified but not defined

```bash
# Error: Profile 'my-custom' not found in configuration
```

**Solution**: Define the profile
```bash
# Edit user configuration
vim ~/.zzcollab/config.yaml

# Add custom profile:
profiles:
  my-custom:
    base-image: "rocker/r-ver:latest"
    libs-bundle: "minimal"
    pkgs-bundle: "tidyverse"
    description: "My custom analysis environment"
```

# R Interface Configuration

ZZCOLLAB provides a complete R interface for configuration management.

## Configuration Functions

```{r config-functions}
library(zzcollab)

# Initialize configuration
init_config()

# Set configuration values
set_config("team_name", "myteam")
set_config("profile_name", "analysis")
set_config("github_account", "myusername")

# Get configuration values
team <- get_config("team_name")
profile <- get_config("profile_name")

# List all configuration
all_config <- list_config()
print(all_config)

# Validate configuration
valid <- validate_config()
if (!valid$valid) {
  message("Configuration errors:")
  print(valid$errors)
}
```

## Configuration-Aware Project Functions

```{r config-projects}
# Functions use configuration defaults automatically

# Create project with defaults from config
init_project("my-analysis")
# Uses: team_name, profile_name from config

# Override specific settings
init_project(
  project_name = "special-study",
  profile = "modeling"  # Override default profile
  # Other settings from config
)

# Explicit full control
init_project(
  team_name = "genomicslab",
  project_name = "cancer-study",
  profile = "modeling"
)
```

## Configuration Inspection

```{r config-inspect}
# Get effective configuration from all sources
effective <- get_effective_config()
str(effective)

# Check configuration sources
sources <- get_config_sources()
print(sources$team_name)  # Shows which file provides value

# Get default values
defaults <- get_config_defaults()
print(defaults)
```

# Best Practices

## Configuration Strategy by Use Case

### Solo Researcher
- **Use**: User configuration only (`~/.zzcollab/config.yaml`)
- **Profiles**: 1-2 profiles (minimal, analysis)
- **Package management**: Add via renv as needed
- **Version control**: Commit renv.lock, not user config

### Small Team (2-5 people)
- **Use**: Project configuration (`./zzcollab.yaml` in Git)
- **Profiles**: 2-3 profiles for team domains
- **Package management**: Team discusses additions, uses renv
- **Version control**: Commit project config, renv.lock

### Large Organization (10+ people)
- **Use**: System + project configurations
- **Profiles**: Comprehensive profile library
- **Package management**: IT manages base images, teams use renv
- **Version control**: Centralized profile library + project configs

## Security Considerations

### Private Repository Settings

```yaml
# ./zzcollab.yaml
collaboration:
  github:
    default-visibility: "private"  # Always private
    branch-protection: true         # Require PR reviews

repository:
  allow-merge-commits: false       # Clean history
  allow-squash-merging: true      # Allow squash
  delete-head-branches: true      # Auto-cleanup
```

### Container Security

```yaml
development:
  container:
    default-user: "analyst"        # Never run as root
    read-only-volumes: true        # Mount shared volumes read-only
```

### Sensitive Data

- **Never commit**: API keys, passwords, credentials
- **Use**: Environment variables for secrets
- **Store**: Credentials outside Git (e.g., `~/.Renviron`)

## Performance Optimization

### Build Time Reduction

1. **Use appropriate profiles**
   - Development: `minimal` or `analysis`
   - CI/CD: `minimal`
   - Production: `modeling` or domain-specific

2. **Layer caching**
   - Base image changes → full rebuild
   - Package additions → partial rebuild
   - Use stable base images

3. **Parallel operations**
   - Enable parallel package installation
   - Build profiles concurrently when possible

### Container Size Optimization

```yaml
# Prefer lightweight profiles for CI/CD
profiles:
  minimal:           # ~650MB
    enabled: true
  analysis:          # ~1.2GB (for development)
    enabled: true
  # Avoid enabling unnecessary heavy profiles
```

## Troubleshooting Strategy

1. **Check configuration hierarchy**
   ```bash
   zzcollab config list --verbose
   ```

2. **Validate configuration**
   ```bash
   zzcollab config validate
   ```

3. **Test with dry-run**
   ```bash
   zzcollab config validate
   ```

4. **Check Docker status**
   ```bash
   docker info
   docker images
   ```

5. **Verbose output**
   ```bash
   zzcollab -v analysis
   ```

# Summary

ZZCOLLAB's configuration system provides control over reproducible research environments through multiple layers:

## Key Concepts

1. **Configuration hierarchy**: Command-line → environment → project → user → system → defaults
2. **Docker profiles**: 14+ specialized environments from 200MB to 3.5GB
3. **Two-layer architecture**: Docker profiles (shared) + renv packages (flexible)
4. **CLI flags**: Comprehensive control over all aspects
5. **R interface**: Complete configuration management within R

## Quick Reference

**Initialize configuration**:
```bash
zzcollab config init
zzcollab config set team-name "myteam"
zzcollab config set profile-name "analysis"
```

**Create project**:
```bash
mkdir study && cd study
zzc analysis                     # Full setup (init + renv + docker)
zzc github                       # Create private GitHub repo
```

**Team collaboration**:
```bash
# Team lead:
make docker-build && make docker-push-team

# Team member:
git clone https://github.com/myteam/study.git && cd study
make docker-build
```

**Package management**:
```bash
# Inside container:
install.packages("package")
# Auto-snapshot on exit - no manual snapshot needed
q()  # Exit R - snapshot happens automatically
```

## Further Reading

- `zzcollab --help` - Complete CLI reference
- `help(package = "zzcollab")` - R package documentation
- `vignette("getting-started")` - Comprehensive tutorial
- `docs/VARIANTS.md` - Detailed profile documentation
- `docs/CONFIGURATION.md` - Advanced configuration guide
