---
title: "Quickstart: Collaborative Analysis Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quickstart: Collaborative Analysis Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This guide walks through a complete collaborative analysis workflow using
zzcollab. Two developers will collaborate on an analysis of the `mtcars`
dataset, demonstrating proper data management, testing, reporting, and
version control practices.

> **Key terms in this vignette.** These are the only terms this guide
> leans on; each is reintroduced on first use. Full definitions are in
> `vignette('glossary')`.
>
> - **Compendium**: your project folder, holding data, code, and
>   write-up together so the whole thing can be re-run as a unit.
> - **Archetype**: the project type you pick at creation (analysis,
>   manuscript, package, simulation, or blog); it shapes the starter
>   layout. This guide uses `analysis`.
> - **Profile**: the Docker image bundle the environment is built from
>   (minimal, analysis, rstudio). Distinct from the archetype: the
>   profile is the image, the archetype is the layout.
> - **Feature**: an independently switchable capability (a package
>   backend, Docker, CI, data hashing, code-quality hooks, unit tests,
>   or cloud launch) toggled with `zzc toggle` or `zzc add`/`zzc rm`.
> - **Capture vs validation**: capture features (the backend and Docker)
>   pin the environment and set the capture level; validation features
>   (tests, CI, code quality) only check it.
> - **Capture level (L0 to L3)**: how locked-down the project is, from
>   'just findable' (L0) up to 'rebuilt and verified' (L3).
> - **Team image**: the Docker image the team lead builds with all
>   `renv.lock` packages baked in, so collaborators skip the install
>   wait.
> - **Image digest**: the manifest checksum (`.team-image-digest`) that
>   lets every member pull the byte-identical team image.

**Workflow Summary**:

1. Developer 1 sets up workspace, prepares data, writes initial analysis
2. Developer 1 pushes to GitHub
3. Developer 2 clones, adds visualization, pushes changes
4. Developer 1 pulls changes and finalizes the report

**Prerequisites**:

- Git installed and configured
- Docker installed and running
- GitHub account with SSH keys configured
- zzcollab repository cloned

## Part 1: Initial Setup (Developer 1)

### 1.1 Install zzcollab

After cloning the zzcollab repository, run the installer:

```bash
git clone https://github.com/rgt47/zzcollab.git
cd zzcollab
./install.sh
```

Verify installation (both commands work identically):

```bash
zzc --help
zzcollab --help
```

### 1.2 Configure User Defaults (One-Time Setup)

Set your personal defaults to avoid repeating common options:

```bash
zzc config init
zzc config set github-account "your-github-username"
zzc config set dockerhub-account "your-dockerhub-username"
```

### 1.3 Create Project Workspace

Create a new directory and initialize the compendium (your project
folder) with `zzc init`:

```bash
mkdir mtcars-analysis
cd mtcars-analysis
zzc init
```

`zzc init` asks two things. The first is the **archetype**, the project
type, which shapes the starter layout:

```
Archetype [analysis/manuscript/package/simulation/blog] (default analysis):
```

This tutorial uses `analysis`, which scaffolds a report at
`analysis/report/report.Rmd`. The second is a short feature wizard
(renv backend, Docker, CI, data-integrity hashing, code-quality hooks,
unit tests, cloud launch). It recommends a sensible default of renv plus
Docker, which is what this guide assumes; accept it. The archetype
(project layout) is separate from the **profile** (the Docker image
bundle), covered in the next step.

To skip every prompt and accept defaults, or to set the archetype as a
flag:

```bash
zzc init --yes                   # non-interactive, accept defaults
zzc init --archetype analysis    # set archetype, still run the wizard
```

The full feature menu, and how to toggle features later, is documented
in `vignette("quickstart3")`.

Optionally add GitHub:

```bash
zzc github                       # Create private GitHub repo (independent command)
```

Or create with public GitHub:

```bash
zzc github --public              # Create public GitHub repo
```

This creates the research compendium structure:

```
mtcars-analysis/
├── analysis/
│   ├── data/
│   │   ├── raw_data/
│   │   └── derived_data/
│   ├── figures/
│   ├── report/
│   │   ├── report.Rmd
│   │   └── references.bib
│   └── scripts/
├── R/
├── inst/
│   └── tinytest/
├── tests/
│   └── tinytest.R
├── DESCRIPTION
├── Dockerfile
├── Makefile
├── renv.lock
└── .Rprofile
```

### 1.4 Build the Docker Image

Build the Docker environment:

```bash
make docker-build
```

This creates a reproducible R environment based on the `analysis`
**profile**, the Docker image bundle (rocker/tidyverse with tidyverse
packages). The profile selects the image; the archetype chosen in the
previous step selected the layout. The three canonical profiles are
minimal, analysis, and rstudio; list them with `zzc list profiles`.
For specialised needs (LaTeX, modeling, Shiny), extend the closest
profile.

## Part 2: Data Preparation (Developer 1)

### 2.1 Enter the Container

Start an interactive R session inside the Docker container:

```bash
make r
```

### 2.2 Prepare and Save Raw Data

Inside R, save the complete mtcars dataset as CSV to the raw_data directory:

```r
# Save complete mtcars as CSV (raw data - read-only after this point)
write.csv(mtcars, "analysis/data/raw_data/mtcars.csv", row.names = TRUE)

# Verify the data
dim(mtcars)
# [1] 32 11

# Document the data source
writeLines(
  c(
    "# mtcars.csv",
    "",
    "## Source",
    "Motor Trend Car Road Tests from the 1974 Motor Trend US magazine.",
    "Complete dataset with 32 observations of 11 variables.",
    "",
    "## Variables",
    paste("-", names(mtcars), collapse = "\n"),
    "",
    "## Creation Date",
    as.character(Sys.Date()),
    "",
    "## Citation",
    "Henderson and Velleman (1981). Building multiple regression models",
    "interactively. Biometrics, 37, 391-411."
  ),
  "analysis/data/raw_data/README.md"
)
```

### 2.3 Create Data Processing Script

Create a script that reads the raw CSV and extracts the first half for analysis.

```bash
# Exit R first with q(), then create the script file on host
```

Create `analysis/scripts/01_prepare_data.R`:

```r
# 01_prepare_data.R
# Purpose: Extract first half of mtcars and add derived variables

library(dplyr)

prepare_mtcars_data <- function(input_path, output_path) {
  raw_data <- read.csv(input_path, row.names = 1)

  n_rows <- nrow(raw_data)
  subset_data <- raw_data[1:(n_rows %/% 2), ]

  processed_data <- subset_data |>
    mutate(
      efficiency_class = case_when(
        mpg >= 20 ~ "high",
        mpg >= 15 ~ "medium",
        TRUE ~ "low"
      ),
      weight_kg = wt * 453.592,
      power_to_weight = hp / wt
    )

  saveRDS(processed_data, output_path)
  processed_data
}

if (sys.nframe() == 0) {
  prepare_mtcars_data(
    input_path = "analysis/data/raw_data/mtcars.csv",
    output_path = "analysis/data/derived_data/mtcars_processed.rds"
  )
}
```

### 2.4 Extract Reusable Functions to R/

Move the core function to the package's R/ directory for testing and reuse.

Create `R/data_functions.R`:

```r
#' Prepare mtcars data for analysis
#'
#' @param data A data frame with mtcars structure
#' @return A data frame with additional computed columns
#' @export
add_derived_variables <- function(data) {
  if (!all(c("mpg", "wt", "hp") %in% names(data))) {
    stop("Input must contain mpg, wt, and hp columns")
  }

  data |>
    dplyr::mutate(
      efficiency_class = dplyr::case_when(
        mpg >= 20 ~ "high",
        mpg >= 15 ~ "medium",
        TRUE ~ "low"
      ),
      weight_kg = wt * 453.592,
      power_to_weight = hp / wt
    )
}

#' Validate mtcars data structure
#'
#' @param data A data frame to validate
#' @return TRUE if valid, throws error otherwise
#' @export
validate_mtcars_data <- function(data) {
  required_cols <- c("mpg", "cyl", "disp", "hp", "drat", "wt",
                     "qsec", "vs", "am", "gear", "carb")

  missing_cols <- setdiff(required_cols, names(data))
  if (length(missing_cols) > 0) {
    stop("Missing required columns: ", paste(missing_cols, collapse = ", "))
  }

  if (nrow(data) == 0) {
    stop("Data frame is empty")
  }

  if (any(data$mpg <= 0, na.rm = TRUE)) {
    stop("mpg must be positive")
  }

  if (any(data$hp <= 0, na.rm = TRUE)) {
    stop("hp must be positive")
  }

  TRUE
}
```

## Part 3: Writing Tests (Developer 1)

### 3.1 Create Data Validation Tests

ZZCOLLAB projects use tinytest. Test files live in `inst/tinytest/` and
contain bare top-level expectations (no `test_that()` wrapper).

Create `inst/tinytest/test-data-validation.R`:

```r
# test-data-validation.R
# Tests for data integrity and validation functions

raw_data_path <- 'analysis/data/raw_data/mtcars.csv'

# Raw data file exists and is readable
expect_true(file.exists(raw_data_path))
data <- read.csv(raw_data_path, row.names = 1)
expect_inherits(data, 'data.frame')

# Expected dimensions
expect_equal(nrow(data), 32L)
expect_equal(ncol(data), 11L)

# Required columns present
required_cols <- c('mpg', 'cyl', 'disp', 'hp', 'drat', 'wt',
                   'qsec', 'vs', 'am', 'gear', 'carb')
expect_true(all(required_cols %in% names(data)))

# Values within expected ranges
expect_true(all(data$mpg > 0 & data$mpg < 50))
expect_true(all(data$cyl %in% c(4, 6, 8)))
expect_true(all(data$hp > 0 & data$hp < 500))
expect_true(all(data$wt > 0))

# Derived data has the correct subset size (only if it has been created)
derived_path <- 'analysis/data/derived_data/mtcars_processed.rds'
if (file.exists(derived_path)) {
  expect_equal(nrow(readRDS(derived_path)), 16L)
}
```

### 3.2 Create Function Tests

Create `inst/tinytest/test-data-functions.R`:

```r
# test-data-functions.R
# Tests for R/data_functions.R

# validate_mtcars_data accepts valid data
valid_data <- data.frame(
  mpg = c(21, 22), cyl = c(6, 4), disp = c(160, 140),
  hp = c(110, 93), drat = c(3.9, 3.85), wt = c(2.6, 2.3),
  qsec = c(16.5, 18.6), vs = c(0, 1), am = c(1, 1),
  gear = c(4, 4), carb = c(4, 2)
)
expect_true(validate_mtcars_data(valid_data))

# validate_mtcars_data rejects missing columns
expect_error(validate_mtcars_data(data.frame(mpg = 21, cyl = 6)),
             pattern = 'Missing required columns')

# validate_mtcars_data rejects empty data
expect_error(validate_mtcars_data(mtcars[0, ]), pattern = 'empty')

# validate_mtcars_data rejects invalid mpg
bad_data <- mtcars[1:2, ]
bad_data$mpg[1] <- -5
expect_error(validate_mtcars_data(bad_data),
             pattern = 'mpg must be positive')

# add_derived_variables creates the expected columns
result <- add_derived_variables(mtcars[1:5, ])
expect_true('efficiency_class' %in% names(result))
expect_true('weight_kg' %in% names(result))
expect_true('power_to_weight' %in% names(result))

# add_derived_variables calculates efficiency_class correctly
ec_data <- data.frame(
  mpg = c(25, 17, 12),
  wt = c(2.5, 3.0, 3.5),
  hp = c(100, 150, 200)
)
expect_equal(add_derived_variables(ec_data)$efficiency_class,
             c('high', 'medium', 'low'))

# add_derived_variables calculates weight_kg correctly
kg_result <- add_derived_variables(data.frame(mpg = 20, wt = 1, hp = 100))
expect_equal(kg_result$weight_kg, 453.592, tolerance = 0.001)

# add_derived_variables requires the correct input columns
expect_error(add_derived_variables(data.frame(x = 1, y = 2)),
             pattern = 'must contain mpg, wt, and hp')
```

### 3.3 Run Tests

Run the tinytest suite:

```bash
make docker-test
```

Expected output (each expectation reported as a passing check):

```
inst/tinytest/test-data-functions.R............    8 tests OK
inst/tinytest/test-data-validation.R...........    7 tests OK
```

To run interactively inside the container instead:

```bash
make r
```

```r
# Inside R
tinytest::run_test_dir('inst/tinytest')
```

Exit R when done:

```r
q()
```

Packages are automatically captured to `renv.lock` when you exit R (auto-snapshot). Auto-restore on startup is disabled in the container; packages are baked into the image at build time.

## Part 4: Initial Analysis Report (Developer 1)

### 4.1 Create the Report

Edit `analysis/report/report.Rmd`:

```{r, eval=FALSE, echo=FALSE}
# The content below should be saved to analysis/report/report.Rmd
```

````
---
title: "Fuel Efficiency Analysis: mtcars Subset"
author: "Developer 1"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: references.bib
---

`r ''````{r report-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

# Load package functions from R/ directory
devtools::load_all(here::here())
```

## Introduction

This analysis examines fuel efficiency patterns in a subset of the Motor
Trend Car Road Tests dataset. The raw data contains all 32 observations;
the derived dataset extracts the first 16 observations for analysis.

## Data Loading

```{r load-data}
library(dplyr)
library(ggplot2)
library(knitr)

# Load processed data
data <- readRDS(here::here("analysis/data/derived_data/mtcars_processed.rds"))

# Display summary
kable(head(data), caption = "First observations of processed data")
```

## Summary Statistics

```{r summary-stats}
summary_stats <- data |>
  group_by(efficiency_class) |>
  summarise(
    n = n(),
    mean_mpg = mean(mpg),
    mean_hp = mean(hp),
    mean_weight_kg = mean(weight_kg),
    .groups = "drop"
  )

kable(summary_stats,
      digits = 2,
      caption = "Summary statistics by efficiency class")
```

## Analysis Results

### Fuel Efficiency Distribution

```{r mpg-histogram}
ggplot(data, aes(x = mpg, fill = efficiency_class)) +
  geom_histogram(bins = 8, alpha = 0.7, color = "white") +
  labs(
    title = "Distribution of Fuel Efficiency",
    x = "Miles Per Gallon",
    y = "Count",
    fill = "Efficiency Class"
  ) +
  theme_minimal()
```

<!-- Developer 2 will add additional visualizations here -->

## Conclusions

*To be completed after collaborative additions.*

## References
````

### 4.2 Run the Analysis Pipeline

Enter container and run the processing script:

```bash
make r
```

```r
source("analysis/scripts/01_prepare_data.R")
```

Verify the derived data was created:

```r
file.exists("analysis/data/derived_data/mtcars_processed.rds")
# [1] TRUE

q()
```

### 4.3 Validate Package Dependencies

After exiting R, validate that all packages are properly tracked:

```bash
make check-renv
```

This runs `zzrenvcheck::check_packages()` inside the container to verify that DESCRIPTION and renv.lock are synchronized with the code's actual `library()` calls.

## Part 5: Push to GitHub (Developer 1)

### 5.1 Initialize Git and Create GitHub Repository

If you haven't created the GitHub repo yet, do it now:

```bash
zzc github                       # Initialize git + create private GitHub repo
```

Or for a public repo:

```bash
zzc github --public              # Create public GitHub repo
```

Or manually:

```bash
git init
git add .
git commit -m "Initial project setup with mtcars analysis

- Add raw data (complete mtcars CSV)
- Create data processing script (extracts first half)
- Add data validation and function tests
- Create initial report structure"

gh repo create myteam/mtcars-analysis --private --source=. --push
```

### 5.2 Build and Push Team Docker Image

```bash
make docker-build

# Multi-arch push (amd64 + arm64) and records digest to .team-image-digest:
make docker-push-team

# Commit the digest so team members can pull the exact same image:
git add .team-image-digest
git commit -m "Record team image digest"
git push
```

`make docker-push-team` builds a multi-architecture image (linux/amd64 and
linux/arm64), pushes it to Docker Hub, and writes the manifest digest to
`.team-image-digest`. Committing that file lets team members run
`make docker-pull-team` to get the byte-identical image.

`zzc dockerhub` is a simpler alternative that pushes a single-architecture
image without recording the digest.

## Part 6: Add Visualization (Developer 2)

### 6.1 Clone the Repository

Developer 2 clones the project and loads the team Docker image:

```bash
git clone git@github.com:myteam/mtcars-analysis.git
cd mtcars-analysis

# Bring framework-managed template files in line with the installed
# zzcollab version, and back-fill local state.
zzc update

# Preferred: pull the exact digest the team lead recorded (bit-identical)
make docker-pull-team          # reads .team-image-digest committed by team lead

# Alternative: build from Dockerfile (always works, takes 5-15 min)
make docker-build
```

`zzc update` is a minor maintenance command. It checks the project's
framework-managed template files (Makefile, CI workflows, and similar)
against the installed zzcollab version, refreshes any that are stale,
and back-fills the state record. Running it right after a clone keeps a
collaborator's framework files consistent with the team lead's, without
touching analysis code or primary feature files. Confirm the result
with `zzc status` (read-only) and `zzc verify` (coherence checks).

### 6.2 Enter Container and Verify Setup

Confirm the framework state before entering the container:

```bash
zzc status                     # read-only: shows features and capture level
zzc verify                     # coherence checks (files agree with state)
```

```bash
make r
```

```r
# Verify data is present
readRDS("analysis/data/derived_data/mtcars_processed.rds") |> head()

# Run existing tests to ensure environment is correct
tinytest::run_test_dir('inst/tinytest')

q()
```

### 6.3 Add Visualization Function

Developer 2 adds a new plotting function. Create `R/plot_functions.R`:

```r
#' Create power-to-weight scatter plot
#'
#' @param data Processed mtcars data frame
#' @param show_labels Logical, whether to show car labels
#' @return A ggplot object
#' @export
plot_power_weight <- function(data, show_labels = TRUE) {
  if (!"power_to_weight" %in% names(data)) {
    stop("Data must contain power_to_weight column")
  }

  p <- ggplot2::ggplot(
    data,
    ggplot2::aes(x = weight_kg, y = hp, color = efficiency_class)
  ) +
    ggplot2::geom_point(size = 3, alpha = 0.8) +
    ggplot2::labs(
      title = "Horsepower vs Weight by Efficiency Class",
      x = "Weight (kg)",
      y = "Horsepower",
      color = "Efficiency"
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(legend.position = "bottom")

  if (show_labels) {
    p <- p + ggplot2::geom_text(
      ggplot2::aes(label = rownames(data)),
      hjust = -0.1,
      vjust = 0.5,
      size = 3
    )
  }

  p
}

#' Create cylinder comparison boxplot
#'
#' @param data Processed mtcars data frame
#' @return A ggplot object
#' @export
plot_mpg_by_cylinder <- function(data) {
  ggplot2::ggplot(data, ggplot2::aes(x = factor(cyl), y = mpg)) +
    ggplot2::geom_boxplot(fill = "steelblue", alpha = 0.7) +
    ggplot2::geom_jitter(width = 0.2, alpha = 0.5) +
    ggplot2::labs(
      title = "Fuel Efficiency by Number of Cylinders",
      x = "Number of Cylinders",
      y = "Miles Per Gallon"
    ) +
    ggplot2::theme_minimal()
}
```

### 6.4 Add Tests for New Functions

Create `inst/tinytest/test-plot-functions.R`:

```r
# test-plot-functions.R
# Tests for R/plot_functions.R

# plot_power_weight returns a ggplot object
pw_data <- mtcars[1:5, ] |> add_derived_variables()
expect_inherits(plot_power_weight(pw_data), 'ggplot')

# plot_power_weight requires the power_to_weight column
expect_error(plot_power_weight(mtcars[1:5, ]),
             pattern = 'power_to_weight')

# plot_power_weight respects the show_labels parameter
expect_inherits(plot_power_weight(pw_data, show_labels = TRUE), 'ggplot')
expect_inherits(plot_power_weight(pw_data, show_labels = FALSE), 'ggplot')

# plot_mpg_by_cylinder returns a ggplot object
expect_inherits(plot_mpg_by_cylinder(mtcars[1:5, ]), 'ggplot')
```

### 6.5 Run All Tests

```bash
make docker-test
```

Expected output (all three test files passing):

```
inst/tinytest/test-data-functions.R............    8 tests OK
inst/tinytest/test-data-validation.R...........    7 tests OK
inst/tinytest/test-plot-functions.R............    5 tests OK
```

### 6.6 Commit and Push Changes

```bash
git add R/plot_functions.R inst/tinytest/test-plot-functions.R
git commit -m "Add visualization functions for power-weight analysis

- Add plot_power_weight() for scatter plots
- Add plot_mpg_by_cylinder() for boxplots
- Include tests for both plotting functions"

git push
```

## Part 7: Finalize Report (Developer 1)

### 7.1 Pull Changes

Developer 1 pulls Developer 2's additions, then refreshes
framework-managed files and confirms state:

```bash
git pull
zzc update                     # refresh stale template files, back-fill state
zzc status                     # confirm features and capture level
```

### 7.2 Enter Container

```bash
make r
```

Verify new packages are auto-restored:

```r
tinytest::run_test_dir('inst/tinytest')
# All three test files should pass

q()
```

### 7.3 Update the Report

Edit `analysis/report/report.Rmd` to include the new visualizations.
Add the following sections after the existing histogram:

````markdown
### Power vs Weight Analysis

```{r power-weight-plot}
# Functions loaded via devtools::load_all() in setup chunk
plot_power_weight(data, show_labels = TRUE)
```

This scatter plot reveals the relationship between engine power and vehicle
weight. Higher efficiency vehicles (green) cluster in the lower-left region,
indicating lower weight and moderate horsepower.

### Cylinder Comparison

```{r cylinder-boxplot}
plot_mpg_by_cylinder(data)
```

The boxplot demonstrates the expected inverse relationship between cylinder
count and fuel efficiency. Four-cylinder vehicles show the highest MPG values,
while eight-cylinder vehicles demonstrate lower fuel efficiency.
````

Update the Conclusions section:

````markdown
## Conclusions

This analysis of the mtcars subset reveals several key findings:

1. **Efficiency Classification**: The dataset contains a mix of low, medium,
   and high efficiency vehicles based on the 15 and 20 MPG thresholds.

2. **Weight-Power Relationship**: There is a clear positive correlation
   between vehicle weight and horsepower, with efficiency class serving as
   a useful categorization.

3. **Cylinder Effect**: Vehicles with fewer cylinders consistently achieve
   better fuel economy, supporting the expected engineering trade-off between
   power and efficiency.

## Session Information

```{r session-info}
sessionInfo()
```
````

### 7.4 Render the Report

```bash
make r
```

```r
rmarkdown::render("analysis/report/report.Rmd")
q()
```

### 7.5 Final Validation

Run the complete test suite:

```bash
make docker-test
```

Validate package environment:

```bash
make check-renv
```

### 7.6 Commit Final Version

```bash
git add analysis/report/report.Rmd analysis/report/report.html
git commit -m "Finalize analysis report with collaborative visualizations

- Integrate power-weight scatter plot
- Add cylinder comparison boxplot
- Write conclusions section
- Add session info for reproducibility"

git push
```

## Part 8: Team Collaboration Details

This section covers the mechanics of inviting collaborators and sharing Docker
images. These details were glossed over in the main workflow.

### 8.1 Inviting Collaborators to GitHub

**Adding a single collaborator:**

```bash
# Add collaborator with write access (can push commits)
gh repo edit --add-collaborator username

# Verify collaborators
gh api repos/:owner/:repo/collaborators --jq '.[].login'
```

**Adding multiple collaborators:**

```bash
# Add several collaborators
for user in alice bob charlie; do
  gh repo edit --add-collaborator "$user"
done
```

**Permission levels** (from least to most access):

| Level | Capabilities |
|-------|-------------|
| read | View code, clone, open issues |
| triage | + Manage issues and PRs |
| write | + Push commits, merge PRs |
| maintain | + Manage repo settings (not destructive) |
| admin | Full access including delete |

```bash
# Add with specific permission (default is 'write' for user repos)
gh api repos/:owner/:repo/collaborators/username -X PUT -f permission=maintain
```

**What happens after invitation:**

1. Collaborator receives email notification
2. They must accept the invitation (check email or github.com/notifications)
3. Once accepted, they can clone and push based on permission level

**Checking pending invitations:**

```bash
gh api repos/:owner/:repo/invitations --jq '.[].invitee.login'
```

### 8.2 DockerHub Image Sharing (Optional but Efficient)

**DockerHub is optional.** Collaborators can always build from the Dockerfile:

```bash
# Option A: Build from Dockerfile (always works, takes 5-15 minutes)
git clone https://github.com/myteam/study.git
cd study
make docker-build
make r
```

**Why use DockerHub?** Pre-installed packages save time:

When the team lead builds the image, all packages from `renv.lock` are
installed into the image. Team members who pull this image get those
packages immediately - no installation wait.

| Approach | What Happens | Time |
|----------|-------------|------|
| Build from Dockerfile | Downloads base + installs all packages | 5-15 min |
| Pull from DockerHub | Downloads pre-built image with packages | 1-3 min |

```bash
# Option B: Pull pre-built image pinned by digest (faster, requires DockerHub setup)
git clone https://github.com/myteam/study.git
cd study
make docker-pull-team    # reads .team-image-digest; byte-identical for all members
make r
```

### 8.3 Setting Up DockerHub (Team Lead)

**One-time DockerHub setup:**

1. Create account at hub.docker.com
2. Create repository (can be public or private)
3. Login from terminal:

```bash
docker login
# Enter DockerHub username and password/token
```

**Build before pushing (important):**

The Docker build process installs all R packages from `renv.lock` into the
image. This is what makes pulling faster for team members - packages are
pre-installed rather than installed on first run.

```bash
# 1. Build image (installs packages from renv.lock into /opt/renv/library)
make docker-build

# 2. Push multi-arch image and record digest (preferred)
make docker-push-team            # writes .team-image-digest; commit this file

# Alternative: single-arch push without digest recording
zzc dockerhub                    # Pushes to dockerhub-account/project:latest
zzc dockerhub --tag v1.0         # Or with an explicit version tag
```

**When to rebuild and push:**

- After adding packages (`renv.lock` changed)
- After modifying Dockerfile
- Before onboarding new team members (ensure they get current image)

```bash
# Check if renv.lock changed since last push
git log --oneline -3 -- renv.lock

# If changed, rebuild and push
make docker-build && zzc dockerhub
```

### 8.4 Private DockerHub Repositories

For private research, use private DockerHub repos:

**Team lead creates private repo:**

1. Go to hub.docker.com → Create Repository
2. Set visibility to "Private"
3. Add collaborators under "Collaborators" tab

**Team members authenticate:**

```bash
# Each team member must login once
docker login
# Enter their DockerHub credentials

# Then pull works
docker pull myteam/private-study:latest
```

**Free tier limitations:**

- Docker Free: 1 private repo, unlimited public
- Docker Pro: Unlimited private repos

For academic teams, public repos are often sufficient since the code is
on private GitHub anyway.

### 8.5 When to Rebuild vs Pull

**Pull the team image when:**

- First time joining the project
- Team lead announces image update
- Dockerfile hasn't changed since last pull

**Rebuild from Dockerfile when:**

- Dockerfile has been modified
- You need to verify reproducibility
- CI/CD (always builds fresh)
- Troubleshooting environment issues

**Check if rebuild needed:**

```bash
# See if Dockerfile changed since last pull
git log --oneline -5 -- Dockerfile

# If changes exist, rebuild
make docker-build
```

### 8.6 Complete Team Workflow Example

**Team Lead (one-time setup):**

```bash
# Create project (archetype + feature wizard; --yes accepts defaults)
mkdir team-analysis && cd team-analysis
zzc init --yes --archetype analysis
zzc github

# Build and push multi-arch image; records digest to .team-image-digest
docker login                     # One-time DockerHub authentication
make docker-push-team
git add .team-image-digest && git commit -m "Record team image digest" && git push

# Add collaborators
gh repo edit --add-collaborator alice
gh repo edit --add-collaborator bob
```

**Team Member (joining):**

```bash
# Clone repository
git clone https://github.com/leader/team-analysis.git
cd team-analysis

# Sync framework-managed files with the installed zzcollab version
zzc update

# Option A: Pull image pinned to the committed digest (fast, reproducible)
docker login                     # If private DockerHub repo
make docker-pull-team            # reads .team-image-digest
make r

# Option B: Build from Dockerfile (always works, verifies reproducibility)
make docker-build
make r
```

**Daily workflow (all team members):**

```bash
git pull                         # Get latest code and any updated digest
zzc update                       # Refresh stale framework files after a pull
make r                           # Enter container (packages baked in)
# ... work, install new packages with install.packages() if needed ...
exit                             # Auto-snapshot writes updated renv.lock
git add . && git commit && git push
```

## Summary

This workflow demonstrated:

1. **Project Setup**: Creating a zzcollab workspace with proper structure
2. **Data Management**: Storing raw data in `analysis/data/raw_data/` with
   documentation
3. **Code Organization**: Separating reusable functions into `R/` directory
4. **Testing**: Writing tests for both data validation and functions
5. **Collaboration**: Two developers working on the same codebase
6. **Docker**: Using containerized environments for reproducibility
7. **Version Control**: Proper commit messages and GitHub workflow
8. **Team Onboarding**: Inviting collaborators and sharing Docker images

The final repository contains:

- Reproducible Docker environment
- Tested and documented R functions
- Complete analysis report with visualizations
- Full version history of collaborative work

## Quick Reference

| Task | Command |
|------|---------|
| Create project (archetype + feature wizard) | `zzc init` |
| Create project non-interactively | `zzc init --yes --archetype analysis` |
| Add GitHub repo | `zzc github` |
| Refresh framework files after a clone or pull | `zzc update` |
| Show features and capture level (read-only) | `zzc status` |
| Verify coherence (L3 with `--full`) | `zzc verify` |
| Add or remove a feature | `zzc add <f>` / `zzc rm <f>` |
| Toggle features interactively | `zzc toggle` |
| Enter container | `make r` |
| Run tests | `make docker-test` |
| Validate packages | `make check-renv` |
| Build Docker image | `make docker-build` |
| Push multi-arch image + record digest | `make docker-push-team` |
| Pull image pinned by digest | `make docker-pull-team` |
| Push single-arch (simpler) | `zzc dockerhub` |
| Push with tag | `zzc dockerhub --tag v1.0` |
| Hash raw data for integrity | `make hash-data` |
| Verify raw data against manifest | `make verify-data` |
| Change base image | `zzcollab docker --base-image rocker/verse` |
| List profiles | `zzc list profiles` |
| Render report | `make docker-render` |

## Next Steps

- Add CI/CD with GitHub Actions for automated testing
  (`zzc add cicd`); the feature toggle model is covered in
  `vignette("quickstart3")`.
- Implement additional analyses on the second half of mtcars
- Create a package vignette documenting the analysis methodology
- Add code coverage reporting with covr

## See also

- `vignette("quickstart3")` for the feature-toggle model in depth: the
  capture-versus-validation distinction, capture levels (L0 to L3), and
  the full `zzc toggle` / `zzc add` / `zzc rm` menu.
