---
title: "Quick Start: RStudio Server Workflow"
author: "ZZCOLLAB Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Quick Start: RStudio Server Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

This quick start guide demonstrates how to use **RStudio Server** inside a Docker container for reproducible analysis. You'll work in a familiar RStudio interface while maintaining complete reproducibility through Docker and renv.

> **Key terms in this vignette.** New to the vocabulary? These are the
> terms this guide uses; each is defined again on first use. Full
> definitions are in `vignette('glossary')`.
>
> - **Profile**: the Docker image bundle the environment is built from;
>   here the `rstudio` profile (RStudio Server in the browser).
> - **Docker / container / image**: a tool that packages software with
>   its environment; a running instance and its template.
> - **renv**: an R package giving each project its own isolated,
>   version-pinned package library.
> - **Continuous integration (CI)**: automatically building and testing
>   code on every push.
> - **Unit test / testthat / tinytest**: a focused check that one piece
>   of code behaves as intended.
> - **Base image**: the starting Docker image a project builds on
>   (e.g. `rocker/rstudio`).
> - **`zzc` / `zzcollab`**: the framework's command-line interface;
>   `zzc` is an alias for `zzcollab`.

**What you will build:**

- RStudio Server environment accessible via browser
- Scatter plot: bill length vs bill depth (Palmer Penguins)
- Complete reproducibility: renv + Docker + Unit Testing + CI/CD
- Private GitHub repository with automated validation

**Why RStudio Profile:**

- Familiar RStudio IDE interface in browser
- No local R installation needed
- Point-and-click package installation
- Visual debugging and data exploration
- Works on any machine with Docker and a browser

**Prerequisites:**

- Docker installed and running
- Web browser
- GitHub account (example: rgt47)
- zzcollab installed (`./install.sh`)

# Step-by-Step Instructions

## Step 1: Configure ZZCOLLAB for RStudio (30 seconds)

Set your defaults once, specifying the RStudio profile:

```bash
# Initialize configuration
zzcollab config init

# Set team name (typically your GitHub username for solo projects)
zzcollab config set team-name "rgt47"

# Set GitHub account
zzcollab config set github-account "rgt47"

# Set RStudio profile as default
zzcollab config set profile-name "rstudio"
```

**What this accomplishes:**

- RStudio Server will be available in all future projects
- Includes R, RStudio Server, essential packages
- ~980MB Docker image (more than minimal, less than full analysis)

## Step 2: Create Project with RStudio Profile (3-4 minutes)

Create your reproducible analysis project:

```bash
# Create project directory and initialize with the rstudio profile
mkdir penguin-bills && cd penguin-bills
zzc rstudio                      # Full setup (init + renv + docker)
zzc github                       # Create private GitHub repo (optional)

# Build Docker image with RStudio Server
make docker-build
```

The profile can also come from your config (Step 1 set `profile-name` to
`rstudio`), in which case `zzc init` selects it automatically.

**What this does:**

1. Creates complete project structure (R/, analysis/, tests/, .github/, etc.)
2. Sets up CI/CD workflows for automated testing
3. Initializes renv for dependency tracking
4. Builds Docker image with RStudio Server included

## Step 3: Start RStudio Server (10 seconds)

Launch RStudio Server in your browser:

```bash
make docker-rstudio
```

**Expected output:**
```
Starting RStudio Server on http://localhost:8787
Username: rstudio, Password: rstudio
```

**Access RStudio:**

1. Open browser to **http://localhost:8787**
2. Login with:
   - Username: `rstudio`
   - Password: `rstudio`
3. You're now in RStudio running inside Docker!

## Step 4: Create Analysis in RStudio (5 minutes)

Work directly in the RStudio interface:

### 4.1 Create Analysis Function

In RStudio, create file `R/bill_analysis.R`:

**File → New File → R Script**

```r
#' Create scatter plot of bill dimensions
#'
#' @param data Palmer penguins data
#' @return ggplot object
#' @export
create_bill_plot <- function(data = palmerpenguins::penguins) {
  data %>%
    dplyr::filter(!is.na(bill_length_mm), !is.na(bill_depth_mm)) %>%
    ggplot2::ggplot(ggplot2::aes(x = bill_length_mm, y = bill_depth_mm,
                                  color = species)) +
    ggplot2::geom_point(size = 3, alpha = 0.7) +
    ggplot2::labs(
      title = 'Palmer Penguins: Bill Dimensions',
      x = 'Bill Length (mm)',
      y = 'Bill Depth (mm)',
      color = 'Species'
    ) +
    ggplot2::theme_minimal()
}
```

Save as: `R/bill_analysis.R`

### 4.2 Create Analysis Report

In RStudio, create file `analysis/report/report.Rmd`:

**File → New File → R Markdown...**

````
---
title: "Palmer Penguins Bill Dimensions Analysis"
author: "Reproducible Research Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

Analysis of bill length and bill depth measurements across three penguin species
from the Palmer Station Antarctica LTER.

## Data Loading

```{r load-data}
library(palmerpenguins)
library(ggplot2)
library(dplyr)

# Source analysis functions from R/
source("../../R/bill_analysis.R")

# Preview data
head(penguins)
```

## Analysis

### Bill Dimensions by Species

```{r bill-plot, fig.width=8, fig.height=6}
# Generate plot using function from R/
plot <- create_bill_plot()
print(plot)

# Save plot
ggsave("../figures/bill_scatter.png", plot, width = 8, height = 6)
```

## Summary

The analysis reveals distinct bill dimension patterns across penguin species.
```
````

Save as: `analysis/report/report.Rmd`

### 4.3 Install Packages in RStudio

Use standard R commands (packages auto-captured when you exit container):

```r
# In RStudio Console:
install.packages("palmerpenguins")
install.packages("ggplot2")
install.packages("dplyr")

# For GitHub packages:
install.packages("remotes")
remotes::install_github("user/package")

# When you stop the RStudio container, packages are automatically saved to renv.lock
# No manual renv::snapshot() needed!
```

**RStudio Advantage:** You can also use the Packages pane:

1. Click "Packages" tab (bottom-right)
2. Click "Install"
3. Type package name
4. Click "Install"
5. Packages automatically captured when you exit container (no manual snapshot needed)

### 4.4 Render Report in RStudio

Render the analysis report:

**Method 1 - Click "Knit" button** in `report.Rmd`

**Method 2 - Console command:**
```r
rmarkdown::render("analysis/report/report.Rmd")
```

Your HTML report appears in `analysis/report/paper.html`!

## Step 5: Add Unit Tests in RStudio (2 minutes)

ZZCOLLAB projects use tinytest. Test files live in `inst/tinytest/` and
contain bare top-level expectations (no `test_that()` wrapper).

### 5.1 Create Test File

**File → New File → R Script**

```r
# inst/tinytest/test-bill_analysis.R
# Source function from R/
source('../../R/bill_analysis.R')

# create_bill_plot produces a valid ggplot with the expected title
plot <- create_bill_plot()
expect_inherits(plot, 'ggplot')
expect_equal(plot$labels$title, 'Palmer Penguins: Bill Dimensions')

# create_bill_plot handles missing data
test_data <- palmerpenguins::penguins
test_data$bill_length_mm[1:5] <- NA
plot_na <- create_bill_plot(test_data)
expect_inherits(plot_na, 'ggplot')

# create_bill_plot includes species in the color aesthetic
expect_true('species' %in% names(plot$labels))
```

Save as: `inst/tinytest/test-bill_analysis.R`

### 5.2 Run Tests

From the terminal:
```bash
make docker-test
```

Or in the RStudio Console:
```r
tinytest::run_test_dir('inst/tinytest')
```

## Step 6: Exit RStudio and Validate (1 minute)

Close your browser tab, then in terminal:

```bash
# Stop RStudio container (Ctrl+C in terminal where docker-rstudio is running)
# Auto-snapshot runs when container exits - packages saved to renv.lock!

# Run tests in clean environment
make docker-test
# Auto-restore runs when R starts - all packages installed automatically!

# Validate dependencies (runs zzrenvcheck in the container)
make check-renv
```

## Step 7: Commit and Push (1 minute)

```bash
git add .
git commit -m "$(cat <<'EOF'
Add bill dimensions scatter plot analysis

Analysis:
- Scatter plot of bill length vs bill depth by species
- Palmer Penguins dataset
- Function-based implementation with unit tests
- Developed in RStudio Server

Reproducibility:
- Level 2: renv dependency tracking
- Level 3: Docker environment isolation (rstudio profile)
- Level 4: Unit tests for computational correctness
- Level 5: CI/CD automated validation
EOF
)"

git push
```

## Step 8: Verify CI/CD (30 seconds)

```bash
# View CI/CD status
gh run list --limit 3

# Or view in browser
gh repo view --web
```

# RStudio Workflow Advantages

## Point-and-Click Development

**Visual Interface Benefits:**

- File browser for easy navigation
- Environment pane shows all objects
- Plots pane displays visualizations
- Help pane with integrated documentation
- Package manager with search

## Interactive Analysis

**Real-time Exploration:**
```r
# In RStudio Console - explore interactively:
library(palmerpenguins)
View(penguins)  # Opens data viewer
summary(penguins)
table(penguins$species)

# Test plot interactively
library(ggplot2)
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point()
```

## Debugging Support

**Visual Debugging:**

1. Set breakpoints (click left of line number)
2. Click "Source" to run with debugging
3. Inspect variables in Environment pane
4. Step through code line by line

## RMarkdown Integration

**Document Authoring:**

- Visual markdown editor (RStudio 1.4+)
- Live preview as you type
- Insert code chunks with shortcuts
- Knit to HTML/PDF/Word with one click

# Accessing RStudio Server

## Starting RStudio

```bash
# From project directory:
make docker-rstudio
```

Opens on: **http://localhost:8787**

## Default Credentials

- **Username:** `rstudio`
- **Password:** `rstudio`

## Customizing Credentials

Edit your project's Makefile to change:

```makefile
docker-rstudio:
	@echo "Starting RStudio Server on http://localhost:8787"
	@echo "Username: myuser, Password: mypass"
	docker run --platform linux/amd64 --rm -p 8787:8787 \
	  -v $$(pwd):/project \
	  -e USER=myuser \
	  -e PASSWORD=mypass \
	  $(PACKAGE_NAME) /init
```

## Persistent Sessions

RStudio session is temporary - work is saved but R workspace isn't:

- **Saved:** All file edits, created files, installed packages (auto-captured in
  renv.lock on exit)
- **Not saved:** R workspace (.RData), unless you explicitly save

**Best Practice:** Save work via files, not R workspace:
```r
# Don't rely on:
save.image()  # Workspace not portable

# Instead use:
saveRDS(my_data, "data/my_data.rds")  # Portable, reproducible
# Packages auto-saved when you exit container - no manual renv::snapshot() needed
```

# Transitioning to Command Line

## When to Use Terminal

RStudio is great for development, but some tasks need terminal:

**Use Terminal for:**

- Running tests: `make docker-test`
- Capture dependencies: `make snapshot` (renv::hydrate + snapshot +
  zzrenvcheck; grows renv.lock and DESCRIPTION before a commit)
- Validation: `make check-renv` (runs zzrenvcheck in the container)
- Git operations: `git commit`, `git push`
- Building images: `make docker-build`
- Note: Auto-snapshot and auto-restore run automatically

**Use RStudio for:**

- Writing analysis code
- Creating visualizations
- Authoring reports
- Exploratory analysis
- Package development

## Hybrid Workflow

**Typical workflow:**

1. **RStudio:** Write and test code interactively
2. **RStudio:** Install packages with standard `install.packages()`
3. **RStudio:** Author report and render locally
4. **Exit RStudio:** Auto-snapshot captures packages in renv.lock
5. **Terminal:** Run full test suite (auto-restore installs packages)
6. **Terminal:** Validate dependencies with `make check-renv` (zzrenvcheck)
7. **Terminal:** Commit and push
8. **CI/CD:** Automated validation runs

# Team Collaboration with RStudio

## Team Lead Setup

```bash
# Set defaults
zzcollab config set team-name "genomicslab"
zzcollab config set profile-name "analysis"

# Create team project
mkdir study && cd study
zzc analysis                     # Full setup (init + renv + docker)
zzc github                       # Create private GitHub repo

# Build and share team image (installs packages from renv.lock)
make docker-build
make docker-push-team            # Optional but efficient for team

# Commit project structure
git add .
git commit -m "Initial project setup with RStudio Server"
git push -u origin main
```

## Team Member Workflow

```bash
# Clone repository
git clone https://github.com/genomicslab/study.git
cd study

# Build Docker image from project Dockerfile
make docker-build

# Start RStudio Server
make docker-rstudio

# Access http://localhost:8787
# Work in identical RStudio environment as team
```

**Note**: DockerHub is optional but efficient. Building from Dockerfile produces
an identical environment if team image is not available.

**Team Benefits:**

- Everyone gets identical RStudio version
- Same R version across team
- Consistent package versions (via renv)
- No "works on my machine" issues

# Troubleshooting

## RStudio Won't Start

```bash
# Check Docker is running
docker info

# Check port 8787 is free
lsof -i :8787

# If port busy, use different port:
docker run --rm -p 8888:8787 -v $(pwd):/project \
  -e USER=analyst -e PASSWORD=analyst \
  $(PACKAGE_NAME) /init

# Then access: http://localhost:8888
```

## Can't Install Packages

```r
# In RStudio Console:
# Use standard install.packages() - auto-captured on container exit
install.packages("package-name")

# If package source unavailable, try different CRAN mirror:
options(repos = c(CRAN = "https://cloud.r-project.org/"))
install.packages("package-name")

# Check renv status if issues persist:
renv::status()

# Auto-restore should handle missing packages, but can manually restore if needed:
# renv::restore()
```

## Files Not Appearing

RStudio mounts `/project` directory:

- Your project root is at `/project`
- All files saved in RStudio appear in host
- Refresh Files pane if needed (click refresh icon)

## Session Crashes

```bash
# Restart RStudio container:
# 1. Stop current container (Ctrl+C)
# 2. Restart:
make docker-rstudio
```

All your file changes are preserved!

# Summary

You created a fully reproducible analysis using RStudio Server:

- **Familiar IDE:** RStudio interface in browser
- **No Local R:** Everything runs in Docker
- **Interactive Development:** Point-and-click + code
- **Reproducible:** Docker + renv ensure consistency
- **Auto-Snapshot:** Packages captured automatically on container exit
- **Auto-Restore:** Missing packages installed automatically on R startup
- **Testable:** Unit tests validate logic
- **Automated:** CI/CD runs on every commit
- **Collaborative:** Team members get identical environment

## RStudio Profile Contents

**What's included (~980MB):**

- R (latest or specified version)
- RStudio Server
- Essential packages: renv, here, usethis, devtools
- Development tooling baked into the image (not renv.lock): the R
  `languageserver` for in-container completion and diagnostics (default on;
  opt out with `zzc config set languageserver false`)
- System libraries: git, pandoc, LaTeX (basic). For full PDF/manuscript
  rendering use the `publishing` profile (rocker/verse), which pre-bakes the
  LaTeX closure
- Auto-snapshot on container exit
- Auto-restore on R startup

## Next Steps

**Extend your analysis:**
```r
# Add statistical models
fit_bill_model <- function(data) {
  lm(bill_depth_mm ~ bill_length_mm * species, data = data)
}

# Add more visualizations
create_flipper_plot <- function(data) {
  # Flipper length vs body mass
}

# Write more tests (tinytest, in inst/tinytest/)
model <- fit_bill_model(penguins)
expect_true('bill_length_mm:species' %in% names(coef(model)))
```

**Share with colleagues:**
```bash
# Team lead: Build and push
make docker-build
make docker-push-team            # Optional but efficient

# Team member: Clone and run
git clone https://github.com/yourteam/project.git && cd project
make docker-build
make docker-rstudio
```

For comprehensive documentation, see `vignette("quickstart")` for command-line workflow or explore other vignettes for advanced topics.
