---
title: "Quickstart: Collaborative Analysis Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quickstart: Collaborative Analysis Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This guide walks through a complete collaborative analysis workflow using
zzcollab. Two developers will collaborate on an analysis of the `mtcars`
dataset, demonstrating proper data management, testing, reporting, and
version control practices.

**Workflow Summary**:

1. Developer 1 sets up workspace, prepares data, writes initial analysis
2. Developer 1 pushes to GitHub

3. Developer 2 clones, adds visualization, pushes changes
4. Developer 1 pulls changes and finalizes the report

**Prerequisites**:

- Git installed and configured
- Docker installed and running
- GitHub account with SSH keys configured
- zzcollab repository cloned

## Part 1: Initial Setup (Developer 1)

### 1.1 Install zzcollab

After cloning the zzcollab repository, run the installer:

```bash
git clone https://github.com/rgt47/zzcollab.git
cd zzcollab
./install.sh
```

Verify installation (both commands work identically):

```bash
zzc --help
zzcollab --help
```

### 1.2 Configure User Defaults (One-Time Setup)

Set your personal defaults to avoid repeating common options:

```bash
zzc config init
zzc config set github-account "your-github-username"
zzc config set dockerhub-account "your-dockerhub-username"
```

### 1.3 Create Project Workspace

Create a new directory and initialize with the analysis profile:

```bash
mkdir mtcars-analysis
cd mtcars-analysis
zzc analysis                     # Full setup: init + renv + docker (prompts to build)
```

Optionally add GitHub:

```bash
zzc github                       # Create private GitHub repo (independent command)
```

Or create with public GitHub:

```bash
zzc github --public              # Create public GitHub repo
```

This creates the research compendium structure:

```
mtcars-analysis/
├── analysis/
│   ├── data/
│   │   ├── raw_data/
│   │   └── derived_data/
│   ├── figures/
│   ├── report/
│   │   ├── report.Rmd
│   │   └── references.bib
│   └── scripts/
├── R/
├── tests/
│   └── testthat/
├── DESCRIPTION
├── Dockerfile
├── Makefile
├── renv.lock
└── .Rprofile
```

### 1.4 Build the Docker Image

Build the Docker environment:

```bash
make docker-build
```

This creates a reproducible R environment based on the `analysis` profile
(rocker/tidyverse with tidyverse packages).

## Part 2: Data Preparation (Developer 1)

### 2.1 Enter the Container

Start an interactive R session inside the Docker container:

```bash
make r
```

### 2.2 Prepare and Save Raw Data

Inside R, save the complete mtcars dataset as CSV to the raw_data directory:

```r
# Save complete mtcars as CSV (raw data - read-only after this point)
write.csv(mtcars, "analysis/data/raw_data/mtcars.csv", row.names = TRUE)

# Verify the data
dim(mtcars)
# [1] 32 11

# Document the data source
writeLines(
  c(
    "# mtcars.csv",
    "",
    "## Source",
    "Motor Trend Car Road Tests from the 1974 Motor Trend US magazine.",
    "Complete dataset with 32 observations of 11 variables.",
    "",
    "## Variables",
    paste("-", names(mtcars), collapse = "\n"),
    "",
    "## Creation Date",
    as.character(Sys.Date()),
    "",
    "## Citation",
    "Henderson and Velleman (1981). Building multiple regression models",
    "interactively. Biometrics, 37, 391-411."
  ),
  "analysis/data/raw_data/README.md"
)
```

### 2.3 Create Data Processing Script

Create a script that reads the raw CSV and extracts the first half for analysis.

```bash
# Exit R first with q(), then create the script file on host
```

Create `analysis/scripts/01_prepare_data.R`:

```r
# 01_prepare_data.R
# Purpose: Extract first half of mtcars and add derived variables

library(dplyr)

prepare_mtcars_data <- function(input_path, output_path) {
  raw_data <- read.csv(input_path, row.names = 1)

  n_rows <- nrow(raw_data)
  subset_data <- raw_data[1:(n_rows %/% 2), ]

  processed_data <- subset_data |>
    mutate(
      efficiency_class = case_when(
        mpg >= 20 ~ "high",
        mpg >= 15 ~ "medium",
        TRUE ~ "low"
      ),
      weight_kg = wt * 453.592,
      power_to_weight = hp / wt
    )

  saveRDS(processed_data, output_path)
  processed_data
}

if (sys.nframe() == 0) {
  prepare_mtcars_data(
    input_path = "analysis/data/raw_data/mtcars.csv",
    output_path = "analysis/data/derived_data/mtcars_processed.rds"
  )
}
```

### 2.4 Extract Reusable Functions to R/

Move the core function to the package's R/ directory for testing and reuse.

Create `R/data_functions.R`:

```r
#' Prepare mtcars data for analysis
#'
#' @param data A data frame with mtcars structure
#' @return A data frame with additional computed columns
#' @export
add_derived_variables <- function(data) {
  if (!all(c("mpg", "wt", "hp") %in% names(data))) {
    stop("Input must contain mpg, wt, and hp columns")
  }

  data |>
    dplyr::mutate(
      efficiency_class = dplyr::case_when(
        mpg >= 20 ~ "high",
        mpg >= 15 ~ "medium",
        TRUE ~ "low"
      ),
      weight_kg = wt * 453.592,
      power_to_weight = hp / wt
    )
}

#' Validate mtcars data structure
#'
#' @param data A data frame to validate
#' @return TRUE if valid, throws error otherwise
#' @export
validate_mtcars_data <- function(data) {
  required_cols <- c("mpg", "cyl", "disp", "hp", "drat", "wt",
                     "qsec", "vs", "am", "gear", "carb")

  missing_cols <- setdiff(required_cols, names(data))
  if (length(missing_cols) > 0) {
    stop("Missing required columns: ", paste(missing_cols, collapse = ", "))
  }

  if (nrow(data) == 0) {
    stop("Data frame is empty")
  }

  if (any(data$mpg <= 0, na.rm = TRUE)) {
    stop("mpg must be positive")
  }

  if (any(data$hp <= 0, na.rm = TRUE)) {
    stop("hp must be positive")
  }

  TRUE
}
```

## Part 3: Writing Tests (Developer 1)

### 3.1 Create Data Validation Tests

Create `tests/testthat/test-data-validation.R`:

```r
# test-data-validation.R
# Tests for data integrity and validation functions

test_that("raw data file exists and is readable", {
  raw_data_path <- "analysis/data/raw_data/mtcars.csv"

  expect_true(file.exists(raw_data_path))

  data <- read.csv(raw_data_path, row.names = 1)
  expect_s3_class(data, "data.frame")
})

test_that("raw data has expected dimensions", {
  data <- read.csv("analysis/data/raw_data/mtcars.csv", row.names = 1)

  expect_equal(nrow(data), 32)
  expect_equal(ncol(data), 11)
})

test_that("raw data contains required columns", {
  data <- read.csv("analysis/data/raw_data/mtcars.csv", row.names = 1)

  required_cols <- c("mpg", "cyl", "disp", "hp", "drat", "wt",
                     "qsec", "vs", "am", "gear", "carb")

  expect_true(all(required_cols %in% names(data)))
})

test_that("raw data values are within expected ranges", {
  data <- read.csv("analysis/data/raw_data/mtcars.csv", row.names = 1)

  expect_true(all(data$mpg > 0 & data$mpg < 50))
  expect_true(all(data$cyl %in% c(4, 6, 8)))
  expect_true(all(data$hp > 0 & data$hp < 500))
  expect_true(all(data$wt > 0))
})

test_that("derived data has correct subset size", {
  skip_if_not(file.exists("analysis/data/derived_data/mtcars_processed.rds"))

  data <- readRDS("analysis/data/derived_data/mtcars_processed.rds")

  expect_equal(nrow(data), 16)
})
```

### 3.2 Create Function Tests

Create `tests/testthat/test-data-functions.R`:

```r
# test-data-functions.R
# Tests for R/data_functions.R

test_that("validate_mtcars_data accepts valid data", {
  valid_data <- data.frame(
    mpg = c(21, 22), cyl = c(6, 4), disp = c(160, 140),
    hp = c(110, 93), drat = c(3.9, 3.85), wt = c(2.6, 2.3),
    qsec = c(16.5, 18.6), vs = c(0, 1), am = c(1, 1),
    gear = c(4, 4), carb = c(4, 2)
  )

  expect_true(validate_mtcars_data(valid_data))
})

test_that("validate_mtcars_data rejects missing columns", {
  incomplete_data <- data.frame(mpg = 21, cyl = 6)

  expect_error(
    validate_mtcars_data(incomplete_data),
    "Missing required columns"
  )
})

test_that("validate_mtcars_data rejects empty data", {
  empty_data <- mtcars[0, ]

  expect_error(validate_mtcars_data(empty_data), "empty")
})

test_that("validate_mtcars_data rejects invalid mpg", {
  bad_data <- mtcars[1:2, ]
  bad_data$mpg[1] <- -5

  expect_error(validate_mtcars_data(bad_data), "mpg must be positive")
})

test_that("add_derived_variables creates expected columns", {
  test_data <- mtcars[1:5, ]
  result <- add_derived_variables(test_data)

  expect_true("efficiency_class" %in% names(result))
  expect_true("weight_kg" %in% names(result))
  expect_true("power_to_weight" %in% names(result))
})

test_that("add_derived_variables calculates efficiency_class correctly", {
  test_data <- data.frame(
    mpg = c(25, 17, 12),
    wt = c(2.5, 3.0, 3.5),
    hp = c(100, 150, 200)
  )

  result <- add_derived_variables(test_data)

  expect_equal(result$efficiency_class, c("high", "medium", "low"))
})

test_that("add_derived_variables calculates weight_kg correctly", {
  test_data <- data.frame(mpg = 20, wt = 1, hp = 100)
  result <- add_derived_variables(test_data)

  expect_equal(result$weight_kg, 453.592, tolerance = 0.001)
})

test_that("add_derived_variables requires correct input columns", {
  bad_data <- data.frame(x = 1, y = 2)

  expect_error(
    add_derived_variables(bad_data),
    "must contain mpg, wt, and hp"
  )
})
```

### 3.3 Run Tests

Enter the container and run tests:

```bash
make r
```

```r
# Inside R
devtools::test()
```

Expected output:

```
== Testing mtcarsanalysis =====================================================
[ FAIL 0 | WARN 0 | SKIP 0 | PASS 11 ]
```

Exit R when done:

```r
q()
```

Packages are automatically captured to renv.lock on exit.

## Part 4: Initial Analysis Report (Developer 1)

### 4.1 Create the Report

Edit `analysis/report/report.Rmd`:

```{r, eval=FALSE, echo=FALSE}
# The content below should be saved to analysis/report/report.Rmd
```

````
---
title: "Fuel Efficiency Analysis: mtcars Subset"
author: "Developer 1"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
bibliography: references.bib
---

`r ''````{r report-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

# Load package functions from R/ directory
devtools::load_all(here::here())
```

## Introduction

This analysis examines fuel efficiency patterns in a subset of the Motor
Trend Car Road Tests dataset. The raw data contains all 32 observations;
the derived dataset extracts the first 16 observations for analysis.

## Data Loading

```{r load-data}
library(dplyr)
library(ggplot2)
library(knitr)

# Load processed data
data <- readRDS(here::here("analysis/data/derived_data/mtcars_processed.rds"))

# Display summary
kable(head(data), caption = "First observations of processed data")
```

## Summary Statistics

```{r summary-stats}
summary_stats <- data |>
  group_by(efficiency_class) |>
  summarise(
    n = n(),
    mean_mpg = mean(mpg),
    mean_hp = mean(hp),
    mean_weight_kg = mean(weight_kg),
    .groups = "drop"
  )

kable(summary_stats,
      digits = 2,
      caption = "Summary statistics by efficiency class")
```

## Analysis Results

### Fuel Efficiency Distribution

```{r mpg-histogram}
ggplot(data, aes(x = mpg, fill = efficiency_class)) +
  geom_histogram(bins = 8, alpha = 0.7, color = "white") +
  labs(
    title = "Distribution of Fuel Efficiency",
    x = "Miles Per Gallon",
    y = "Count",
    fill = "Efficiency Class"
  ) +
  theme_minimal()
```

<!-- Developer 2 will add additional visualizations here -->

## Conclusions

*To be completed after collaborative additions.*

## References
````

### 4.2 Run the Analysis Pipeline

Enter container and run the processing script:

```bash
make r
```

```r
source("analysis/scripts/01_prepare_data.R")
```

Verify the derived data was created:

```r
file.exists("analysis/data/derived_data/mtcars_processed.rds")
# [1] TRUE

q()
```

### 4.3 Validate Package Dependencies

After exiting R, validate that all packages are properly tracked:

```bash
make check-renv
```

This pure-shell validation ensures DESCRIPTION and renv.lock are synchronized.

## Part 5: Push to GitHub (Developer 1)

### 5.1 Initialize Git and Create GitHub Repository

If you haven't created the GitHub repo yet, do it now:

```bash
zzc github                       # Initialize git + create private GitHub repo
```

Or for a public repo:

```bash
zzc github --public              # Create public GitHub repo
```

Or manually:

```bash
git init
git add .
git commit -m "Initial project setup with mtcars analysis

- Add raw data (complete mtcars CSV)
- Create data processing script (extracts first half)
- Add data validation and function tests
- Create initial report structure

🤖 Generated with [Claude Code](https://claude.com/claude-code)

Co-Authored-By: Claude <noreply@anthropic.com>"

gh repo create myteam/mtcars-analysis --private --source=. --push
```

### 5.2 Build and Push Team Docker Image

```bash
make docker-build
zzc dockerhub                    # Push to DockerHub
```

This makes the Docker image available for Developer 2.

## Part 6: Add Visualization (Developer 2)

### 6.1 Clone the Repository

Developer 2 clones the project and pulls the team Docker image:

```bash
git clone git@github.com:myteam/mtcars-analysis.git
cd mtcars-analysis
docker pull myteam/mtcars-analysis:latest
```

### 6.2 Enter Container and Verify Setup

```bash
make r
```

```r
# Verify data is present
readRDS("analysis/data/derived_data/mtcars_processed.rds") |> head()

# Run existing tests to ensure environment is correct
devtools::test()

q()
```

### 6.3 Add Visualization Function

Developer 2 adds a new plotting function. Create `R/plot_functions.R`:

```r
#' Create power-to-weight scatter plot
#'
#' @param data Processed mtcars data frame
#' @param show_labels Logical, whether to show car labels
#' @return A ggplot object
#' @export
plot_power_weight <- function(data, show_labels = TRUE) {
  if (!"power_to_weight" %in% names(data)) {
    stop("Data must contain power_to_weight column")
  }

  p <- ggplot2::ggplot(
    data,
    ggplot2::aes(x = weight_kg, y = hp, color = efficiency_class)
  ) +
    ggplot2::geom_point(size = 3, alpha = 0.8) +
    ggplot2::labs(
      title = "Horsepower vs Weight by Efficiency Class",
      x = "Weight (kg)",
      y = "Horsepower",
      color = "Efficiency"
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(legend.position = "bottom")

  if (show_labels) {
    p <- p + ggplot2::geom_text(
      ggplot2::aes(label = rownames(data)),
      hjust = -0.1,
      vjust = 0.5,
      size = 3
    )
  }

  p
}

#' Create cylinder comparison boxplot
#'
#' @param data Processed mtcars data frame
#' @return A ggplot object
#' @export
plot_mpg_by_cylinder <- function(data) {
  ggplot2::ggplot(data, ggplot2::aes(x = factor(cyl), y = mpg)) +
    ggplot2::geom_boxplot(fill = "steelblue", alpha = 0.7) +
    ggplot2::geom_jitter(width = 0.2, alpha = 0.5) +
    ggplot2::labs(
      title = "Fuel Efficiency by Number of Cylinders",
      x = "Number of Cylinders",
      y = "Miles Per Gallon"
    ) +
    ggplot2::theme_minimal()
}
```

### 6.4 Add Tests for New Functions

Create `tests/testthat/test-plot-functions.R`:

```r
# test-plot-functions.R
# Tests for R/plot_functions.R

test_that("plot_power_weight returns ggplot object", {
  test_data <- mtcars[1:5, ] |> add_derived_variables()

  result <- plot_power_weight(test_data)

  expect_s3_class(result, "ggplot")
})

test_that("plot_power_weight requires power_to_weight column", {
  bad_data <- mtcars[1:5, ]

  expect_error(
    plot_power_weight(bad_data),
    "power_to_weight"
  )
})

test_that("plot_power_weight respects show_labels parameter", {
  test_data <- mtcars[1:5, ] |> add_derived_variables()

  with_labels <- plot_power_weight(test_data, show_labels = TRUE)
  without_labels <- plot_power_weight(test_data, show_labels = FALSE)

  expect_s3_class(with_labels, "ggplot")
  expect_s3_class(without_labels, "ggplot")
})

test_that("plot_mpg_by_cylinder returns ggplot object", {
  test_data <- mtcars[1:5, ]

  result <- plot_mpg_by_cylinder(test_data)

  expect_s3_class(result, "ggplot")
})
```

### 6.5 Run All Tests

```bash
make r
```

```r
devtools::test()
```

Expected output:

```
== Testing mtcarsanalysis =====================================================
[ FAIL 0 | WARN 0 | SKIP 0 | PASS 15 ]
```

Exit R:

```r
q()
```

### 6.6 Commit and Push Changes

```bash
git add R/plot_functions.R tests/testthat/test-plot-functions.R
git commit -m "Add visualization functions for power-weight analysis

- Add plot_power_weight() for scatter plots
- Add plot_mpg_by_cylinder() for boxplots
- Include tests for both plotting functions

🤖 Generated with [Claude Code](https://claude.com/claude-code)

Co-Authored-By: Claude <noreply@anthropic.com>"

git push
```

## Part 7: Finalize Report (Developer 1)

### 7.1 Pull Changes

Developer 1 pulls Developer 2's additions:

```bash
git pull
```

### 7.2 Enter Container

```bash
make r
```

Verify new packages are auto-restored:

```r
devtools::test()
# All 15 tests should pass

q()
```

### 7.3 Update the Report

Edit `analysis/report/report.Rmd` to include the new visualizations.
Add the following sections after the existing histogram:

````markdown
### Power vs Weight Analysis

```{r power-weight-plot}
# Functions loaded via devtools::load_all() in setup chunk
plot_power_weight(data, show_labels = TRUE)
```

This scatter plot reveals the relationship between engine power and vehicle
weight. Higher efficiency vehicles (green) cluster in the lower-left region,
indicating lower weight and moderate horsepower.

### Cylinder Comparison

```{r cylinder-boxplot}
plot_mpg_by_cylinder(data)
```

The boxplot demonstrates the expected inverse relationship between cylinder
count and fuel efficiency. Four-cylinder vehicles show the highest MPG values,
while eight-cylinder vehicles demonstrate lower fuel efficiency.
````

Update the Conclusions section:

````markdown
## Conclusions

This analysis of the mtcars subset reveals several key findings:

1. **Efficiency Classification**: The dataset contains a mix of low, medium,
   and high efficiency vehicles based on the 15 and 20 MPG thresholds.

2. **Weight-Power Relationship**: There is a clear positive correlation
   between vehicle weight and horsepower, with efficiency class serving as
   a useful categorization.

3. **Cylinder Effect**: Vehicles with fewer cylinders consistently achieve
   better fuel economy, supporting the expected engineering trade-off between
   power and efficiency.

## Session Information

```{r session-info}
sessionInfo()
```
````

### 7.4 Render the Report

```bash
make r
```

```r
rmarkdown::render("analysis/report/report.Rmd")
q()
```

### 7.5 Final Validation

Run the complete test suite:

```bash
make docker-test
```

Validate package environment:

```bash
make check-renv
```

### 7.6 Commit Final Version

```bash
git add analysis/report/report.Rmd analysis/report/report.html
git commit -m "Finalize analysis report with collaborative visualizations

- Integrate power-weight scatter plot
- Add cylinder comparison boxplot
- Write conclusions section
- Add session info for reproducibility

🤖 Generated with [Claude Code](https://claude.com/claude-code)

Co-Authored-By: Claude <noreply@anthropic.com>"

git push
```

## Part 8: Team Collaboration Details

This section covers the mechanics of inviting collaborators and sharing Docker
images. These details were glossed over in the main workflow.

### 8.1 Inviting Collaborators to GitHub

**Adding a single collaborator:**

```bash
# Add collaborator with write access (can push commits)
gh repo edit --add-collaborator username

# Verify collaborators
gh api repos/:owner/:repo/collaborators --jq '.[].login'
```

**Adding multiple collaborators:**

```bash
# Add several collaborators
for user in alice bob charlie; do
  gh repo edit --add-collaborator "$user"
done
```

**Permission levels** (from least to most access):

| Level | Capabilities |
|-------|-------------|
| read | View code, clone, open issues |
| triage | + Manage issues and PRs |
| write | + Push commits, merge PRs |
| maintain | + Manage repo settings (not destructive) |
| admin | Full access including delete |

```bash
# Add with specific permission (default is 'write' for user repos)
gh api repos/:owner/:repo/collaborators/username -X PUT -f permission=maintain
```

**What happens after invitation:**

1. Collaborator receives email notification
2. They must accept the invitation (check email or github.com/notifications)
3. Once accepted, they can clone and push based on permission level

**Checking pending invitations:**

```bash
gh api repos/:owner/:repo/invitations --jq '.[].invitee.login'
```

### 8.2 DockerHub Image Sharing (Optional but Efficient)

**DockerHub is optional.** Collaborators can always build from the Dockerfile:

```bash
# Option A: Build from Dockerfile (always works, takes 5-15 minutes)
git clone https://github.com/myteam/study.git
cd study
make docker-build
make r
```

**Why use DockerHub?** Pre-installed packages save time:

When the team lead builds the image, all packages from `renv.lock` are
installed into the image. Team members who pull this image get those
packages immediately - no installation wait.

| Approach | What Happens | Time |
|----------|-------------|------|
| Build from Dockerfile | Downloads base + installs all packages | 5-15 min |
| Pull from DockerHub | Downloads pre-built image with packages | 1-3 min |

```bash
# Option B: Pull pre-built image (faster, requires DockerHub setup)
git clone https://github.com/myteam/study.git
cd study
docker pull myteam/study:latest
make r
```

### 8.3 Setting Up DockerHub (Team Lead)

**One-time DockerHub setup:**

1. Create account at hub.docker.com
2. Create repository (can be public or private)
3. Login from terminal:

```bash
docker login
# Enter DockerHub username and password/token
```

**Build before pushing (important):**

The Docker build process installs all R packages from `renv.lock` into the
image. This is what makes pulling faster for team members - packages are
pre-installed rather than installed on first run.

```bash
# 1. Build image (installs packages from renv.lock)
make docker-build

# 2. Push the built image to DockerHub
zzc dockerhub                    # Pushes to dockerhub-account/project:latest

# Or with version tag
zzc dockerhub --tag v1.0
```

**When to rebuild and push:**

- After adding packages (renv.lock changed)
- After modifying Dockerfile
- Before onboarding new team members (ensure they get latest)

```bash
# Check if renv.lock changed since last push
git log --oneline -3 -- renv.lock

# If changed, rebuild and push
make docker-build && zzc dockerhub
```

### 8.4 Private DockerHub Repositories

For private research, use private DockerHub repos:

**Team lead creates private repo:**

1. Go to hub.docker.com → Create Repository
2. Set visibility to "Private"
3. Add collaborators under "Collaborators" tab

**Team members authenticate:**

```bash
# Each team member must login once
docker login
# Enter their DockerHub credentials

# Then pull works
docker pull myteam/private-study:latest
```

**Free tier limitations:**

- Docker Free: 1 private repo, unlimited public
- Docker Pro: Unlimited private repos

For academic teams, public repos are often sufficient since the code is
on private GitHub anyway.

### 8.5 When to Rebuild vs Pull

**Pull the team image when:**

- First time joining the project
- Team lead announces image update
- Dockerfile hasn't changed since last pull

**Rebuild from Dockerfile when:**

- Dockerfile has been modified
- You need to verify reproducibility
- CI/CD (always builds fresh)
- Troubleshooting environment issues

**Check if rebuild needed:**

```bash
# See if Dockerfile changed since last pull
git log --oneline -5 -- Dockerfile

# If changes exist, rebuild
make docker-build
```

### 8.6 Complete Team Workflow Example

**Team Lead (one-time setup):**

```bash
# Create project
mkdir team-analysis && cd team-analysis
zzc analysis
zzc github

# Push Docker image
docker login                     # One-time authentication
zzc dockerhub

# Add collaborators
gh repo edit --add-collaborator alice
gh repo edit --add-collaborator bob
```

**Team Member (joining):**

```bash
# Clone repository
git clone https://github.com/leader/team-analysis.git
cd team-analysis

# Option A: Fast path (pull image)
docker login                     # If private DockerHub repo
docker pull leader/team-analysis:latest
make r

# Option B: Full reproducibility (build from scratch)
make docker-build                # Takes longer but verifies Dockerfile
make r
```

**Daily workflow (all team members):**

```bash
git pull                         # Get latest code
make r                           # Enter container (auto-restores packages)
# ... work ...
exit                             # Auto-snapshots packages
git add . && git commit && git push
```

## Summary

This workflow demonstrated:

1. **Project Setup**: Creating a zzcollab workspace with proper structure
2. **Data Management**: Storing raw data in `analysis/data/raw_data/` with
   documentation
3. **Code Organization**: Separating reusable functions into `R/` directory
4. **Testing**: Writing tests for both data validation and functions
5. **Collaboration**: Two developers working on the same codebase
6. **Docker**: Using containerized environments for reproducibility
7. **Version Control**: Proper commit messages and GitHub workflow
8. **Team Onboarding**: Inviting collaborators and sharing Docker images

The final repository contains:

- Reproducible Docker environment
- Tested and documented R functions
- Complete analysis report with visualizations
- Full version history of collaborative work

## Quick Reference

| Task | Command |
|------|---------|
| Create project (full setup) | `zzc analysis` |
| Add GitHub repo | `zzc github` |
| Enter container | `make r` |
| Run tests | `devtools::test()` inside R |
| Validate packages | `make check-renv` |
| Build Docker image | `make docker-build` |
| Push to DockerHub | `zzc dockerhub` |
| Push with tag | `zzc dockerhub --tag v1.0` |
| Switch profile | `zzc publishing` |
| List profiles | `zzc list profiles` |
| Render report | `make docker-render` |

## Next Steps

- Add CI/CD with GitHub Actions for automated testing
- Implement additional analyses on the second half of mtcars
- Create a package vignette documenting the analysis methodology
- Add code coverage reporting with covr
