---
title: "Quick Start: RStudio Server Workflow"
author: "ZZCOLLAB Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Quick Start: RStudio Server Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

This quick start guide demonstrates how to use **RStudio Server** inside a Docker container for reproducible analysis. You'll work in a familiar RStudio interface while maintaining complete reproducibility through Docker and renv.

**What you will build:**
- RStudio Server environment accessible via browser
- Scatter plot: bill length vs bill depth (Palmer Penguins)
- Complete reproducibility: renv + Docker + Unit Testing + CI/CD
- Private GitHub repository with automated validation

**Why RStudio Profile:**
- Familiar RStudio IDE interface in browser
- No local R installation needed
- Point-and-click package installation
- Visual debugging and data exploration
- Works on any machine with Docker and a browser

**Prerequisites:**
- Docker installed and running
- Web browser
- GitHub account (example: rgt47)
- zzcollab installed (`./install.sh`)
- Dotfiles directory at `~/dotfiles` (optional)

# Step-by-Step Instructions

## Step 1: Configure ZZCOLLAB for RStudio (30 seconds)

Set your defaults once, specifying the RStudio profile:

```bash
# Initialize configuration
zzcollab --config init

# Set team name (typically your GitHub username for solo projects)
zzcollab --config set team-name "rgt47"

# Set GitHub account
zzcollab --config set github-account "rgt47"

# Set RStudio profile as default
zzcollab --config set profile-name "rstudio"

# Set dotfiles directory (optional)
zzcollab --config set dotfiles-dir "~/dotfiles"
```

**What this accomplishes:**
- RStudio Server will be available in all future projects
- Includes R, RStudio Server, essential packages
- ~980MB Docker image (more than minimal, less than full analysis)

## Step 2: Create Project with RStudio Profile (3-4 minutes)

Create your reproducible analysis project:

```bash
# Create project directory and initialize
mkdir penguin-bills && cd penguin-bills
zzcollab  # Uses rstudio profile from config

# Build Docker image with RStudio Server
make docker-build
```

**What this does:**
1. Creates complete project structure (R/, analysis/, tests/, .github/, etc.)
2. Sets up CI/CD workflows for automated testing
3. Initializes renv for dependency tracking
4. Builds Docker image with RStudio Server included

## Step 3: Start RStudio Server (10 seconds)

Launch RStudio Server in your browser:

```bash
make docker-rstudio
```

**Expected output:**
```
Starting RStudio Server on http://localhost:8787
Username: analyst, Password: analyst
```

**Access RStudio:**
1. Open browser to **http://localhost:8787**
2. Login with:
   - Username: `analyst`
   - Password: `analyst`
3. You're now in RStudio running inside Docker!

## Step 4: Create Analysis in RStudio (5 minutes)

Work directly in the RStudio interface:

### 4.1 Create Analysis Function

In RStudio, create file `R/bill_analysis.R`:

**File → New File → R Script**

```r
#' Create scatter plot of bill dimensions
#'
#' @param data Palmer penguins data
#' @return ggplot object
#' @export
create_bill_plot <- function(data = palmerpenguins::penguins) {
  data %>%
    dplyr::filter(!is.na(bill_length_mm), !is.na(bill_depth_mm)) %>%
    ggplot2::ggplot(ggplot2::aes(x = bill_length_mm, y = bill_depth_mm,
                                  color = species)) +
    ggplot2::geom_point(size = 3, alpha = 0.7) +
    ggplot2::labs(
      title = 'Palmer Penguins: Bill Dimensions',
      x = 'Bill Length (mm)',
      y = 'Bill Depth (mm)',
      color = 'Species'
    ) +
    ggplot2::theme_minimal()
}
```

Save as: `R/bill_analysis.R`

### 4.2 Create Analysis Report

In RStudio, create file `analysis/paper/paper.Rmd`:

**File → New File → R Markdown...**

````
---
title: "Palmer Penguins Bill Dimensions Analysis"
author: "Reproducible Research Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

Analysis of bill length and bill depth measurements across three penguin species
from the Palmer Station Antarctica LTER.

## Data Loading

```{r load-data}
library(palmerpenguins)
library(ggplot2)
library(dplyr)

# Source analysis functions from R/
source("../../R/bill_analysis.R")

# Preview data
head(penguins)
```

## Analysis

### Bill Dimensions by Species

```{r bill-plot, fig.width=8, fig.height=6}
# Generate plot using function from R/
plot <- create_bill_plot()
print(plot)

# Save plot
ggsave("../figures/bill_scatter.png", plot, width = 8, height = 6)
```

## Summary

The analysis reveals distinct bill dimension patterns across penguin species.
```
````

Save as: `analysis/paper/paper.Rmd`

### 4.3 Install Packages in RStudio

Use RStudio's **Packages** pane or the Console:

```r
# In RStudio Console:
renv::install("palmerpenguins")
renv::install("ggplot2")
renv::install("dplyr")

# Save dependencies
renv::snapshot()
```

**RStudio Advantage:** You can also use the Packages pane:
1. Click "Packages" tab (bottom-right)
2. Click "Install"
3. Type package name
4. Click "Install"
5. Then run `renv::snapshot()` to save

### 4.4 Render Report in RStudio

Render the analysis report:

**Method 1 - Click "Knit" button** in `paper.Rmd`

**Method 2 - Console command:**
```r
rmarkdown::render("analysis/paper/paper.Rmd")
```

Your HTML report appears in `analysis/paper/paper.html`!

## Step 5: Add Unit Tests in RStudio (2 minutes)

Create tests in RStudio:

### 5.1 Initialize Testing

In RStudio Console:
```r
usethis::use_testthat()
```

### 5.2 Create Test File

**File → New File → R Script**

```r
# Source function from R/
source('../../R/bill_analysis.R')

test_that('create_bill_plot produces valid ggplot', {
  plot <- create_bill_plot()

  # Valid ggplot object
  expect_s3_class(plot, 'ggplot')

  # Correct title
  expect_equal(plot$labels$title, 'Palmer Penguins: Bill Dimensions')
})

test_that('create_bill_plot handles missing data', {
  # Data with missing values
  test_data <- palmerpenguins::penguins
  test_data$bill_length_mm[1:5] <- NA

  plot <- create_bill_plot(test_data)

  # Should still produce plot
  expect_s3_class(plot, 'ggplot')
})

test_that('create_bill_plot includes all species', {
  plot <- create_bill_plot()

  # Should have species in color aesthetic
  expect_true('species' %in% names(plot$labels))
})
```

Save as: `tests/testthat/test-bill_analysis.R`

### 5.3 Run Tests in RStudio

In RStudio Console:
```r
devtools::test()
```

Or use **Build → Test Package** (if package structure)

## Step 6: Exit RStudio and Validate (1 minute)

Close your browser tab, then in terminal:

```bash
# Stop RStudio container (Ctrl+C in terminal where docker-rstudio is running)

# Run tests in clean environment
make docker-test

# Validate dependencies (pure shell - NO R REQUIRED!)
make check-renv  # Pure shell validation (grep, awk, jq)
```

## Step 7: Commit and Push (1 minute)

```bash
git add .
git commit -m "$(cat <<'EOF'
Add bill dimensions scatter plot analysis

Analysis:
- Scatter plot of bill length vs bill depth by species
- Palmer Penguins dataset
- Function-based implementation with unit tests
- Developed in RStudio Server

Reproducibility:
- Level 2: renv dependency tracking
- Level 3: Docker environment isolation (rstudio profile)
- Level 4: Unit tests for computational correctness
- Level 5: CI/CD automated validation

🤖 Generated with [Claude Code](https://claude.com/claude-code)

Co-Authored-By: Claude <noreply@anthropic.com>
EOF
)"

git push
```

## Step 8: Verify CI/CD (30 seconds)

```bash
# View CI/CD status
gh run list --limit 3

# Or view in browser
gh repo view --web
```

# RStudio Workflow Advantages

## Point-and-Click Development

**Visual Interface Benefits:**
- File browser for easy navigation
- Environment pane shows all objects
- Plots pane displays visualizations
- Help pane with integrated documentation
- Package manager with search

## Interactive Analysis

**Real-time Exploration:**
```r
# In RStudio Console - explore interactively:
library(palmerpenguins)
View(penguins)  # Opens data viewer
summary(penguins)
table(penguins$species)

# Test plot interactively
library(ggplot2)
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point()
```

## Debugging Support

**Visual Debugging:**
1. Set breakpoints (click left of line number)
2. Click "Source" to run with debugging
3. Inspect variables in Environment pane
4. Step through code line by line

## RMarkdown Integration

**Document Authoring:**
- Visual markdown editor (RStudio 1.4+)
- Live preview as you type
- Insert code chunks with shortcuts
- Knit to HTML/PDF/Word with one click

# Accessing RStudio Server

## Starting RStudio

```bash
# From project directory:
make docker-rstudio
```

Opens on: **http://localhost:8787**

## Default Credentials

- **Username:** `analyst`
- **Password:** `analyst`

## Customizing Credentials

Edit your project's Makefile to change:

```makefile
docker-rstudio:
	@echo "Starting RStudio Server on http://localhost:8787"
	@echo "Username: myuser, Password: mypass"
	docker run --platform linux/amd64 --rm -p 8787:8787 \
	  -v $$(pwd):/project \
	  -e USER=myuser \
	  -e PASSWORD=mypass \
	  $(PACKAGE_NAME) /init
```

## Persistent Sessions

RStudio session is temporary - work is saved but R workspace isn't:
- **Saved:** All file edits, created files, installed packages (via renv)
- **Not saved:** R workspace (.RData), unless you explicitly save

**Best Practice:** Save work via files, not R workspace:
```r
# Don't rely on:
save.image()  # Workspace not portable

# Instead use:
saveRDS(my_data, "data/my_data.rds")  # Portable, reproducible
renv::snapshot()  # Save package versions
```

# Transitioning to Command Line

## When to Use Terminal

RStudio is great for development, but some tasks need terminal:

**Use Terminal for:**
- Running tests: `make docker-test`
- Validation: `make check-renv` (pure shell - NO R required!)
- Git operations: `git commit`, `git push`
- Building images: `make docker-build`

**Use RStudio for:**
- Writing analysis code
- Creating visualizations
- Authoring reports
- Exploratory analysis
- Package development

## Hybrid Workflow

**Typical workflow:**
1. **RStudio:** Write and test code interactively
2. **RStudio:** Author report and render locally
3. **Terminal:** Run full test suite
4. **Terminal:** Validate dependencies
5. **Terminal:** Commit and push
6. **CI/CD:** Automated validation runs

# Team Collaboration with RStudio

## Team Lead Setup

```bash
# Set RStudio as team profile
zzcollab --config set team-name "genomicslab"
zzcollab --config set profile-name "rstudio"

# Create team project
mkdir study && cd study
zzcollab -t genomicslab -p study --profile-name rstudio

# Build and share team image
make docker-build
make docker-push-team

# Commit project structure
git add .
git commit -m "Initial project setup with RStudio Server"
git push -u origin main
```

## Team Member Workflow

```bash
# Clone repository
git clone https://github.com/genomicslab/study.git
cd study

# Pull and use team RStudio image
zzcollab --use-team-image

# Start RStudio Server
make docker-rstudio

# Access http://localhost:8787
# Work in identical RStudio environment as team
```

**Team Benefits:**
- Everyone gets identical RStudio version
- Same R version across team
- Consistent package versions (via renv)
- No "works on my machine" issues

# Troubleshooting

## RStudio Won't Start

```bash
# Check Docker is running
docker info

# Check port 8787 is free
lsof -i :8787

# If port busy, use different port:
docker run --rm -p 8888:8787 -v $(pwd):/project \
  -e USER=analyst -e PASSWORD=analyst \
  $(PACKAGE_NAME) /init

# Then access: http://localhost:8888
```

## Can't Install Packages

```r
# In RStudio Console:
# Check renv status
renv::status()

# Restore from lockfile
renv::restore()

# If package source unavailable, try:
options(repos = c(CRAN = "https://cloud.r-project.org/"))
install.packages("package-name")
renv::snapshot()
```

## Files Not Appearing

RStudio mounts `/project` directory:
- Your project root is at `/project`
- All files saved in RStudio appear in host
- Refresh Files pane if needed (click refresh icon)

## Session Crashes

```bash
# Restart RStudio container:
# 1. Stop current container (Ctrl+C)
# 2. Restart:
make docker-rstudio
```

All your file changes are preserved!

# Summary

You created a fully reproducible analysis using RStudio Server:

- ✅ **Familiar IDE:** RStudio interface in browser
- ✅ **No Local R:** Everything runs in Docker
- ✅ **Interactive Development:** Point-and-click + code
- ✅ **Reproducible:** Docker + renv ensure consistency
- ✅ **Testable:** Unit tests validate logic
- ✅ **Automated:** CI/CD runs on every commit
- ✅ **Collaborative:** Team members get identical environment

## RStudio Profile Contents

**What's included (~980MB):**
- R (latest or specified version)
- RStudio Server
- Essential packages: renv, here, usethis, devtools
- System libraries: git, pandoc, LaTeX (basic)

## Next Steps

**Extend your analysis:**
```r
# Add statistical models
fit_bill_model <- function(data) {
  lm(bill_depth_mm ~ bill_length_mm * species, data = data)
}

# Add more visualizations
create_flipper_plot <- function(data) {
  # Flipper length vs body mass
}

# Write more tests
test_that("model includes species interaction", {
  model <- fit_bill_model(penguins)
  expect_true("bill_length_mm:species" %in% names(coef(model)))
})
```

**Share with colleagues:**
```bash
# Team lead: Build and push
make docker-build
make docker-push-team

# Team member: Pull and use
zzcollab --use-team-image
make docker-rstudio
```

For comprehensive documentation, see `vignette("quickstart")` for command-line workflow or explore other vignettes for advanced topics.
