---
title: "Quick Start: Complete Reproducible Data Workflow"
author: "ZZCOLLAB Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Quick Start: Complete Reproducible Data Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

This quick start guide demonstrates how to create a fully reproducible analysis
with all 5 pillars of reproducibility. You will create a complete data workflow
from raw data through processing, analysis, testing, and reporting using Palmer
Penguins bill measurements.

**What you will build:**

- Complete data workflow: raw data → processing → analysis → visualization
- All 5 pillars: Dockerfile + renv + .Rprofile + Source Code + Research Data
- Functions, tests, and documentation following best practices
- Private GitHub repository with automated CI/CD validation

**Prerequisites:**

- Docker installed and running
- GitHub account (example: rgt47)
- Personal access token for GitHub (with repo permissions)
- zzcollab installed (`./install.sh`)
- `gum` installed for styled TUI prompts (optional but recommended):
  `brew install gum`

> **Key terms in this vignette.** Each is defined again on first use.
> Full definitions are in `vignette('glossary')`.
>
> - **Compendium**: your project folder, holding data, code, and
>   write-up together so the whole thing can be re-run as a unit.
> - **Archetype**: the project type you choose at creation (analysis,
>   manuscript, package, simulation, or blog), which shapes the starter
>   layout.
> - **Profile**: the Docker image bundle (which R packages are baked
>   into the container): minimal, analysis, or rstudio. Distinct
>   from the archetype.
> - **Feature**: an independently switchable capability (a package
>   backend, Docker, CI, data hashing, code-quality hooks, unit tests,
>   or cloud launch) toggled with `zzc toggle` or `zzc add` / `zzc rm`.
> - **Capture level (L0 to L3)**: how locked-down the project is, from
>   'just findable' (L0) up to 'rebuilt and verified' (L3).
> - **renv / lockfile**: the package backend that records exact package
>   versions in `renv.lock` so installs are repeatable.
> - **Auto-snapshot**: the automatic `renv::snapshot()` that runs when
>   you exit R in the container, capturing new packages into `renv.lock`.
> - **Five Pillars**: Dockerfile, renv.lock, .Rprofile, source code, and
>   research data, the necessary and sufficient parts of a reproducible
>   compendium.

# Step-by-Step Instructions

## Step 1: One-time Identity Setup

The first time you run `zzc init`, it detects missing required fields and
guides you through a short identity prompt. This runs once and is never
asked again once your user config is populated.

The identity gate captures three fields and saves them to
`~/.zzcollab/config.yaml`:

- **Full name** -- written into `DESCRIPTION`'s `Authors@R` field
- **Email address** -- written into `DESCRIPTION`
- **Personal GitHub username** -- your default account; can be overridden
  per project in the next step

With `gum` installed the prompts render as a styled TUI with pre-filled
values and cursor navigation. Without `gum` they fall back to plain readline.

To inspect or change these defaults at any time:

```bash
zzc config list                          # show all current settings
zzc config set author-name "Your Name"   # update a specific field
zzc config set github-account "youruser"
```

**Note:** Both `zzc` and `zzcollab` work identically.

## Step 2: Create Project

Create the project directory and run `zzc init`:

```bash
mkdir penguin-bills && cd penguin-bills
zzc init
```

`zzc init` runs an identity gate on first use, then asks two questions.

**Identity gate** (first run only)

Fires when `~/.zzcollab/config.yaml` is missing name or email. Captures
the three identity fields described in Step 1 and saves them to your user
config. Skipped on subsequent projects.

**Question 1 -- Archetype**

The archetype is the project type, which shapes the starter layout:

```
Archetype [analysis/manuscript/package/simulation/blog] (default analysis):
```

- `analysis`, `manuscript`, and `blog` scaffold a report
  (`analysis/report/report.Rmd`); `blog` also gets an `analysis/posts/`
  directory.
- `package` and `simulation` do not scaffold a report; `simulation` gets
  a seeded, parallel-ready starter script instead.

For this tutorial accept the default, `analysis`.

**Question 2 -- Feature wizard**

The second question is the feature wizard (the same one `zzc toggle`
uses). It recommends a sensible default, renv plus Docker, which you can
accept or untick. Each toggle is an independently switchable *feature*:
the package backend (renv), Docker, CI, data-integrity hashing,
code-quality hooks, unit tests, and cloud launch. You do not have to
decide everything now; features can be added or removed at any time (see
`vignette("quickstart3")` for the full toggle model). For this tutorial,
accept the recommended renv plus Docker.

Only fields that differ from your user defaults are written to the
project's `./zzcollab.yaml`, keeping the file minimal. Commit this file;
it documents project-specific decisions and allows collaborators to
reproduce the same configuration.

Example `./zzcollab.yaml` when only the GitHub account differs:

```yaml
# Project-specific zzcollab configuration
# Overrides user config (~/.zzcollab/config.yaml)
# Commit this file to version control.
github:
  account: genomicslab
```

**Non-interactive forms.** `--yes` skips every prompt and accepts the
defaults; `--archetype <x>` sets the archetype as a flag:

```bash
# Scaffold an analysis project with default features, no prompts.
zzc init --yes

# Set the archetype explicitly, still ask the feature wizard.
zzc init --archetype manuscript
```

After init, create the GitHub repository:

```bash
zzc github          # creates private repo and pushes
```

**Quickstart alternative** (uses config defaults, no interactive prompts):

```bash
mkdir penguin-bills && cd penguin-bills
zzc analysis        # init + renv + docker in one command
zzc github
```

### Profiles versus archetypes

These are two distinct choices, and it is worth keeping them apart:

- An **archetype** is the project type and layout (chosen in Question 1).
- A **profile** is the Docker image bundle, that is, which R packages are
  baked into the container.

A profile command is also smart about context:

- **New project**: full setup (init + renv + docker + build prompt).
- **Existing project**: switches the profile and regenerates the
  Dockerfile only.

```bash
# Create a new project with the analysis profile.
mkdir my-project && cd my-project
zzc analysis                   # Full setup (init + renv + docker)

# Switch profile on an existing project.
zzc rstudio                    # Regenerates Dockerfile with rstudio profile

# List available profiles.
zzc list profiles
```

**Available profiles** (the three canonical profiles):

- `minimal` - Command-line R environment (~650MB)
- `analysis` - Tidyverse data analysis (~1.2GB), recommended
- `rstudio` - RStudio Server (~980MB)

For specialised needs (LaTeX, modeling, Shiny), start with the closest one and add
packages inside the container with `install.packages()`, then commit the
updated `renv.lock`.

## Step 3: Add Research Data and Processing Pipeline

This step demonstrates the complete data workflow following the Five Pillars of Reproducibility:

- **Pillar 5**: Raw data in `analysis/data/raw_data/`
- **Pillar 4**: Processing scripts in `analysis/scripts/`
- **Pillar 4**: Reusable functions in `R/`
- **Pillar 4**: Unit tests in `inst/tinytest/`

Start the Docker development environment:

```bash
# Enter Docker container
make r
```

### Add Raw Data

Create a raw data file (simulating field-collected penguin measurements).

Create `analysis/data/raw_data/penguins_raw.csv`:

```csv
species,island,bill_length_mm,bill_depth_mm,flipper_length_mm,body_mass_g,sex,year
Adelie,Torgersen,39.1,18.7,181,3750,male,2007
Adelie,Torgersen,39.5,17.4,186,3800,female,2007
Adelie,Torgersen,40.3,18.0,195,3250,female,2007
Adelie,Torgersen,NA,NA,NA,NA,NA,2007
Gentoo,Biscoe,46.1,13.2,211,4500,female,2007
Gentoo,Biscoe,50.0,16.3,230,5700,male,2007
Chinstrap,Dream,46.5,17.9,192,3500,female,2007
Chinstrap,Dream,50.0,19.5,196,3900,male,2007
```

Create `analysis/data/raw_data/README.md`:

```markdown
# Raw Data: Palmer Penguins

## Source
Simulated penguin measurements from Palmer Station Antarctica LTER.

## Collection Methods
- Field measurements of adult penguins
- Three species: Adelie, Gentoo, Chinstrap
- Three islands: Torgersen, Biscoe, Dream

## Variables
- species: Penguin species (Adelie, Gentoo, Chinstrap)
- island: Island where observed
- bill_length_mm: Bill length in millimeters
- bill_depth_mm: Bill depth in millimeters
- flipper_length_mm: Flipper length in millimeters
- body_mass_g: Body mass in grams
- sex: Penguin sex (male, female)
- year: Year of observation

## Known Issues
- Row 4 contains missing values (equipment failure)
- All measurements subject to ±0.5mm measurement error
```

### Create Data Processing Script

Create a script to clean and process the raw data.

Create `analysis/scripts/01_process_data.R`:

```r
# Data Processing Script: Clean raw penguin data
# Input: analysis/data/raw_data/penguins_raw.csv
# Output: analysis/data/derived_data/penguins_clean.csv

library(dplyr)

# Load raw data
raw_data <- read.csv("analysis/data/raw_data/penguins_raw.csv")

# Process data
clean_data <- raw_data %>%
  # Remove rows with missing values
  filter(!is.na(bill_length_mm), !is.na(bill_depth_mm)) %>%
  # Add derived variables
  mutate(
    bill_ratio = bill_length_mm / bill_depth_mm,
    size_category = case_when(
      body_mass_g < 3500 ~ "small",
      body_mass_g < 4500 ~ "medium",
      TRUE ~ "large"
    )
  )

# Create output directory if needed
dir.create("analysis/data/derived_data", showWarnings = FALSE, recursive = TRUE)

# Save processed data
write.csv(clean_data, "analysis/data/derived_data/penguins_clean.csv", row.names = FALSE)

# Log processing results
cat("Processed", nrow(clean_data), "valid records from", nrow(raw_data), "raw records\n")
cat("Removed", nrow(raw_data) - nrow(clean_data), "incomplete records\n")
```

Create `analysis/data/derived_data/README.md`:

```markdown
# Processed Data: Palmer Penguins (Clean)

## Source
Processed from `analysis/data/raw_data/penguins_raw.csv`

## Processing Steps
See: `analysis/scripts/01_process_data.R`

1. Loaded raw data
2. Removed rows with missing bill measurements
3. Added derived variables:
   - bill_ratio: bill_length_mm / bill_depth_mm
   - size_category: small (<3500g), medium (3500-4500g), large (>4500g)

## Quality Checks
- All rows have complete bill measurements
- All derived variables computed successfully
- No outliers detected beyond expected range
```

Run the processing script:

```bash
Rscript analysis/scripts/01_process_data.R
```

### Create Reusable Analysis Functions

Create functions in `R/` directory that use the processed data.

Create `R/data_utils.R`:

```r
#' Load processed penguin data
#'
#' @param path Path to clean data file
#' @return data.frame of processed penguin data
#' @export
load_penguin_data <- function(path = "analysis/data/derived_data/penguins_clean.csv") {
  if (!file.exists(path)) {
    stop("Clean data file not found. Run analysis/scripts/01_process_data.R first.")
  }

  data <- read.csv(path, stringsAsFactors = FALSE)

  # Validate expected columns
  required_cols <- c("species", "bill_length_mm", "bill_depth_mm", "bill_ratio")
  missing_cols <- setdiff(required_cols, names(data))

  if (length(missing_cols) > 0) {
    stop("Missing required columns: ", paste(missing_cols, collapse = ", "))
  }

  data
}

#' Calculate summary statistics by species
#'
#' @param data Penguin data
#' @return data.frame of summary statistics
#' @export
summarize_by_species <- function(data) {
  data %>%
    dplyr::group_by(species) %>%
    dplyr::summarize(
      n = dplyr::n(),
      mean_bill_length = mean(bill_length_mm, na.rm = TRUE),
      mean_bill_depth = mean(bill_depth_mm, na.rm = TRUE),
      mean_bill_ratio = mean(bill_ratio, na.rm = TRUE),
      .groups = "drop"
    )
}
```

Install required packages:

```bash
# Packages will be captured in renv.lock when you exit the container
Rscript -e 'install.packages(c("dplyr", "here"))'
```

### Create Unit Tests

Add tests for the data processing functions. ZZCOLLAB projects use tinytest:
test files live in `inst/tinytest/` and contain bare top-level expectations
(no `test_that()` wrapper).

Create `inst/tinytest/test-data_utils.R`:

```r
# Tests for data utility functions
# make docker-test loads functions from R/ via devtools::load_all()

clean_csv <- 'analysis/data/derived_data/penguins_clean.csv'

# load_penguin_data loads valid data (only when processing has been run)
if (file.exists(clean_csv)) {
  data <- load_penguin_data()
  expect_true(is.data.frame(data))
  expect_true(nrow(data) > 0)
  expect_true('species' %in% names(data))
  expect_true('bill_length_mm' %in% names(data))
  expect_true('bill_depth_mm' %in% names(data))
  expect_true('bill_ratio' %in% names(data))

  # summarize_by_species produces valid output
  summary <- summarize_by_species(data)
  expect_true(is.data.frame(summary))
  expect_true('species' %in% names(summary))
  expect_true('n' %in% names(summary))
  expect_true('mean_bill_length' %in% names(summary))
  expect_true(is.double(summary$mean_bill_length))
  expect_true(is.double(summary$mean_bill_depth))
  expect_true(all(summary$mean_bill_length > 0))
  expect_true(all(summary$mean_bill_depth > 0))

  # bill_ratio is calculated correctly
  expect_equal(data$bill_ratio,
               data$bill_length_mm / data$bill_depth_mm)
}

# load_penguin_data fails gracefully with a missing file
expect_error(load_penguin_data('/nonexistent/path.csv'),
             pattern = 'Clean data file not found')
```

Run tests:

```bash
make docker-test
```

**What this accomplishes:**

1. Raw Data (Pillar 5): Original measurements in `analysis/data/raw_data/`
2. Processing Pipeline: Script in `analysis/scripts/01_process_data.R`
3. Derived Data: Cleaned data in `analysis/data/derived_data/`
4. Reusable Functions: Data utilities in `R/data_utils.R`
5. Unit Tests: Tests in `inst/tinytest/test-data_utils.R`
6. Documentation: README files documenting data provenance and processing

This establishes the complete data workflow that ensures reproducibility from raw data through to analysis.

## Step 4: Create Analysis Report

Now create a report that uses the processed data and functions we created in Step 3.

### Create Visualization Function

Add a plotting function.

Create `R/bill_analysis.R`:

```r
#' Create scatter plot of bill dimensions
#'
#' @param data Penguin data with bill measurements
#' @return ggplot object
#' @export
create_bill_plot <- function(data) {
  ggplot2::ggplot(data, ggplot2::aes(x = bill_length_mm, y = bill_depth_mm,
                                      color = species)) +
    ggplot2::geom_point(size = 3, alpha = 0.7) +
    ggplot2::labs(
      title = 'Palmer Penguins: Bill Dimensions',
      x = 'Bill Length (mm)',
      y = 'Bill Depth (mm)',
      color = 'Species'
    ) +
    ggplot2::theme_minimal()
}
```

Install ggplot2:

```bash
Rscript -e 'install.packages("ggplot2")'
```

### About GUI Support (Automatic)

The `make r` command auto-detects your profile and configures GUI support
automatically:

- **X11 profiles** (e.g., `x11_minimal`): Automatically enables DISPLAY for
  interactive graphics
- **RStudio profiles** (e.g., `analysis`, `rstudio`): Launches RStudio Server at http://localhost:8787
- **Shiny profiles**: Launches Shiny Server at http://localhost:3838
- **Standard profiles**: Opens interactive shell

**For X11 GUI support** (if using an X11 profile):

One-time XQuartz setup (macOS):
```bash
brew install --cask xquartz
open -a XQuartz
# In Preferences > Security: Enable "Allow connections from network clients"
# Restart XQuartz, then:
export DISPLAY=:0
/opt/X11/bin/xhost +localhost
```

For this tutorial, we use the default profile and save plots to files (works in all environments including CI/CD).

### Create Analysis Report

Create the report that uses our data workflow (raw data → processing → functions → visualization).

Create `analysis/report/report.Rmd`:

````
---
title: "Palmer Penguins Bill Dimensions Analysis"
author: "Reproducible Research Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load package functions from R/ directory
devtools::load_all(here::here())
```

## Introduction

Analysis of bill length and bill depth measurements across three penguin species
from the Palmer Station Antarctica LTER.

This report demonstrates the complete data workflow:
- Raw data → Processing → Analysis → Visualization

## Data Processing

```{r data-processing}
library(dplyr)
library(ggplot2)

# Load the processed data (function from R/data_utils.R)
penguins <- load_penguin_data()

# Preview data
head(penguins)
```

## Summary Statistics

```{r summary-stats}
# Calculate summary by species (function from R/data_utils.R)
summary_stats <- summarize_by_species(penguins)
knitr::kable(summary_stats, digits = 2, caption = "Bill dimensions by species")
```

## Visualization

### Bill Dimensions by Species

```{r bill-plot, fig.width=8, fig.height=6}
# Generate plot using function from R/bill_analysis.R
bill_plot <- create_bill_plot(penguins)
print(bill_plot)

# Save plot
dir.create(here::here("analysis/figures"), showWarnings = FALSE, recursive = TRUE)
ggplot2::ggsave(here::here("analysis/figures/bill_scatter.png"),
                bill_plot, width = 8, height = 6)
```

## Summary

The analysis reveals distinct bill dimension patterns across penguin species:
- Data processed from raw measurements (Step 3)
- Functions in R/ provide reusable analysis components
- Unit tests ensure correctness
````

### Render the Report

```bash
# Render the report
Rscript -e 'rmarkdown::render("analysis/report/report.Rmd")'

# Exit container (auto-snapshot runs here!)
exit
```

**Auto-Snapshot Architecture**

When you **exit R** (inside the container), zzcollab automatically:

1. Runs `renv::snapshot()` to capture package dependencies in `renv.lock`
2. Reports success and reminds you to commit changes

**You do not need to manually run `renv::snapshot()`** — just exit R and commit the updated `renv.lock`.

**Package restoration**: The container image is built with all packages from `renv.lock` baked in at `/opt/renv/library`. Auto-restore on startup is intentionally disabled (`ZZCOLLAB_AUTO_RESTORE=false` in the Dockerfile) so the baked library is authoritative and the container does not re-resolve packages from the network on every startup. If you add packages with `install.packages()` inside the container, exit R (auto-snapshot runs), rebuild the image (`make docker-build`), and the new packages will be baked in.

## Step 5: Validate and Commit

Validate everything works in a clean environment:

```bash
# Exit the Docker container (if still inside from Step 4)
# This automatically runs renv::snapshot() and validates packages!
exit

# Run tests in clean environment
make docker-test

# Optional: Manually validate dependencies (runs zzrenvcheck in the container)
make check-renv
```

Now commit your work:

```bash
git add .
git commit -m "$(cat <<'EOF'
Add complete data workflow: raw data to analysis

Data Workflow (5 Pillars):
- Raw data: analysis/data/raw_data/penguins_raw.csv (Pillar 5)
- Processing script: analysis/scripts/01_process_data.R (Pillar 4)
- Derived data: analysis/data/derived_data/penguins_clean.csv (Pillar 5)
- Functions: R/data_utils.R, R/bill_analysis.R (Pillar 4)
- Unit tests: inst/tinytest/test-data_utils.R (Pillar 4)
- Analysis report: analysis/report/report.Rmd (Pillar 4)

Reproducibility:
- Pillar 1: Dockerfile (computational environment)
- Pillar 2: renv.lock (dependency tracking)
- Pillar 3: .Rprofile (R session configuration)
- Pillar 4: Source code (scripts + functions + tests + reports)
- Pillar 5: Research data (raw + derived + documentation)
EOF
)"

git push
```

## Step 6: Verify CI/CD

Check that automated validation runs:

```bash
# View CI/CD status
gh run list --limit 3

# Or view in browser
gh repo view --web
```

GitHub Actions will automatically:

1. Build Docker container from scratch
2. Validate package dependencies (runs zzrenvcheck via make check-renv)
3. Run unit tests
4. Report results

**Note**: CI/CD runs `make check-renv`, which invokes
`zzrenvcheck::check_packages()` inside the container.

## Step 7: Verify Report Rendering CI/CD (Optional)

GitHub Actions is already configured to automatically render your report on every
push. The default workflow renders `analysis/report/report.Rmd`.

Verify the configuration:

```bash
# Check the render-report workflow (should already be correct)
cat .github/workflows/render-report.yml | grep "rmarkdown::render"

# Expected output:
# compendium-env Rscript -e 'rmarkdown::render("analysis/report/report.Rmd")'
```

If you used a different filename, update the workflow:
```bash
vim .github/workflows/render-report.yml
# Change the render command to match your report filename
git add .github/workflows/render-report.yml
git commit -m "Update CI/CD render path"
git push
```

**What this enables:**

- Automatic report rendering on every push to main
- Fresh computational environment for each render
- Downloadable PDF/HTML artifacts
- Verification that analysis is fully reproducible

## Step 8: Keeping Your Project Current (Optional)

zzcollab manages a set of template files in your compendium (the
Makefile, CI workflows, the `.Rprofile` template, and similar). Two minor
maintenance commands keep those in step with the installed zzcollab
version:

```bash
zzc update          # refresh framework-managed templates to this version
zzc doctor          # report on the project's state record
zzc doctor --fix    # back-fill or refresh the state record
```

`zzc update` is a minor maintenance command: it checks that the
project's framework-managed template files match the installed zzcollab
version, refreshes them when they have drifted, and back-fills the state
record. `zzc doctor` reports on the project's state record and, with
`--fix`, brings an older project's record up to date. Neither command
touches your data, code, or analysis. Run them after upgrading zzcollab
or when adopting a compendium created with an earlier version.

# What You Built

## All 5 Reproducibility Levels

**Level 1**: Basic R Project ✗ (skipped for reproducibility)

**Level 2**: renv ✓

- `renv.lock` records exact package versions
- `renv::snapshot()` captured dependencies

**Level 3**: Docker ✓

- Dockerfile specifies complete environment
- Anyone can reproduce with `make r` (launches image for current Dockerfile)

**Level 4**: Unit Testing ✓

- Tests validate plot creation
- Tests verify missing data handling
- Tests confirm species included

**Level 5**: CI/CD ✓

- GitHub Actions runs on every push
- Automated test suite execution
- Dependency validation

## The Five Pillars of Reproducibility

Your analysis is now fully reproducible thanks to five essential components that represent the necessary and sufficient elements for complete reproducibility:

### 1. Dockerfile - Computational Environment

- **R version**: Specifies exact R version (4.4.0)
- **System dependencies**: All system libraries (libcurl, libssl, etc.)
- **Base image**: Foundation for consistent environment
- **Environment variables**: Locale and system settings

```dockerfile
# Generated by zzcollab -- do not edit manually.
FROM rocker/tidyverse:4.4.2@sha256:<digest>   # pinned by zzc at init time

# OCI provenance labels
LABEL zzcollab.template.version="2.4.0" \
      zzcollab.r.version="4.4.2" \
      zzcollab.ppm.snapshot="2026-06-01"

# Library lives outside the project bind-mount so the baked environment
# is not shadowed when the host working tree is mounted at runtime.
ENV LANG=en_US.UTF-8 LC_ALL=en_US.UTF-8 TZ=UTC \
    RENV_PATHS_LIBRARY=/opt/renv/library \
    RENV_PATHS_CACHE=/opt/renv/cache \
    RENV_CONFIG_REPOS_OVERRIDE="https://packagemanager.posit.co/cran/__linux__/noble/2026-06-01" \
    ZZCOLLAB_CONTAINER=true \
    ZZCOLLAB_AUTO_RESTORE=false

RUN apt-get update && apt-get install -y --no-install-recommends \
    libcurl4-openssl-dev libssl-dev libxml2-dev && \
    rm -rf /var/lib/apt/lists/*

RUN R -e "install.packages('renv')"
RUN mkdir -p /opt/renv/library /opt/renv/cache
COPY renv.lock renv.lock
RUN R -e "renv::init(bare=TRUE, force=TRUE, restart=FALSE); renv::restore()"
```

### 2. renv.lock - Exact Package Versions

- **Every R package**: Exact version for all packages used
- **Complete dependency tree**: All package dependencies
- **Source repositories**: CRAN, Bioconductor, GitHub

```json
{
  "tidyverse": {
    "Version": "2.0.0",
    "Repository": "CRAN"
  }
}
```

### 3. .Rprofile - R Session Configuration

- **Critical R options**: Version controlled and copied into Docker
- **Pinned settings**: `stringsAsFactors`, `contrasts`, `na.action`, `digits`, `OutDec`
- **RNG discipline**: `RNGkind()` pinned to ensure seed reproducibility across R versions

```r
# .Rprofile (version controlled, generated by zzcollab)

# Pin RNG algorithm so stochastic analyses are reproducible across R versions.
# R 3.6.0 changed the default sample.kind, silently breaking seeded analyses.
RNGkind("Mersenne-Twister", "Inversion", "Rejection")

# Reproducibility options — applied on both host and in container
options(
  stringsAsFactors = FALSE,
  contrasts = c("contr.treatment", "contr.poly"),
  na.action = "na.omit",
  digits = 7,
  OutDec = "."
)
# renv activation and auto-snapshot-on-exit are handled automatically.
# Do not add source("renv/activate.R") manually; it is in the template.
```

**Important**: zzcollab's `.Rprofile` template pins the RNG kind and critical
session options. These settings are version-controlled and copied into Docker,
ensuring consistent behaviour across team members and CI/CD runs.

### 4. Analysis Source Code - Computational Logic

- **Processing scripts**: `analysis/scripts/01_process_data.R`
- **Reusable functions**: `R/data_utils.R`, `R/bill_analysis.R`
- **Reports**: `analysis/report/report.Rmd`
- **Tests**: `inst/tinytest/test-data_utils.R`
- **Random seeds**: Set explicitly in code for stochastic analyses. The
  `.Rprofile` pins `RNGkind()` to Mersenne-Twister/Inversion/Rejection so
  seeds behave consistently across R versions.

```r
# Example: Set seed for reproducible stochastic processes
set.seed(42)  # .Rprofile pins RNGkind so this seed is version-stable
model <- randomForest(species ~ ., data = penguins)
```

### 5. Research Data - Empirical Foundation

- **Raw data**: `analysis/data/raw_data/` (original, unmodified files)
- **Derived data**: `analysis/data/derived_data/` (processed, analysis-ready)
- **Data documentation**: `data/README.md` (data dictionary, provenance, processing lineage)
- **Quality assurance**: Validation checks, known issues, outlier documentation

```
analysis/data/
├── raw_data/
│   ├── penguins_raw.csv      # Original data (read-only)
│   └── README.md             # Data provenance and collection methods
└── derived_data/
    ├── penguins_clean.csv    # Processed data
    └── README.md             # Processing steps and transformations
```

**Critical insight**: Without the original research data, perfect computational reproducibility is meaningless. The analysis code operates on data, and that data must be preserved, documented, and version-controlled alongside the computational environment.

### Why Environment Variables Matter

Environment variables affect computational results silently and are critical for reproducibility:

**Locale settings** (`LANG`, `LC_ALL`):

- Control string sorting order (e.g., "Åland" sorts differently in US vs Swedish
  locales)
- Affect number parsing (comma vs period as decimal separator)
- Impact factor level ordering in statistical models

**Timezone** (`TZ`):

- Affects date-time arithmetic and aggregation
- Eliminates daylight saving time complications
- Ensures consistent temporal analyses across team members

**Parallel processing** (`OMP_NUM_THREADS`):

- Controls thread count for deterministic computation
- Single-threaded (`=1`) ensures reproducible results
- Multi-threaded can produce different results each run due to floating-point order

```dockerfile
# Environment variables set in Dockerfile
ENV LANG=en_US.UTF-8        # Standardize locale
ENV LC_ALL=en_US.UTF-8      # Override all locale categories
ENV TZ=UTC                  # No daylight saving time
ENV OMP_NUM_THREADS=1       # Deterministic single-threaded execution
```

**Real-world impact**: Two researchers running identical code in different locales or timezones can produce different results without explicit environment variable specification.

### How They Work Together:

```bash
# Someone reproduces your analysis 5 years from now:
git clone https://github.com/rgt47/penguin-bills.git
cd penguin-bills

# 1. Dockerfile builds environment (R version + system deps)
docker build -t penguin-env .
docker run -it -v $(pwd):/project penguin-env

# 2. renv.lock installs exact package versions
renv::restore()

# 3. Source code runs the analysis
source("analysis/scripts/01_process_data.R")
rmarkdown::render("analysis/report/report.Rmd")

# Result: Identical output to your original analysis
```

**Key Insight**:

- **Dockerfile** = Environment foundation (R version, system libraries,
  environment variables)
- **renv.lock** = Exact packages (source of truth for R dependencies)
- **.Rprofile** = R session settings (critical options that affect behavior)
- **Source Code** = Analytical logic (what actually runs, including random seeds)
- **Research Data** = Empirical observations (what analyses operate on)

All five pillars are version controlled in Git, representing the necessary and sufficient components for complete reproducibility. Without any one pillar, independent reproduction of the analysis becomes impossible.

### Reproducibility Checklist

When preparing analysis for publication or sharing, verify all five pillars.

For comprehensive best practices, troubleshooting, and team collaboration
patterns, see `vignette("reproducibility-layers")` and
`docs/collaborative-reproducibility.md`.

**Pillar 1 - Dockerfile**:

- R version explicitly specified
- System dependencies documented
- Environment variables set (LANG, LC_ALL, TZ, OMP_NUM_THREADS)

**Pillar 2 - renv.lock**:

- All packages from all contributors included
- Exact versions with repository sources
- Committed to version control

**Pillar 3 - .Rprofile**:

- Critical options version controlled and copied into Docker
- RNGkind pinned for seed reproducibility across R versions
- renv activation and auto-snapshot managed by the template automatically

**Pillar 4 - Source Code**:

- Random seeds explicit for all stochastic operations
- Analysis scripts, functions, tests all committed
- Clear computational workflow documented

**Pillar 5 - Research Data**:

- Raw data preserved (read-only, original files)
- Data documentation complete (data dictionary, provenance)
- Processing scripts link raw to derived data
- Data quality notes documented

## Project Structure

```
penguin-bills/
├── .github/
│   └── workflows/
│       └── r-package.yml        # Automated CI/CD validation
├── R/
│   ├── data_utils.R             # Data loading functions
│   └── bill_analysis.R          # Visualization functions
├── analysis/
│   ├── data/
│   │   ├── raw_data/            # Original, unmodified data
│   │   ├── derived_data/        # Processed analysis-ready data
│   │   └── README.md            # Data dictionary and documentation
│   ├── report/
│   │   ├── report.Rmd           # Analysis report (reproducible)
│   │   └── report.html          # Rendered output
│   ├── figures/
│   │   └── bill_scatter.png     # Generated plot
│   └── tables/                  # Generated tables
├── data/                        # R package data (optional, for package distribution)
├── scripts/                     # Working scripts and exploration
├── inst/
│   └── tinytest/
│       └── test-data_utils.R    # Unit tests
├── tests/
│   └── tinytest.R               # tinytest driver
├── man/                         # Generated function documentation
├── vignettes/                   # Package tutorials
├── docs/                        # Project documentation
├── renv/                        # renv package cache
├── .Rbuildignore                # Files excluded from package build
├── .gitignore                   # Git exclusions
├── DESCRIPTION                  # Project metadata and dependencies
├── NAMESPACE                    # Function exports (managed by roxygen2)
├── LICENSE                      # GPL-3 license
├── Makefile                     # Development workflow automation
├── Dockerfile                   # Environment specification
├── renv.lock                    # Exact package versions
└── penguin-bills.Rproj          # RStudio project file
```

## Verification Checklist

- Docker image built locally
- GitHub repository created (private)
- Analysis produces scatter plot
- Unit tests pass locally
- Dependencies captured in renv.lock
- CI/CD validates on every push
- Complete reproducibility achieved

# Next Steps

## Extend Your Analysis

```{r eval=FALSE}
# Add more plots
create_flipper_plot <- function(data = penguins) {
  # Flipper length vs body mass
}

# Add statistical analysis
fit_bill_model <- function(data) {
  lm(bill_depth_mm ~ bill_length_mm * species, data = data)
}

# Write more tests (tinytest, in inst/tinytest/)
model <- fit_bill_model(penguins)
expect_true('bill_length_mm:species' %in% names(coef(model)))
```

## Collaborate

The collaborative workflow below uses the profile commands for a quick
setup. To add or remove individual capabilities (CI, code-quality hooks,
data hashing, cloud launch) on a shared project, use the feature toggles
described in `vignette("quickstart3")`.

### Team Lead Workflow

Create a team project and share the Docker image:

```bash
# Create team project
mkdir study && cd study
zzc analysis                     # Full setup, prompts to build
zzc github                       # Create private GitHub repo

# Build image (installs packages from renv.lock into image)
make docker-build

# Optional: Push to DockerHub for faster team onboarding
zzc dockerhub                    # Team members can pull instead of build

# Add collaborators
gh repo edit --add-collaborator username
```

**Why build before push?** The build process installs all R packages from
`renv.lock` into the image. Team members who pull this image get packages
pre-installed (1-3 min) instead of building from scratch (5-15 min).

**Team Lead Responsibilities:**

1. Creates project with `zzc <profile>` (e.g., `zzc analysis`)
2. Adds GitHub with `zzc github`
3. Builds image with `make docker-build` (installs packages)
4. Optionally pushes to DockerHub with `zzc dockerhub`
5. Adds collaborators to GitHub repository
6. Rebuilds and pushes when renv.lock changes

### Team Member Workflow

Clone and join the team project:

```bash
# Clone the repository
git clone https://github.com/genomicslab/study.git
cd study

# Option A: Build from Dockerfile (always works)
make docker-build                # 5-15 minutes, verifies reproducibility
make r

# Option B: Pull pre-built image (faster, if team uses DockerHub)
docker pull genomicslab/study:latest   # 1-3 minutes
make r
```

**DockerHub is optional.** Building from the Dockerfile always works and
verifies full reproducibility. Pulling from DockerHub is faster for teams
with frequent onboarding but requires the team lead to push images.

**What happens for team members:**

1. Clone gets Dockerfile + renv.lock (reproducibility pillars)
2. Build or pull Docker image for identical environment
3. Add packages as needed using `install.packages()`
4. Exit container - packages automatically captured in renv.lock

**Environment Consistency:**

- **Docker environment**: Identical for all team members (build or pull)
- **Package management**: Add packages dynamically with `install.packages()`
- **Automatic capture**: Packages tracked automatically when you exit containers

## Publish Results

```bash
# Create manuscript
mkdir -p analysis/report
# Add report.Rmd with analysis

# Render report automatically via CI/CD
# (already configured in .github/workflows/render-report.yml)
```

# Troubleshooting

## GitHub Repository Creation Fails

```bash
# Create repository manually if needed
cd penguin-bills
gh repo create rgt47/penguin-bills --private --source=. --remote=origin --push
```

## Docker Build Errors

```bash
# Check Docker daemon running
docker info

# Clean rebuild (uses config defaults)
docker system prune
rm -rf penguin-bills  # Remove failed directory

# Retry project creation
mkdir penguin-bills && cd penguin-bills
zzc analysis
make docker-build
```

## Test Failures

```bash
# Run the tinytest suite (shows detailed failure messages)
make docker-test

# Or interactively in the container
make r
# tinytest::run_test_dir('inst/tinytest')
```

## CI/CD Failures

```bash
# View workflow logs
gh run view --log

# Validate locally first (make check-renv runs zzrenvcheck)
make check-renv
make docker-test
```

# Customizing Profiles and Features

## View Available Profiles

A profile is the Docker image bundle (which R packages are baked into the
container). It is separate from the archetype (the project type) and from
the features you toggle.

```bash
# List all predefined profiles
zzc list profiles

# List available library bundles (system dependencies)
zzc list libs

# List available package bundles (R packages)
zzc list pkgs
```

**Available profiles** (the three canonical profiles):

- `minimal` - Command-line R environment (~650MB)
- `analysis` - Tidyverse data analysis (~1.2GB), recommended
- `rstudio` - RStudio Server (~980MB)

For specialised needs (LaTeX, modeling, Shiny), start with the closest one and add
packages inside the container with `install.packages()`, then commit the
updated `renv.lock`.

## Switching Profiles

```bash
# Profile commands are smart:
# - New project: full setup (init + renv + docker + build prompt)
# - Existing project: switch profile (regenerate Dockerfile)

mkdir my-project && cd my-project
zzc analysis                     # Full setup with analysis profile
zzc github                       # Add GitHub (independent)

mkdir new-project && cd new-project
zzc rstudio                      # rstudio profile
zzc github --public              # Create public GitHub repo
```

## Toggling Features

Capabilities beyond the profile, such as Docker, CI, data-integrity
hashing, code-quality hooks, unit tests, and cloud launch, are *features*
you turn on and off at any time. Use the interactive wizard, or the
explicit `add` / `rm` forms:

```bash
zzc toggle                       # interactive feature checklist
zzc add code-quality             # add a single feature (styler + lintr)
zzc rm cloud                     # remove the devcontainer
zzc status                       # read-only view of the current state
```

Capture features (the backend and Docker) set the project's *capture
level* (L0 locatable, L1 pinned packages, L2 pinned environment, L3
verified); validation features (tests, CI, code quality) only check the
result without moving the level. The full toggle model, capture levels,
and the `zzc verify` workflow are documented in `vignette("quickstart3")`.

**Package management within a profile:**

- Packages are added dynamically using `install.packages()` inside
  containers.
- **Auto-snapshot**: packages are captured in `renv.lock` when you exit R
  in the container.
- The `renv.lock` file accumulates packages from all team members
  automatically.

# Summary

You created a fully reproducible analysis with:

- **Environment consistency**: Docker ensures identical R, packages, system
  libraries
- **Computational correctness**: Unit tests validate analytical logic
- **Automated verification**: CI/CD catches errors on every commit
- **Private collaboration**: GitHub private repo ready for team members
- **Professional workflow**: Industry-standard practices for research software

This is the gold standard for reproducible computational research. All 5 levels
working together ensure your analysis is not only reproducible, but also
correct and maintainable.

## Understanding the Architecture

ZZCOLLAB uses a two-layer package management system:

### Layer 1: Docker Environment (Team/Shared)

- **Controlled by**: Profile selection (e.g., `zzc analysis`)
- **Purpose**: Pre-installed packages in Docker image
- **Shared**: All team members get identical Docker environment

### Layer 2: Dynamic Package Installation (Personal)

- **Controlled by**: `install.packages()` inside containers (auto-captured on
  exit)
- **Purpose**: Add packages as needed for specific analyses
- **Collaborative**: renv.lock accumulates packages from all contributors

### Practical Example

Team lead sets up project:
```bash
mkdir study && cd study
zzc analysis                     # Full setup (init + renv + docker)
zzc github                       # Create private GitHub repo
zzc dockerhub                    # Optional: push to DockerHub for faster onboarding
```

Team members join and add packages:
```bash
git clone https://github.com/genomicslab/study.git
cd study

# Either build (always works) or pull (if team uses DockerHub)
make docker-build                # Option A: build from Dockerfile
# docker pull genomicslab/study:latest  # Option B: pull pre-built image

make r

# Inside container - add packages as needed
install.packages("tidymodels")
exit  # Auto-snapshot captures in renv.lock!

# Commit changes
git add renv.lock
git commit -m "Add tidymodels"
git push
```

**Key Insight**: The Docker image provides the foundation, but `renv.lock` ensures complete reproducibility by specifying exact versions of ALL packages.

For the feature-toggle model and capture levels, see
`vignette("quickstart3")`. For the capture/validation layers in depth,
see `vignette("reproducibility-layers")`.

---
*Rendered on 2026-06-19 at 08:36 PDT.*<br>
*Source: ~/prj/sfw/07-zzcollab/zzcollab/vignettes/quickstart1.Rmd*
