---
title: "ZZCOLLAB for R Users: Team Collaboration Workflow"
author: "ZZCOLLAB Team"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{ZZCOLLAB for R Users: Team Collaboration Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# ZZCOLLAB for R Users: Team Collaboration Workflow

This vignette describes collaborative research workflows using R functions that abstract containerization technologies. This approach is suitable for research teams with R expertise who require reproducible computational environments.

## The Team Collaboration Challenge

**Common challenges in collaborative research:**

- Version inconsistencies across different computational environments
- Unclear dependency specifications and requirements
- Non-reproducible analytical results due to environmental differences
- Complex setup procedures for new collaborators

**ZZCOLLAB approach:** Standardized computational environments accessible through R function calls.

## Team Roles and Workflow

### Role 1: Team Lead (Project Creator)
**Responsibilities:** Create the shared reproducible environment, set up the project infrastructure
**Skills needed:** R programming (Docker handled automatically)

### Role 2: Team Members (Collaborators)  
**Responsibilities:** Join the project, contribute analysis, maintain reproducibility
**Skills needed:** R programming only

### Role 3: Data Scientists/Analysts
**Responsibilities:** Focus on analysis without worrying about environment setup
**Skills needed:** R and your domain expertise

## Team Lead: Setting Up the Collaborative Environment

### Initial Team Configuration

The team lead sets up shared defaults that all team members will inherit:

```{r eval=FALSE}
library(zzcollab)

# Initialize team lead configuration
init_config()

# Set team-wide defaults (these become shared standards)
set_config("team_name", "datascience-lab")      # Your team's DockerHub namespace
set_config("github_account", "university-lab")  # Your organization's GitHub
set_config("build_mode", "comprehensive")       # Rich package environment for team
set_config("dotfiles_dir", "~/team-dotfiles")   # Shared configuration files

# View configuration to confirm
list_config()
```

### Creating a Team Research Project

```{r eval=FALSE}
# Create collaborative project with shared environment
# This is the ONLY command that requires Docker knowledge (handled automatically)
init_project(
  project_name = "customer-churn-analysis",
  team_name = "datascience-lab",
  build_mode = "comprehensive",  # Ensures rich package environment for team
  github_account = "university-lab"
)
```

**What this creates:**
- ✅ **Shared Docker environment** with all team packages pre-installed
- ✅ **GitHub repository** with proper access controls and CI/CD
- ✅ **R package structure** with team development standards
- ✅ **Reproducible infrastructure** that team members can join with one function

### Communicating with Team Members

After project creation, share these details with your team:

```{r eval=FALSE}
# Information to share with team members:
team_info <- list(
  team_name = "datascience-lab",
  project_name = "customer-churn-analysis",
  github_repo = "https://github.com/university-lab/customer-churn-analysis",
  instructions = "
    1. Install zzcollab R package
    2. Clone GitHub repository 
    3. Run: join_project('datascience-lab', 'customer-churn-analysis')
    4. Start analyzing with: start_rstudio()
  "
)

print(team_info)
```

## Team Members: Joining a Collaborative Project

### One-Time Setup for Team Members

```{r eval=FALSE}
# Install and configure zzcollab (one-time per team member)
library(zzcollab)

# Set personal configuration (inherits team defaults)
init_config()
set_config("github_account", "my-username")  # Your personal GitHub
set_config("dotfiles_dir", "~/my-dotfiles")  # Your personal config files (optional)

# View inherited configuration
list_config()
```

### Joining the Team Project

```{r eval=FALSE}
# Step 1: Clone the team repository (standard Git workflow)
# In terminal or RStudio: git clone https://github.com/university-lab/customer-churn-analysis.git
# Or use RStudio: File > New Project > Version Control > Git

# Step 2: Navigate to project and join team environment
setwd("customer-churn-analysis")  # or open .Rproj file
library(zzcollab)

# Join the team project (gets exact same environment as team lead)
join_project(
  team_name = "datascience-lab",
  project_name = "customer-churn-analysis"
)
```

**What this does:**
- ✅ **Downloads team environment** - Same R version, same packages, same versions
- ✅ **Sets up local development** - Configures your RStudio to use team environment  
- ✅ **Enables reproducibility** - Your results match teammates exactly
- ✅ **Handles complexity** - No Docker commands, all through R functions

### Starting Your Development Session

```{r eval=FALSE}
# Start RStudio in the team's reproducible environment
start_rstudio()  # Access at http://localhost:8787

# Alternative: Command-line R session
start_r_session()

# Your environment now matches your teammates exactly!
```

## Collaborative Development Example

Let's walk through a realistic team collaboration scenario with three team members analyzing customer churn.

### Team Lead: Sarah (Data Science Manager)

```{r eval=FALSE}
# Sarah sets up the collaborative infrastructure
library(zzcollab)

# Create team project for customer churn analysis
init_project(
  project_name = "churn-prediction-2024",
  team_name = "retail-analytics",
  build_mode = "comprehensive"  # Includes ML packages for the team
)

# Define the analysis structure for the team
# Creates analysis/01_data_prep.R, analysis/02_modeling.R, etc.
```

**Sarah shares with team:**
- GitHub repository: `https://github.com/retail-analytics/churn-prediction-2024`
- Team name: `retail-analytics`
- Project name: `churn-prediction-2024`

### Team Member 1: Mike (Data Scientist)

```{r eval=FALSE}
# Mike joins and focuses on data preparation
library(zzcollab)

# Join the team project
setwd("churn-prediction-2024")  # After cloning from GitHub
join_project("retail-analytics", "churn-prediction-2024")

# Start development
start_rstudio()

# In RStudio (identical environment to Sarah's):
library(tidyverse)
library(janitor)
library(skimr)

# Mike's contribution: data cleaning functions
clean_customer_data <- function(raw_data) {
  raw_data %>%
    clean_names() %>%
    filter(!is.na(customer_id)) %>%
    mutate(
      churn = as.factor(churn),
      tenure_years = tenure_months / 12,
      monthly_charges = as.numeric(monthly_charges)
    ) %>%
    select(-tenure_months)  # Remove original column
}

# Test function with team data
customer_data <- read_csv("data/raw_customer_data.csv")
clean_data <- clean_customer_data(customer_data)

# Save cleaned data for team
write_csv(clean_data, "data/clean_customer_data.csv")

# Commit Mike's contribution
git_commit("Add data cleaning pipeline

- Implement clean_customer_data() function
- Standardize column names and data types  
- Create tenure_years from tenure_months
- Save clean dataset for team modeling")

git_push()
```

### Team Member 2: Lisa (Machine Learning Specialist)

```{r eval=FALSE}
# Lisa joins and focuses on modeling
library(zzcollab)

# Join the same team project  
setwd("churn-prediction-2024")
join_project("retail-analytics", "churn-prediction-2024")

# Start development (same environment as Sarah and Mike)
start_rstudio()

# In RStudio:
library(tidymodels)
library(randomForest)
library(xgboost)

# Pull Mike's latest changes
git_pull()  # Through R interface

# Lisa's contribution: modeling functions
build_churn_models <- function(training_data) {
  # Split data
  data_split <- initial_split(training_data, strata = churn)
  train_set <- training(data_split)
  test_set <- testing(data_split)
  
  # Random forest model
  rf_spec <- rand_forest(trees = 500) %>%
    set_engine("randomForest") %>%
    set_mode("classification")
  
  rf_fit <- rf_spec %>%
    fit(churn ~ ., data = train_set)
  
  # XGBoost model  
  xgb_spec <- boost_tree(trees = 500) %>%
    set_engine("xgboost") %>%
    set_mode("classification")
    
  xgb_fit <- xgb_spec %>%
    fit(churn ~ ., data = train_set)
  
  # Return models and test set for evaluation
  list(
    rf_model = rf_fit,
    xgb_model = xgb_fit,
    test_data = test_set
  )
}

# Use Mike's cleaned data
clean_data <- read_csv("data/clean_customer_data.csv")
models <- build_churn_models(clean_data)

# Save models for team
saveRDS(models, "models/churn_models.rds")

# Commit Lisa's contribution
git_commit("Add machine learning models

- Implement build_churn_models() with RF and XGBoost
- Use tidymodels framework for consistency
- Save trained models for team evaluation
- Ready for performance analysis")

git_push()
```

### Team Collaboration: Combining Contributions

Now all team members can work with each other's contributions seamlessly:

```{r eval=FALSE}
# Any team member can run the complete pipeline
library(zzcollab)
setwd("churn-prediction-2024")
start_rstudio()

# In RStudio - combining everyone's work:
library(tidyverse)
library(tidymodels)

# Pull latest changes from all team members
git_pull()

# Use Mike's data cleaning
source("R/data_cleaning.R")
raw_data <- read_csv("data/raw_customer_data.csv")
clean_data <- clean_customer_data(raw_data)

# Use Lisa's models
source("R/modeling.R") 
models <- build_churn_models(clean_data)

# Evaluate models (could be Sarah or another team member's contribution)
evaluate_models <- function(models) {
  test_predictions_rf <- predict(models$rf_model, models$test_data)
  test_predictions_xgb <- predict(models$xgb_model, models$test_data)
  
  # Calculate metrics
  rf_accuracy <- accuracy_vec(models$test_data$churn, test_predictions_rf$.pred_class)
  xgb_accuracy <- accuracy_vec(models$test_data$churn, test_predictions_xgb$.pred_class)
  
  results <- tibble(
    model = c("Random Forest", "XGBoost"),
    accuracy = c(rf_accuracy, xgb_accuracy)
  )
  
  return(results)
}

# Everyone gets identical results!
model_performance <- evaluate_models(models)
print(model_performance)
```

## Team Management Functions

### For Team Leads

```{r eval=FALSE}
# Monitor team project status
team_images()  # See shared Docker environments
status()       # Check active containers

# Manage team environment
add_package("new_package")  # Add packages to team environment
sync_env()                  # Update team package requirements  
rebuild()                   # Rebuild team environment with new packages

# Coordinate team work
git_status()           # Check repository state
validate_repro()       # Ensure reproducibility across team
```

### For Team Members

```{r eval=FALSE}
# Stay synchronized with team
git_pull()            # Get latest changes from teammates
sync_env()            # Update your environment to match team
validate_repro()      # Check your setup matches team standards

# Contribute to team project
create_branch("feature/customer-segments")  # Work on feature branch
git_commit("Add customer segmentation analysis")
create_pr("Add customer segmentation", 
          "This analysis identifies three key customer segments for targeted retention")

# Get help when needed
zzcollab_help()       # Built-in help system
status()              # Check your environment status
```

## Advanced Team Collaboration Patterns

### Feature Branch Workflow

```{r eval=FALSE}
# Team member working on new feature
create_branch("feature/survival-analysis")
start_rstudio()

# Develop feature in RStudio...
# Add survival analysis for customer lifetime value

git_commit("Add survival analysis for customer LTV

- Implement Cox proportional hazards model
- Add Kaplan-Meier survival curves by segment  
- Generate LTV predictions based on survival probabilities
- Include comprehensive tests and documentation")

create_pr("Add customer survival analysis", 
          "Extends churn prediction with survival analysis to estimate customer lifetime value")
```

### Code Review Workflow

```{r eval=FALSE}
# Reviewer can perfectly reproduce the analysis
join_project("retail-analytics", "churn-prediction-2024")
git_checkout("feature/survival-analysis")  # Review branch
start_rstudio()

# Run the proposed analysis in identical environment
source("R/survival_analysis.R")
test_survival_analysis()  # All tests pass in identical environment

# Provide feedback through GitHub interface
```

### Multi-Project Team Coordination

```{r eval=FALSE}
# Team lead managing multiple related projects
projects <- c("churn-prediction-2024", "customer-segmentation", "pricing-optimization")

# Set up related projects with shared environments
for (project in projects) {
  init_project(
    project_name = project,
    team_name = "retail-analytics", 
    build_mode = "comprehensive"  # Consistent environment
  )
}

# Team members can join any project with consistent experience
# Same packages, same R version, same development environment
```

## Best Practices for Team Collaboration

### Communication Patterns

```{r eval=FALSE}
# Include reproducibility info in commit messages
git_commit("Add customer clustering analysis

Analysis Details:
- K-means clustering with 4 customer segments
- Validated using silhouette analysis
- Generated segment profiles and visualizations

Reproducibility:
- All package versions locked in renv.lock
- Random seed set to 12345 for consistent results  
- Analysis tested in team Docker environment")

# Use GitHub Issues for analysis coordination
# Use Pull Requests for code review
# Use project README for team setup instructions
```

### Environment Management

```{r eval=FALSE}
# Regular team environment maintenance
sync_env()          # Keep packages synchronized
validate_repro()    # Check reproducibility regularly
rebuild()           # Update environment when needed

# Document package additions
add_package("corrplot")  # Team lead adds package
git_commit("Add corrplot package for correlation analysis")
# Team members sync: sync_env() to get new package
```

### Quality Assurance

```{r eval=FALSE}
# Built-in reproducibility checks
validate_repro()  # Checks:
# - All team members have same package versions
# - All required packages are installed
# - All scripts can run successfully  
# - All data dependencies are satisfied

# Regular testing
run_script("tests/test_all_analysis.R")  # Run comprehensive tests
render_report("reports/team_analysis_report.Rmd")  # Generate team report
```

## Troubleshooting Team Issues

### Common Team Problems

```{r eval=FALSE}
# Problem: "It works for me but not for my teammate"
# Solution: Check environment synchronization
validate_repro()  # Both team members run this
sync_env()        # Update to match team environment

# Problem: "New team member can't reproduce results"  
# Solution: Ensure proper project joining
join_project("team-name", "project-name")  # Re-run join process
start_rstudio()                              # Verify environment

# Problem: "Package conflicts between team members"
# Solution: Use team environment management
rebuild()      # Team lead rebuilds shared environment
sync_env()     # All team members sync to new environment
```

### Getting Team Help

```{r eval=FALSE}
# Built-in team coordination help
zzcollab_help(team = TRUE)     # Team-specific help
zzcollab_next_steps(team = TRUE)  # What should the team do next
team_images()                  # Check shared environments
status()                       # Check individual setup
```

## Real-World Team Success Story

**Scenario:** University research lab with 5 members analyzing climate data

**Challenge:** Different R versions, package conflicts, "works on my laptop" syndrome

**ZZCOLLAB Solution:**

```{r eval=FALSE}
# Team Lead (Dr. Smith): One-time setup
init_project("climate-trends-2024", 
             team_name = "climate-lab", 
             build_mode = "comprehensive")

# 5 Team Members: Simple joining
join_project("climate-lab", "climate-trends-2024")
start_rstudio()

# Result: Identical environments, reproducible science, 6-month project completed in 3 months
```

**Observed benefits:**
- **Environmental consistency** across different systems
- **Cross-platform collaboration** independent of operating system
- **Reproducible research** suitable for publication requirements
- **Streamlined onboarding** for new team members
- **Standardized practices** without specialized technical knowledge requirements

## Summary for R Teams

ZZCOLLAB implements team collaboration through standard R interfaces:

**Team Lead:** Create shared environment with `init_project()`
**Team Members:** Join with `join_project()`, access via `start_rstudio()`  
**Reproducible Environments:** Consistent computational environments across team
**Standard Workflow:** Integration with RStudio, R functions, and Git
**Abstracted Complexity:** Container management handled programmatically
**Scalable Framework:** Supports varying team sizes from small groups to large research laboratories

Start your next team research project:

```{r eval=FALSE}
# Team Lead
library(zzcollab)
init_project("our-research-project", team_name = "our-lab")

# Team Members  
join_project("our-lab", "our-research-project")
start_rstudio()
# Begin collaborative analysis!
```

This framework provides systematic reproducibility and collaborative workflows through established R programming interfaces.