---
title: "Testing Data Analysis Workflows"
author: "ZZCOLLAB Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Testing Data Analysis Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Comprehensive Guide to Testing Data Analysis Workflows in R

## Table of Contents

1. [Introduction](#introduction)
2. [Testing Philosophy for Data Analysis](#testing-philosophy-for-data-analysis)
3. [Testing Framework Overview](#testing-framework-overview)
4. [Types of Tests for Data Analysis](#types-of-tests-for-data-analysis)
5. [Setting Up Your Testing Environment](#setting-up-your-testing-environment)
6. [Unit Testing for Analysis Functions](#unit-testing-for-analysis-functions)
7. [Integration Testing for Data Pipelines](#integration-testing-for-data-pipelines)
8. [Data Validation and Quality Testing](#data-validation-and-quality-testing)
9. [Reproducibility Testing](#reproducibility-testing)
10. [Testing Analysis Scripts](#testing-analysis-scripts)
11. [Testing Report Generation](#testing-report-generation)
12. [Continuous Integration for Data Analysis](#continuous-integration-for-data-analysis)
13. [Best Practices and Common Pitfalls](#best-practices-and-common-pitfalls)
14. [Tools and Packages](#tools-and-packages)
15. [Examples and Templates](#examples-and-templates)
16. [References](#references)

## Introduction

Testing data analysis workflows presents unique challenges compared to 
traditional software development. While package development focuses on 
testing isolated functions with predictable inputs and outputs, data 
analysis involves testing complex pipelines, validating data quality, 
ensuring reproducibility, and verifying that analytical results are 
correct and meaningful (Wilson et al., 2017; Wickham, 2015).

This guide provides comprehensive strategies for testing data analysis 
workflows in R, building upon established software engineering practices 
while addressing the specific needs of computational research and data 
science (Poldrack, 2019; Wilson et al., 2014).

## Testing Philosophy for Data Analysis

### Core Principles

Data analysis testing serves two primary goals:

1. **Computational Reproducibility**: Ensuring another researcher can use 
   your code and data to independently obtain identical results
2. **Result Correctness**: Validating that the results generated by the 
   code are accurate and meaningful

As noted by Wilson et al. (2017), "An important aspect of this is to 
validate the code using software development practices that prevent 
errors and software testing methods that can help detect them when they 
occur."

### Three-Phase Testing Approach

Following the reproducible research workflow principles (Poldrack, 2019), 
testing should align with the three phases of data analysis:

- **Explore Phase**: Rapid prototyping with lightweight validation
- **Refine Phase**: Comprehensive testing of stabilized analysis 
  components  
- **Produce Phase**: Full integration and reproducibility testing

## Testing Framework Overview

### The tinytest Package

ZZCOLLAB projects use the `tinytest` package for testing. It is a
lightweight, zero-dependency framework that requires no scaffolding:
test files contain bare, top-level expectation calls rather than tests
wrapped in a `test_that()` block. It offers:

- Rich expectation functions (`expect_equal()`, `expect_true()`,
  `expect_identical()`, `expect_inherits()`, and others)
- A flat file structure with no helper boilerplate
- Tests that double as runnable scripts during development

Test files live in `inst/tinytest/`, are driven by `tests/tinytest.R`,
and run via `make docker-test`, which executes
`tinytest::run_test_dir('inst/tinytest')` inside the container.

```{r, eval=FALSE}
# inst/tinytest/test_data_loading.R
data <- load_data('sample.csv')
expect_true(nrow(data) > 0)
expect_true('outcome' %in% names(data))
```

### Extended Testing Ecosystem

For data analysis, additional packages complement `tinytest`:

- **testdat**: Data validation testing (Hope, 2022)
- **validate**: Rule-based data validation (van der Loo & de Jonge, 2021)
- **assertr**: Assertive programming for data analysis pipelines 
  (Fischetti, 2021)
- **targets**: Pipeline testing and validation (Landau, 2021)

## Types of Tests for Data Analysis

### 1. Unit Tests

Test individual analysis functions in isolation:

```{r, eval=FALSE}
# inst/tinytest/test_outlier_detection.R
test_data <- c(1, 2, 3, 100, 4, 5)
outliers <- detect_outliers(test_data, method = 'iqr')
expect_equal(outliers, 100)
expect_equal(length(outliers), 1)
```

### 2. Integration Tests

Test how different components work together:

```{r, eval=FALSE}
# inst/tinytest/test_data_pipeline.R
# Test entire workflow from raw data to results
raw_data <- load_raw_data()
clean_data <- clean_data(raw_data)
model <- fit_model(clean_data)
results <- generate_results(model)

expect_inherits(results, 'analysis_results')
expect_true(results$converged)
```

### 3. Data Validation Tests

Ensure data quality and consistency:

```{r, eval=FALSE}
# inst/tinytest/test_data_quality.R
data <- load_analysis_data()

# Check data structure
expect_equal(ncol(data), 15)
expect_true(all(c('id', 'outcome', 'treatment') %in% names(data)))

# Check data ranges
expect_true(all(data$age >= 0 & data$age <= 120))
expect_true(all(data$outcome %in% c('success', 'failure')))

# Check for missing data patterns
missing_rate <- sum(is.na(data)) / (nrow(data) * ncol(data))
expect_true(missing_rate < 0.05)  # Less than 5% missing
```

### 4. Reproducibility Tests

Verify analysis can be reproduced:

```{r, eval=FALSE}
# inst/tinytest/test_reproducibility.R
set.seed(12345)
results1 <- run_analysis()

set.seed(12345)
results2 <- run_analysis()

expect_equal(results1$estimates, results2$estimates)
expect_equal(results1$p_values, results2$p_values)
```

## Setting Up Your Testing Environment

### Directory Structure

ZZCOLLAB projects place all test files in `inst/tinytest/`, with the
test runner in `tests/tinytest.R`. A flat layout is sufficient; group
related concerns by file name rather than by subdirectory:

```
tests/
└── tinytest.R              # Test runner
inst/
└── tinytest/
    ├── test_data_loading.R
    ├── test_data_cleaning.R
    ├── test_statistical_models.R
    ├── test_visualization.R
    ├── test_data_pipeline.R
    ├── test_report_rendering.R
    └── data/
        ├── sample-data.csv
        └── expected-outputs.rds
```

### Test Configuration

Create a `tests/tinytest.R` file that runs the suite when
`R CMD check` invokes it:

```{r, eval=FALSE}
# tests/tinytest.R
if (requireNamespace('tinytest', quietly = TRUE)) {
  tinytest::test_package('myproject')
}
```

Within the container, the same suite runs via `make docker-test`,
which calls `tinytest::run_test_dir('inst/tinytest')`.

### Helper Functions

Because tinytest files are plain scripts, define reusable test
utilities at the top of a file, or in a shared file sourced by the
tests that need it:

```{r, eval=FALSE}
# inst/tinytest/helpers.R
create_test_dataset <- function(n = 100) {
  data.frame(
    id = 1:n,
    treatment = sample(c('A', 'B'), n, replace = TRUE),
    outcome = rnorm(n, mean = 50, sd = 10),
    age = sample(18:80, n, replace = TRUE)
  )
}

expect_valid_model <- function(model) {
  expect_inherits(model, 'lm')
  expect_true(length(coef(model)) > 0)
  expect_true(summary(model)$r.squared >= 0)
}
```

## Unit Testing for Analysis Functions

### Statistical Functions

Test statistical computations with known results:

```{r, eval=FALSE}
# inst/tinytest/test_confidence_interval.R
# Test with known values
data <- c(10, 12, 14, 16, 18, 20)
ci <- calculate_ci(data, confidence = 0.95)

# Verify structure
expect_equal(length(ci), 2)
expect_equal(names(ci), c('lower', 'upper'))

# Verify mathematical properties
expect_true(ci$lower < mean(data))
expect_true(ci$upper > mean(data))
expect_true(ci$upper > ci$lower)
```

### Data Transformation Functions

Test data manipulation and cleaning:

```{r, eval=FALSE}
# inst/tinytest/test_imputation.R
test_data <- data.frame(
  x = c(1, 2, NA, 4, 5),
  y = c(10, NA, 30, 40, 50)
)

imputed <- impute_missing(test_data, method = 'mean')

# Check no missing values remain
expect_false(any(is.na(imputed)))

# Check imputation values are reasonable
expect_equal(imputed$x[3], mean(c(1, 2, 4, 5)))
expect_equal(imputed$y[2], mean(c(10, 30, 40, 50)))
```

### Visualization Functions

Test plot generation and properties:

```{r, eval=FALSE}
# inst/tinytest/test_scatter_plot.R
test_data <- create_test_dataset()

# Test plot creation
p <- create_scatter_plot(test_data, x = 'age', y = 'outcome')

# Check plot object
expect_inherits(p, 'ggplot')
expect_equal(length(p$layers), 1)

# Check plot data
plot_data <- ggplot2::ggplot_build(p)$data[[1]]
expect_equal(nrow(plot_data), nrow(test_data))
```

## Integration Testing for Data Pipelines

### Complete Pipeline Testing

Test the entire data analysis workflow:

```{r, eval=FALSE}
# inst/tinytest/test_full_pipeline.R
# Create temporary directory for outputs
temp_dir <- tempdir()
output_dir <- file.path(temp_dir, 'analysis_output')
dir.create(output_dir, recursive = TRUE)

# Run complete pipeline
results <- run_full_analysis(
  input_file = 'inst/tinytest/data/sample-data.csv',
  output_dir = output_dir
)

# Verify outputs exist
expect_true(file.exists(file.path(output_dir, 'results.csv')))
expect_true(file.exists(file.path(output_dir, 'plots.pdf')))

# Verify results structure
expect_inherits(results, 'analysis_results')
expect_true('model_summary' %in% names(results))
expect_true('diagnostics' %in% names(results))

# Cleanup
unlink(output_dir, recursive = TRUE)
```

### Cross-Script Dependencies

Test that analysis scripts work together:

```{r, eval=FALSE}
# inst/tinytest/test_script_sequence.R
scripts <- c(
  '01_data_preparation.R',
  '02_exploratory_analysis.R',
  '03_statistical_modeling.R',
  '04_results_visualization.R'
)

# Execute each script in an isolated environment
for (script in scripts) {
  script_path <- file.path('scripts', script)
  expect_true(file.exists(script_path))
  source(script_path, local = new.env())
}
```

## Data Validation and Quality Testing

### Using testdat for Data Validation

The `testdat` package provides specialized functions for testing data 
frames:

```{r, eval=FALSE}
# inst/tinytest/test_data_quality_rules.R
library(testdat)

data <- load_analysis_data()

# Test data completeness
expect_data(data, has_all_names, c('id', 'treatment', 'outcome'))
expect_data(data, has_nrows, min = 100)
expect_data(data, has_ncols, 10)

# Test data types
expect_data(data, is_col_type, 'id', 'integer')
expect_data(data, is_col_type, 'treatment', 'factor')
expect_data(data, is_col_type, 'outcome', 'numeric')

# Test value ranges
expect_data(data, has_range, 'age', min = 0, max = 120)
expect_data(data, has_no_missing, 'id')
```

### Custom Validation Rules

Create domain-specific validation functions:

```{r, eval=FALSE}
# inst/tinytest/test_clinical_validation.R
data <- load_clinical_data()

# Patient ID format
expect_true(all(grepl('^P\\d{6}$', data$patient_id)))

# Visit dates are reasonable
expect_true(all(data$visit_date >= as.Date('2020-01-01')))
expect_true(all(data$visit_date <= Sys.Date()))

# Biomarker values within expected ranges
expect_true(all(data$biomarker >= 0 & data$biomarker <= 1000))

# Treatment groups are balanced
treatment_table <- table(data$treatment)
expect_true(all(treatment_table >= 0.3 * nrow(data)))
```

## Reproducibility Testing

### Seed Management Testing

Ensure random processes are reproducible:

```{r, eval=FALSE}
# inst/tinytest/test_random_sampling.R
# Test bootstrap procedure
test_data <- create_test_dataset(100)

set.seed(42)
bootstrap1 <- bootstrap_analysis(test_data, n_boots = 1000)

set.seed(42)
bootstrap2 <- bootstrap_analysis(test_data, n_boots = 1000)

expect_equal(bootstrap1$estimates, bootstrap2$estimates)
expect_equal(bootstrap1$confidence_intervals,
             bootstrap2$confidence_intervals)
```

### Cross-Platform Reproducibility

Test that analysis works across different systems:

```{r, eval=FALSE}
# inst/tinytest/test_platform_independence.R
# Test numerical stability
test_data <- create_test_dataset(1000)

# Run analysis multiple times
results <- replicate(10, {
  model <- fit_complex_model(test_data)
  summary(model)$coefficients[1, 1]  # First coefficient
})

# Check consistency (allowing for floating point precision)
expect_true(sd(results) < 1e-10)
```

### Package Version Testing

Ensure results are stable across package versions:

```{r, eval=FALSE}
# inst/tinytest/test_package_versions.R
# Document package versions
packages <- c('lme4', 'ggplot2', 'dplyr')
versions <- sapply(packages, function(pkg) {
  as.character(packageVersion(pkg))
})

# Store with results for reproducibility tracking
for (pkg in packages) requireNamespace(pkg, quietly = TRUE)
expect_true(all(packages %in% loadedNamespaces()))

# Test that known results match
known_results <- readRDS('inst/tinytest/data/known_results_v1.rds')
current_results <- run_standard_analysis()

expect_equal(current_results$coefficients,
             known_results$coefficients,
             tolerance = 1e-6)
```

## Testing Analysis Scripts

### Script Execution Testing

Test that numbered analysis scripts run without error:

```{r, eval=FALSE}
# inst/tinytest/test_scripts_execute.R
script_dir <- 'scripts'
scripts <- list.files(script_dir, pattern = '^\\d+_.*\\.R$',
                      full.names = TRUE)
scripts <- sort(scripts)  # Ensure numerical order

# Execute each script in a clean environment
for (script in scripts) {
  script_env <- new.env()
  source(script, local = script_env)
}
```

### Output Validation

Test that scripts produce expected outputs:

```{r, eval=FALSE}
# inst/tinytest/test_script_outputs.R
# Run data preparation script
source('scripts/01_data_preparation.R', local = new.env())

# Check that cleaned data was created
expect_true(file.exists('data/derived_data/cleaned_data.rds'))

# Validate cleaned data
cleaned_data <- readRDS('data/derived_data/cleaned_data.rds')
expect_true(nrow(cleaned_data) > 0)
expect_false(any(is.na(cleaned_data$primary_outcome)))
```

## Testing Report Generation

### R Markdown Testing

Test that reports can be rendered:

```{r, eval=FALSE}
# inst/tinytest/test_report_render.R
report_path <- 'analysis/report/report.Rmd'
output_dir <- tempdir()

# Test YAML parsing
yaml_content <- rmarkdown::yaml_front_matter(report_path)
expect_true('title' %in% names(yaml_content))
expect_true('author' %in% names(yaml_content))

# Test dependencies
required_packages <- c('rmarkdown', 'knitr', 'ggplot2')
for (pkg in required_packages) {
  expect_true(requireNamespace(pkg, quietly = TRUE))
}

# Test rendering only when LaTeX is available. tinytest has no
# skip helper, so guard with a conditional instead.
if (requireNamespace('tinytex', quietly = TRUE)) {
  rmarkdown::render(report_path,
                    output_dir = output_dir,
                    quiet = TRUE)

  # Verify PDF was created
  pdf_file <- file.path(output_dir, 'report.pdf')
  expect_true(file.exists(pdf_file))
}
```

### Content Validation

Test report content quality:

```{r, eval=FALSE}
# inst/tinytest/test_report_content.R
report_path <- 'analysis/report/report.Rmd'
content <- readLines(report_path)

# Check for required sections
expect_true(any(grepl('# Introduction', content)))
expect_true(any(grepl('# Methods', content)))
expect_true(any(grepl('# Results', content)))
expect_true(any(grepl('# Discussion', content)))

# Check for figure references
expect_true(any(grepl('Figure \\\\@ref', content)))

# Check for table references
expect_true(any(grepl('Table \\\\@ref', content)))
```

## Continuous Integration for Data Analysis

### GitHub Actions Configuration

Example workflow for testing data analysis projects:

```yaml
# .github/workflows/analysis-testing.yml
name: Analysis Testing

on: [push, pull_request]

jobs:
  test:
    runs-on: ubuntu-latest
    
    steps:
    - uses: actions/checkout@v3
    
    - uses: r-lib/actions/setup-r@v2
      with:
        r-version: '4.3.0'
        
    - uses: r-lib/actions/setup-r-dependencies@v2
      with:
        cache-version: 2
        
    - name: Run tinytest suite
      run: |
        Rscript -e "tinytest::run_test_dir('inst/tinytest')"

    - name: Check reproducibility
      run: |
        Rscript -e "source('scripts/99_reproducibility_check.R')"

    - name: Validate package dependencies
      run: |
        make check-renv
```

The `make check-renv` target runs the companion `zzrenvcheck` package
inside the project's Docker image. It executes
`zzrenvcheck::check_packages(auto_fix = TRUE, strict = TRUE)`, which
verifies that every package referenced in code is declared in both
`DESCRIPTION` and `renv.lock`. The host wrapper `zzcollab validate`
performs the same check. Variants `make check-renv-no-fix` (report
only) and `make check-renv-no-strict` (skip `tests/` and `vignettes/`)
are available.

### Quality Gates

Implement automated quality checks:

```{r, eval=FALSE}
# scripts/99_reproducibility_check.R
library(tinytest)

# Check renv lockfile is current
expect_true(renv::status()$synchronized)

# Check R options are documented
source('check_rprofile_options.R')

# Check data integrity
data_files <- list.files('data/raw_data', full.names = TRUE)
for (file in data_files) {
  checksum <- digest::digest(file, file = TRUE)
  # Compare with stored checksums
  expect_true(verify_checksum(file, checksum))
}
```

## Best Practices and Common Pitfalls

### Best Practices

1. **Test Early and Often**: Implement tests during the exploratory phase, 
   not just at the end
2. **Use Meaningful Test Names**: Describe what the test validates, not 
   just what it does
3. **Test Edge Cases**: Include tests for missing data, extreme values, 
   and empty datasets
4. **Mock External Dependencies**: Use sample data instead of requiring 
   database connections
5. **Document Expected Behavior**: Use informative error messages and 
   comments

### Common Pitfalls

1. **Testing Implementation Instead of Behavior**: Focus on what the 
   function should do, not how
2. **Overly Specific Tests**: Avoid tests that break with minor, 
   acceptable changes
3. **Ignoring Random Elements**: Always set seeds when testing functions 
   with randomness
4. **Insufficient Data Validation**: Test data assumptions explicitly
5. **Skipping Integration Tests**: Unit tests alone do not catch pipeline 
   failures

### Performance Considerations

```{r, eval=FALSE}
# inst/tinytest/test_performance.R
start_time <- Sys.time()

results <- run_analysis(large_dataset)

end_time <- Sys.time()
duration <- as.numeric(difftime(end_time, start_time, units = 'mins'))

expect_true(duration < 5)  # Should complete within 5 minutes
```

## Tools and Packages

### Core Testing Framework

- **tinytest**: Primary testing framework for ZZCOLLAB projects; a
  lightweight, zero-dependency alternative to testthat (van der Loo,
  2020)
- **devtools**: Development tools, including helpers for running checks
- **renv**: Environment reproducibility, validated via `make check-renv`

### Data Validation

- **testdat**: Specialized data frame testing (Hope, 2022)
- **validate**: Rule-based data validation (van der Loo & de Jonge, 2021)
- **assertr**: Pipeline assertions and data verification (Fischetti, 2021)
- **checkmate**: Fast argument checks and assertions

### Reproducibility

- **targets**: Pipeline management and testing (Landau, 2021)
- **renv**: Environment reproducibility (Ushey, 2022)
- **here**: Path management for reproducible scripts
- **conflicted**: Managing package conflicts

### Continuous Integration

- **rcmdcheck**: R CMD check automation
- **covr**: Code coverage analysis
- **lintr**: Code style checking

## Examples and Templates

### Simple Examples with Built-in Datasets

#### Basic Data Validation Tests with iris Dataset

```{r, eval=FALSE}
# inst/tinytest/test_iris_analysis.R
data(iris)

# Test basic structure
expect_equal(nrow(iris), 150)
expect_equal(ncol(iris), 5)
expect_true(all(c('Sepal.Length', 'Sepal.Width', 'Petal.Length',
                  'Petal.Width', 'Species') %in% names(iris)))

# Test data types
expect_true(is.numeric(iris$Sepal.Length))
expect_true(is.numeric(iris$Sepal.Width))
expect_true(is.factor(iris$Species))

# Test value ranges
expect_true(all(iris$Sepal.Length > 0))
expect_true(all(iris$Petal.Length >= 0))
expect_true(max(iris$Sepal.Length) <= 10)  # Reasonable upper bound

# Test species levels
expected_species <- c('setosa', 'versicolor', 'virginica')
expect_equal(levels(iris$Species), expected_species)
expect_equal(as.vector(table(iris$Species)), rep(50L, 3))

# Measurements should be biologically plausible
expect_true(all(iris$Sepal.Length >= 4 & iris$Sepal.Length <= 8))
expect_true(all(iris$Sepal.Width >= 1.5 & iris$Sepal.Width <= 5))
expect_true(all(iris$Petal.Length >= 1 & iris$Petal.Length <= 7))
expect_true(all(iris$Petal.Width >= 0.1 & iris$Petal.Width <= 3))

# No missing values
expect_false(any(is.na(iris)))
```

#### Simple Statistical Functions with mtcars

```{r, eval=FALSE}
# inst/tinytest/test_mtcars_stats.R

# Helper function for testing
calculate_mpg_summary <- function(data) {
  list(
    mean_mpg = mean(data$mpg),
    median_mpg = median(data$mpg),
    sd_mpg = sd(data$mpg),
    range_mpg = range(data$mpg)
  )
}

data(mtcars)

# mpg summary statistics are correct
summary_stats <- calculate_mpg_summary(mtcars)

expect_true(summary_stats$mean_mpg > 0)
expect_true(summary_stats$median_mpg > 0)
expect_true(summary_stats$sd_mpg > 0)
expect_equal(length(summary_stats$range_mpg), 2)
expect_true(summary_stats$range_mpg[1] <= summary_stats$range_mpg[2])

# Test against known values (mtcars is static)
expect_equal(summary_stats$mean_mpg, mean(mtcars$mpg))
expect_equal(summary_stats$median_mpg, median(mtcars$mpg))
expect_true(abs(summary_stats$mean_mpg - 20.09) < 0.1)

# Correlation analysis: mpg versus weight
cor_mpg_wt <- cor(mtcars$mpg, mtcars$wt)
expect_true(cor_mpg_wt >= -1 & cor_mpg_wt <= 1)
expect_true(cor_mpg_wt < 0)  # Should be negative correlation
expect_true(abs(cor_mpg_wt) > 0.5)  # Should be strong correlation

# Linear model fitting
model <- lm(mpg ~ wt + hp, data = mtcars)
expect_inherits(model, 'lm')
expect_equal(length(coef(model)), 3)  # Intercept + 2 predictors
expect_true(summary(model)$r.squared > 0.5)
expect_true(summary(model)$r.squared <= 1)

# Test residuals
residuals <- resid(model)
expect_equal(length(residuals), nrow(mtcars))
expect_true(abs(mean(residuals)) < 1e-10)  # Approximately zero
```

#### Basic Visualization Testing with penguins

```{r, eval=FALSE}
# inst/tinytest/test_penguins_viz.R
library(ggplot2)
library(palmerpenguins)

# Helper function
create_penguin_scatterplot <- function(data, x_var, y_var) {
  ggplot(data, aes(x = .data[[x_var]], y = .data[[y_var]],
                   color = species)) +
    geom_point() +
    theme_minimal() +
    labs(title = paste('Penguins:', y_var, 'vs', x_var))
}

data(penguins, package = 'palmerpenguins')

# Remove missing values for plotting
clean_penguins <- penguins[complete.cases(penguins), ]
expect_true(nrow(clean_penguins) > 0)

# Create plot
p <- create_penguin_scatterplot(clean_penguins, 'bill_length_mm',
                                'body_mass_g')

# Test plot object
expect_inherits(p, 'ggplot')
expect_equal(length(p$layers), 1)  # One geom_point layer

# Test plot data
plot_data <- ggplot_build(p)$data[[1]]
expect_equal(nrow(plot_data), nrow(clean_penguins))
expect_true(all(c('x', 'y', 'colour') %in% names(plot_data)))

# Validate species separation
species_summary <- aggregate(body_mass_g ~ species,
                             data = clean_penguins,
                             FUN = function(x) c(mean = mean(x),
                                                 sd = sd(x)))

expect_equal(nrow(species_summary), 3)  # Three species
expect_true(all(species_summary$body_mass_g[, 'mean'] > 0))
expect_true(all(species_summary$body_mass_g[, 'sd'] > 0))

# Gentoo penguins should be heaviest on average
gentoo_mean <- species_summary[species_summary$species == 'Gentoo',
                               'body_mass_g'][1, 'mean']
adelie_mean <- species_summary[species_summary$species == 'Adelie',
                               'body_mass_g'][1, 'mean']
expect_true(gentoo_mean > adelie_mean)
```

### Medium Complexity Examples

#### Data Pipeline Testing with iris

```{r, eval=FALSE}
# inst/tinytest/test_iris_pipeline.R

# Analysis pipeline functions
prepare_iris_data <- function(data) {
  # Add derived variables
  data$sepal_ratio <- data$Sepal.Length / data$Sepal.Width
  data$petal_ratio <- data$Petal.Length / data$Petal.Width
  data$total_length <- data$Sepal.Length + data$Petal.Length
  data
}

analyze_species_differences <- function(data) {
  # Perform ANOVA for each measurement
  results <- list()
  measurements <- c('Sepal.Length', 'Sepal.Width', 'Petal.Length',
                    'Petal.Width')

  for (measure in measurements) {
    formula_str <- paste(measure, '~ Species')
    aov_result <- aov(as.formula(formula_str), data = data)
    results[[measure]] <- summary(aov_result)
  }

  results
}

data(iris)

# Data preparation pipeline works
prepared_data <- prepare_iris_data(iris)

expect_true('sepal_ratio' %in% names(prepared_data))
expect_true('petal_ratio' %in% names(prepared_data))
expect_true('total_length' %in% names(prepared_data))

expect_true(all(prepared_data$sepal_ratio > 0))
expect_true(all(prepared_data$petal_ratio > 0))
expect_true(all(prepared_data$total_length > prepared_data$Sepal.Length))
expect_true(all(prepared_data$total_length > prepared_data$Petal.Length))

expect_false(any(is.na(prepared_data$sepal_ratio)))
expect_false(any(is.na(prepared_data$petal_ratio)))

# Species analysis produces valid results
analysis_results <- analyze_species_differences(prepared_data)

expect_true(is.list(analysis_results))
expect_equal(length(analysis_results), 4)
expect_true(all(c('Sepal.Length', 'Sepal.Width', 'Petal.Length',
                  'Petal.Width') %in% names(analysis_results)))

for (result in analysis_results) {
  expect_true(is.list(result))
  expect_true(length(result) == 1)  # One ANOVA table

  # Check F-statistic exists and is positive
  f_stat <- result[[1]][1, 'F value']
  expect_true(f_stat > 0)

  # Check p-value is between 0 and 1
  p_value <- result[[1]][1, 'Pr(>F)']
  expect_true(p_value >= 0 & p_value <= 1)
}

# Complete pipeline is reproducible
set.seed(123)
prepared1 <- prepare_iris_data(iris)
results1 <- analyze_species_differences(prepared1)

set.seed(123)
prepared2 <- prepare_iris_data(iris)
results2 <- analyze_species_differences(prepared2)

expect_equal(prepared1, prepared2)
expect_equal(results1, results2)
```

#### Advanced Statistical Testing with mtcars

```{r, eval=FALSE}
# inst/tinytest/test_mtcars_advanced.R

# Advanced analysis functions
perform_regression_analysis <- function(data, response, predictors) {
  formula_str <- paste(response, "~", paste(predictors, collapse = " + "))
  model <- lm(as.formula(formula_str), data = data)
  
  # Add model diagnostics
  list(
    model = model,
    summary = summary(model),
    diagnostics = list(
      residuals = resid(model),
      fitted = fitted(model),
      r_squared = summary(model)$r.squared,
      adj_r_squared = summary(model)$adj.r.squared,
      aic = AIC(model),
      bic = BIC(model)
    )
  )
}

validate_model_assumptions <- function(model_result) {
  residuals <- model_result$diagnostics$residuals
  fitted <- model_result$diagnostics$fitted
  
  # Shapiro-Wilk test for normality (if sample size allows)
  if (length(residuals) <= 5000) {
    normality_test <- shapiro.test(residuals)
  } else {
    normality_test <- NULL
  }
  
  # Durbin-Watson test for autocorrelation
  if (requireNamespace("lmtest", quietly = TRUE)) {
    dw_test <- lmtest::dwtest(model_result$model)
  } else {
    dw_test <- NULL
  }
  
  list(
    normality = normality_test,
    durbin_watson = dw_test,
    residual_range = range(residuals),
    mean_residual = mean(residuals)
  )
}

data(mtcars)

# Regression analysis function works correctly
result <- perform_regression_analysis(mtcars, 'mpg',
                                      c('wt', 'hp', 'cyl'))

expect_true(is.list(result))
expect_true(all(c('model', 'summary', 'diagnostics') %in% names(result)))

expect_inherits(result$model, 'lm')
expect_equal(length(coef(result$model)), 4)  # Intercept + 3 predictors

expect_true(result$diagnostics$r_squared >= 0 &
            result$diagnostics$r_squared <= 1)
expect_true(result$diagnostics$adj_r_squared <=
            result$diagnostics$r_squared)
expect_true(result$diagnostics$aic > 0)
expect_true(result$diagnostics$bic > 0)
expect_equal(length(result$diagnostics$residuals), nrow(mtcars))

# Model validation runs and reports diagnostics
good_result <- perform_regression_analysis(mtcars, 'mpg', c('wt', 'hp'))
good_validation <- validate_model_assumptions(good_result)

expect_true(is.list(good_validation))
expect_true(abs(good_validation$mean_residual) < 1e-10)  # Near zero
expect_true(length(good_validation$residual_range) == 2)
expect_true(good_validation$residual_range[1] <=
            good_validation$residual_range[2])

# Regression handles different model specifications
simple_model <- perform_regression_analysis(mtcars, 'mpg', 'wt')
complex_model <- perform_regression_analysis(mtcars, 'mpg',
                                             c('wt', 'hp', 'cyl',
                                               'disp', 'qsec'))

expect_true(length(coef(simple_model$model)) <
            length(coef(complex_model$model)))
expect_true(complex_model$diagnostics$r_squared >=
            simple_model$diagnostics$r_squared)
expect_inherits(simple_model$model, 'lm')
expect_inherits(complex_model$model, 'lm')
```

#### Integration Testing with penguins Dataset

```{r, eval=FALSE}
# inst/tinytest/test_penguins_integration.R
library(palmerpenguins)
library(ggplot2)

# Complete analysis workflow
run_penguin_analysis <- function(remove_na = TRUE) {
  # Load and prepare data
  data(penguins, package = "palmerpenguins")
  
  if (remove_na) {
    clean_data <- penguins[complete.cases(penguins), ]
  } else {
    clean_data <- penguins
  }
  
  # Descriptive statistics
  summary_stats <- list()
  numeric_vars <- c("bill_length_mm", "bill_depth_mm", "flipper_length_mm", 
                   "body_mass_g")
  
  for (var in numeric_vars) {
    if (remove_na) {
      values <- clean_data[[var]]
    } else {
      values <- clean_data[[var]][!is.na(clean_data[[var]])]
    }
    
    summary_stats[[var]] <- list(
      mean = mean(values),
      sd = sd(values),
      median = median(values),
      range = range(values)
    )
  }
  
  # Statistical models
  models <- list()
  
  # Model 1: Body mass predicted by bill dimensions
  if (remove_na) {
    models$body_mass_model <- lm(body_mass_g ~ bill_length_mm + 
                                bill_depth_mm, data = clean_data)
  }
  
  # Model 2: Flipper length by species
  if (remove_na) {
    models$flipper_species_model <- aov(flipper_length_mm ~ species, 
                                       data = clean_data)
  }
  
  # Visualizations
  plots <- list()
  
  if (remove_na && nrow(clean_data) > 0) {
    # Scatterplot
    plots$bill_scatter <- ggplot(clean_data, 
                                aes(x = bill_length_mm, 
                                    y = bill_depth_mm, 
                                    color = species)) +
      geom_point() +
      theme_minimal() +
      labs(title = "Bill Dimensions by Species")
    
    # Box plot
    plots$mass_boxplot <- ggplot(clean_data, 
                                aes(x = species, y = body_mass_g, 
                                    fill = species)) +
      geom_boxplot() +
      theme_minimal() +
      labs(title = "Body Mass by Species")
  }
  
  # Return comprehensive results
  list(
    data = clean_data,
    summary_stats = summary_stats,
    models = models,
    plots = plots,
    sample_size = nrow(clean_data)
  )
}

# Complete workflow executes successfully
results <- run_penguin_analysis(remove_na = TRUE)

expect_true(is.list(results))
expect_true(all(c('data', 'summary_stats', 'models', 'plots',
                  'sample_size') %in% names(results)))

expect_inherits(results$data, 'data.frame')
expect_true(results$sample_size > 300)  # Substantial sample
expect_false(any(is.na(results$data)))  # No missing values after cleaning

expect_equal(length(results$summary_stats), 4)
for (stat in results$summary_stats) {
  expect_true(all(c('mean', 'sd', 'median', 'range') %in% names(stat)))
  expect_true(stat$mean > 0)
  expect_true(stat$sd > 0)
}

expect_inherits(results$models$body_mass_model, 'lm')
expect_inherits(results$models$flipper_species_model, 'aov')

expect_inherits(results$plots$bill_scatter, 'ggplot')
expect_inherits(results$plots$mass_boxplot, 'ggplot')

# Results are biologically meaningful: Gentoo are heaviest
species_masses <- aggregate(body_mass_g ~ species, data = results$data,
                            mean)
gentoo_mass <- species_masses[species_masses$species == 'Gentoo',
                              'body_mass_g']
adelie_mass <- species_masses[species_masses$species == 'Adelie',
                              'body_mass_g']
chinstrap_mass <- species_masses[species_masses$species == 'Chinstrap',
                                 'body_mass_g']

expect_true(gentoo_mass > adelie_mass)
expect_true(gentoo_mass > chinstrap_mass)

# Body mass model has explanatory power
model_summary <- summary(results$models$body_mass_model)
expect_true(model_summary$r.squared > 0.1)
expect_true(model_summary$fstatistic[1] > 1)

# Species differences in flipper length are significant
anova_result <- summary(results$models$flipper_species_model)
f_stat <- anova_result[[1]][1, 'F value']
p_value <- anova_result[[1]][1, 'Pr(>F)']
expect_true(f_stat > 1)
expect_true(p_value < 0.05)

# Analysis is reproducible across repeated runs
results1 <- run_penguin_analysis(remove_na = TRUE)
results2 <- run_penguin_analysis(remove_na = TRUE)

expect_equal(results1$data, results2$data)
expect_equal(results1$sample_size, results2$sample_size)
expect_equal(results1$summary_stats, results2$summary_stats)
expect_equal(coef(results1$models$body_mass_model),
             coef(results2$models$body_mass_model))
```

### Complete Test Suite Template

```{r, eval=FALSE}
# inst/tinytest/test_complete_analysis.R
library(myanalysis)

# Unit tests for core functions: data loading
csv_data <- load_csv_data('inst/tinytest/data/sample.csv')
expect_inherits(csv_data, 'data.frame')

xlsx_data <- load_excel_data('inst/tinytest/data/sample.xlsx')
expect_inherits(xlsx_data, 'data.frame')

# Statistical functions are correct
test_data <- data.frame(
  group = rep(c('A', 'B'), each = 50),
  value = c(rnorm(50, 10, 2), rnorm(50, 12, 2))
)

result <- compare_groups(test_data, 'value', 'group')
expect_inherits(result, 'htest')
expect_true(result$p.value > 0 & result$p.value <= 1)

# Integration: complete workflow executes
data <- load_analysis_data()
cleaned <- clean_and_validate(data)
models <- fit_models(cleaned)
results <- summarize_results(models)
plots <- create_visualizations(results)

expect_inherits(results, 'analysis_summary')
expect_true(length(plots) > 0)

# Reproducibility: identical seeds yield identical results
set.seed(123)
result1 <- run_bootstrap_analysis(n_bootstrap = 100)

set.seed(123)
result2 <- run_bootstrap_analysis(n_bootstrap = 100)

expect_equal(result1, result2)
```

### Data Validation Template

```{r, eval=FALSE}
# inst/tinytest/test_data_validation.R
library(testdat)

data <- load_trial_data()

# Structure validation
expect_data(data, has_all_names,
            c('patient_id', 'treatment', 'outcome', 'baseline_score'))
expect_data(data, has_nrows, min = 200)  # Minimum sample size

# Data type validation
expect_data(data, is_col_type, 'patient_id', 'character')
expect_data(data, is_col_type, 'treatment', 'factor')
expect_data(data, is_col_type, 'outcome', 'numeric')

# Value validation
expect_data(data, has_range, 'outcome', min = 0, max = 100)
expect_data(data, has_no_missing, 'patient_id')

# Business logic validation
treatment_groups <- levels(data$treatment)
expect_true(all(c('placebo', 'active') %in% treatment_groups))

# Check randomization balance
group_sizes <- table(data$treatment)
expect_true(all(group_sizes >= 90 & group_sizes <= 110))
```

## References

Fischetti, T. (2021). assertr: Assertive Programming for R Analysis 
Pipelines. R package version 2.8. 
https://CRAN.R-project.org/package=assertr

Hope, R. (2022). testdat: Data Unit Testing for R. R package version 
0.4.0. https://CRAN.R-project.org/package=testdat

Landau, W. M. (2021). The targets R package: a dynamic Make-like 
function-oriented pipeline toolkit for reproducibility and 
high-performance computing. Journal of Open Source Software, 6(57), 2959. 
https://doi.org/10.21105/joss.02959

Poldrack, R. A. (2019). The new mind readers: What neuroimaging can and 
cannot reveal about our thoughts. Princeton University Press.

Poldrack, R. A., et al. (2019). Reproducible data analysis. In Stanford 
Psychology Guide to Doing Open Science. 
https://poldrack.github.io/psych-open-science-guide/4_reproducibleanalysis.html

Ushey, K. (2022). renv: Project Environments. R package version 0.16.0. 
https://CRAN.R-project.org/package=renv

van der Loo, M. (2020). A method for deriving information from running R
code (tinytest: lightweight, no-dependency unit testing). The R Journal,
12(1), 454-462. https://CRAN.R-project.org/package=tinytest

van der Loo, M., & de Jonge, E. (2021). Data validation infrastructure
for R. Journal of Statistical Software, 97(10), 1-31.
https://doi.org/10.18637/jss.v097.i10

Wickham, H. (2011). testthat: Get started with testing. The R Journal, 
3(1), 5-10. 
https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf

Wickham, H. (2015). R packages: organize, test, document, and share your 
code. O'Reilly Media.

Wilson, G., Aruliah, D. A., Brown, C. T., Hong, N. P. C., Davis, M., Guy, 
R. T., ... & Wilson, P. (2014). Best practices for scientific computing. 
PLoS biology, 12(1), e1001745. 
https://doi.org/10.1371/journal.pbio.1001745

Wilson, G., Bryan, J., Cranston, K., Kitzes, J., Nederbragt, L., & Teal, 
T. K. (2017). Good enough practices in scientific computing. PLoS 
computational biology, 13(6), e1005510. 
https://doi.org/10.1371/journal.pcbi.1005510

---

*This guide is part of the ZZCOLLAB framework for reproducible research 
collaboration. For more information, see the project documentation and 
workflow guides.*