---
title: "Plot Viewing with zzvim-R Terminal Graphics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plot Viewing with zzvim-R Terminal Graphics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

The `zzvim-R` vim plugin provides automatic plot display in kitty and iTerm2
terminals. Unlike manual save-and-view approaches, zzvim-R intercepts R's
graphics device and displays plots automatically when they are created.

This vignette covers:

- Installation and setup
- Auto-display functions (`zzplot`, `zzggplot`)
- Plot history navigation
- Configuration options (size, alignment, relative sizing)
- Split pane viewing
- Integration with zzcollab Docker workflows

## Prerequisites

- [Kitty terminal](https://sw.kovidgoyal.net/kitty/) or iTerm2 (macOS)
- [zzvim-R plugin](https://github.com/rgt47/zzvim-R) installed in vim/neovim
- R project with `.Rprofile` or `DESCRIPTION` file

## Installation

### Install zzvim-R Plugin

**vim-plug:**

```vim
Plug 'rgt47/zzvim-R'
```

**Manual:**

```bash
git clone https://github.com/rgt47/zzvim-R ~/.vim/pack/plugins/start/zzvim-R
```

### Automatic Setup

When you open vim in a kitty terminal within an R project, zzvim-R automatically:

1. Detects the terminal type (kitty or iTerm2)
2. Creates `.Rprofile.local` with terminal graphics functions
3. Adds `.Rprofile.local` to `.gitignore`

No manual configuration required.

### Manual Setup

If automatic setup does not activate, copy the template manually:

```bash
# From zzvim-R plugin directory
cp ~/.vim/pack/plugins/start/zzvim-R/templates/.Rprofile.local .

# Or from zzcollab if using Docker
cp /path/to/zzvim-R/templates/.Rprofile.local .
```

## Terminal Detection

zzvim-R detects the terminal type automatically:

| Terminal | Detection Method | Features |
|----------|------------------|----------|
| **Kitty** | `$KITTY_WINDOW_ID` | Full support + alignment control |
| **iTerm2** | `$ITERM_SESSION_ID` | Full support (no alignment) |
| **Other** | - | Disabled |

Check your terminal type in R:

```{r}
Sys.getenv("KITTY_WINDOW_ID")  # Non-empty if kitty
Sys.getenv("ITERM_SESSION_ID") # Non-empty if iTerm2
```

## Basic Usage

### Auto-Display Functions

zzvim-R provides wrapper functions that automatically display plots:

```{r}
# Base R plots
zzplot(mtcars$wt, mtcars$mpg, main = "Weight vs MPG")

# ggplot2 plots
library(ggplot2)
p <- ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  theme_minimal()
zzggplot(p)
```
Plots appear directly in the terminal without manual `kitty +kitten icat` calls.

### How It Works

zzvim-R overrides `dev.off()` to automatically display plots after rendering:

1. `zzplot()` or `zzggplot()` creates a temporary PNG file
2. When the device closes, the plot displays via `kitty +kitten icat`
3. The plot is added to history for later navigation

## Plot History

Navigate through previously displayed plots:

```{r}
# View plot history
plot_history()
#> Plot history: 5 plots
#> Current: Plot 3
#>
#>   [1] /tmp/plot_abc123.png
#>   [2] /tmp/plot_def456.png
#> * [3] /tmp/plot_ghi789.png
#>   [4] /tmp/plot_jkl012.png
#>   [5] /tmp/plot_mno345.png

# Navigate backward
plot_prev()
#> Plot 2 of 5

# Navigate forward
plot_next()
#> Plot 3 of 5
```

## Saving Plots

### Save as PNG

```{r}
# Create and display a plot
zzggplot(p)

# Save the current plot
save_plot("analysis/figures/mpg_by_weight.png")
#> Plot saved to: analysis/figures/mpg_by_weight.png
```

### Save as PDF

```{r}
# Convert current plot to PDF
plot_to_pdf("analysis/figures/mpg_by_weight.pdf")
#> PDF saved to: analysis/figures/mpg_by_weight.pdf
```

## Configuration Options

### Fixed Plot Size

Set specific pixel dimensions:

```{r}
# Set plot dimensions (width, height, dpi)
set_plot_size(1200, 800, 150)
#> Plot size set to 1200x800 @ 150dpi

# Future plots use these dimensions
zzggplot(p)
```

### Relative Sizing (Recommended)

Size plots as a percentage of terminal dimensions:

```{r}
# Default: 75% width, 100% height
set_plot_size_relative()
#> Relative sizing enabled: 720x576px
#> (75% x 100% of terminal)
#> Plots will auto-resize when terminal is resized

# Custom percentages
set_plot_size_relative(width_pct = 0.5, height_pct = 0.8)
#> Relative sizing enabled: 480x460px
#> (50% x 80% of terminal)
```

### Plot Alignment (Kitty Only)

Control horizontal alignment of plots in the terminal:

```{r}
# Align to right side (default)
set_plot_align("right")

# Center the plot
set_plot_align("center")

# Align to left side
set_plot_align("left")
```

### Redisplay After Resize

If you resize your terminal, redisplay the current plot:

```{r}
plot_redisplay_if_resized()
#> Terminal resized from 120x40 to 160x50
#> Redisplaying current plot...
```

## Split Pane Viewing

Open the current plot in a separate kitty split pane:

```{r}
# Create a plot
zzggplot(p)

# Open in vertical split
plot_split()
#> Plot opened in split pane (use Ctrl+Shift+] to focus plot, Ctrl+Shift+[ to return to R)
```

### Kitty Navigation

| Shortcut | Action |
|----------|--------|
| `Ctrl+Shift+]` | Focus next pane (plot) |
| `Ctrl+Shift+[` | Focus previous pane (R) |
| `Ctrl+Shift+w` | Close current pane |

### Scaled Split View

Open a scaled version of the plot:

```{r}
# Open at 150% scale
plot_split(scale = 1.5)

# Open at 50% scale (for high-resolution plots)
plot_split(scale = 0.5)
```

## Docker Integration

When using zzcollab Docker containers with kitty, the terminal graphics work
automatically if `.Rprofile.local` is in the mounted project directory.

### Workflow

```bash
# 1. Start vim in kitty terminal
cd ~/projects/myproject
vim analysis/scripts/analysis.R

# 2. Open terminal split and start Docker R
:below terminal
make r
```

```{r}
# 3. In R (inside Docker container)
library(ggplot2)

p <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  theme_minimal()

# Auto-displays in kitty
zzggplot(p)

# Save for reports
save_plot("analysis/figures/scatter.png")
```

### Ensuring .Rprofile.local is Loaded

The zzcollab `.Rprofile` template automatically sources `.Rprofile.local`:

```{r}
# In .Rprofile (auto-generated by zzcollab)
if (file.exists(".Rprofile.local")) {
  source(".Rprofile.local")
}
```

If plots are not auto-displaying, verify:

```{r}
# Check if terminal graphics are loaded
exists("zzplot")
#> [1] TRUE

# Check terminal detection
.terminal
#> [1] "kitty"
```

## Function Reference

### Plot Display

| Function | Description |
|----------|-------------|
| `zzplot(...)` | Base R plot with auto-display |
| `zzggplot(p)` | ggplot2 plot with auto-display |
| `plot_split(scale)` | Open current plot in split pane |

### Plot History

| Function | Description |
|----------|-------------|
| `plot_history()` | List all plots in session |
| `plot_prev()` | Navigate to previous plot |
| `plot_next()` | Navigate to next plot |

### Saving

| Function | Description |
|----------|-------------|
| `save_plot(filename)` | Save current plot as PNG |
| `plot_to_pdf(filename)` | Save current plot as PDF |

### Configuration

| Function | Description |
|----------|-------------|
| `set_plot_size(w, h, dpi)` | Set fixed plot dimensions |
| `set_plot_size_relative(w_pct, h_pct)` | Set size as % of terminal |
| `set_plot_align(align)` | Set alignment (kitty only) |
| `plot_redisplay_if_resized()` | Redisplay after terminal resize |

## Startup Message

When terminal graphics are enabled, R displays:

```
✓ Terminal graphics enabled (KITTY)

Quick reference:
  zzplot(...)                : Base R plots with auto-display
  zzggplot(p)                : ggplot2 plots with auto-display
  plot_history()             : View plot history
  plot_prev() / plot_next()  : Navigate history
  save_plot(file)            : Save current plot as PNG
  plot_to_pdf(file)          : Save current plot as PDF
  set_plot_size(w,h,dpi)     : Fixed dimensions
  set_plot_size_relative()   : Size by % of terminal (default: 75% x 100%)
  set_plot_align(a)          : Alignment (left/center/right)
  plot_split()               : Open current plot in split pane
  plot_redisplay_if_resized(): Redisplay after terminal resize
```

## Troubleshooting

### Plots Not Displaying

1. **Verify kitty terminal:**

```bash
echo $KITTY_WINDOW_ID
# Should output a non-empty value
```

2. **Check .Rprofile.local exists:**

```bash
ls -la .Rprofile.local
```

3. **Verify functions are loaded:**

```{r}
exists("zzplot")
exists(".terminal")
```

4. **Test icat manually:**

```bash
kitty +kitten icat --help
```

### Wrong Terminal Detected

Check environment variables:

```{r}
Sys.getenv("KITTY_WINDOW_ID")
Sys.getenv("ITERM_SESSION_ID")
Sys.getenv("TERM_PROGRAM")
```

### Plots Too Large/Small

Use relative sizing for automatic adjustment:

```{r}
set_plot_size_relative(width_pct = 0.5, height_pct = 0.75)
```

### Split Pane Not Opening

Verify kitty remote control is enabled. Add to `~/.config/kitty/kitty.conf`:

```
allow_remote_control yes
```

## Comparison with Manual Approach

| Feature | Manual (`kp()`) | zzvim-R |
|---------|-----------------|---------|
| **Auto-display** | No | Yes |
| **Plot history** | No | Yes |
| **Alignment control** | No | Yes (kitty) |
| **Relative sizing** | No | Yes |
| **Split pane** | No | Yes |
| **Docker compatible** | Yes | Yes |
| **Setup required** | Function in .Rprofile | Plugin + template |

For simple workflows, the manual `kp()` function (see `vignette("workflow-kitty-graphics")`)
may be sufficient. For vim-based development with frequent plotting, zzvim-R provides
a more integrated experience.

## See Also

- [zzvim-R Documentation](https://github.com/rgt47/zzvim-R)
- [Kitty Graphics Protocol](https://sw.kovidgoyal.net/kitty/graphics-protocol/)
- [ZZCOLLAB Kitty Graphics Workflow](workflow-kitty-graphics.html)
