---
title: "R Package Development with ZZCOLLAB"
author: "ZZCOLLAB Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{R Package Development with ZZCOLLAB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

This vignette demonstrates the recommended workflow for R package development using ZZCOLLAB, following the unified research compendium framework [@marwick2018] and R package development best practices [@wickham2015].

**Key Philosophy**: Use the **same structure** for package development as data analysis. No separate "package paradigm"—just add package infrastructure progressively as needed [@marwick2018].

## What You Will Learn

- Developing R packages following rrtools conventions
- Test-driven development with comprehensive coverage [@wickham2011]
- Documentation with roxygen2 [@wickham2015]
- CRAN compliance and submission
- Continuous integration for multi-platform testing
- Publishing to CRAN and beyond

## Prerequisites

- Docker installed and running
- ZZCOLLAB installed (`./install.sh`)
- Familiarity with R package structure [@wickham2015]
- Git configured with credentials
- GitHub account for package hosting

# The Unified Compendium Approach to Packages

## There Is No "Package Paradigm"

Traditional approach: "Analysis" vs "Package" are separate project types.

**ZZCOLLAB approach** [@marwick2018]: **Same structure, progressive disclosure**.

```
Day 1 (Analysis):        Month 3 (Add Package Docs):    Month 6 (CRAN-ready):
analysis/                analysis/                       analysis/
├── data/               ├── data/                       ├── data/
└── scripts/            └── scripts/                    └── scripts/
R/                      R/                              R/
└── functions.R         └── functions.R                 └── functions.R
                        man/                            man/
                        └── *.Rd (auto-generated)       └── *.Rd
                        tests/                          tests/
                        └── testthat/                   └── testthat/
                                                        vignettes/
                                                        └── tutorial.Rmd
                                                        NEWS.md
                                                        cran-comments.md
```

**No migration needed**—add CRAN infrastructure when ready [@wickham2015].

## When to Develop as a Package

Develop as a package when [@wickham2015]:
- **Reusable functions** - Code will be used across multiple projects
- **Team collaboration** - Multiple developers contributing
- **Public distribution** - Sharing via CRAN, Bioconductor, GitHub
- **Formal documentation** - Need help files and vignettes
- **Quality assurance** - Want automated testing and validation

# Step-by-Step: Package Development Workflow

## Step 1: Project Initialization (Same as Analysis!)

The package development workflow starts **identically** to data analysis [@marwick2018].

### Configure ZZCOLLAB

```bash
# One-time configuration
zzcollab config init
zzcollab config set team-name "myusername"
zzcollab config set github-account "myusername"
zzcollab config set profile-name "ubuntu_standard_analysis"  # Analysis profile works for packages!
```

### Create Package Project

```bash
# Create project directory (use package name)
mkdir penguintools && cd penguintools

# Initialize ZZCOLLAB project (identical to analysis workflow)
zzcollab

# Build Docker environment
make docker-build
```

**What this creates** - **Identical to analysis workflow**:

```
penguintools/
├── analysis/
│   ├── data/
│   │   ├── raw_data/              # Example data for development
│   │   └── derived_data/          # Processed examples
│   ├── figures/                   # Generated plots for README/vignettes
│   └── scripts/                   # Development scripts
├── R/                             # Package functions
├── tests/                         # Unit tests
├── DESCRIPTION                    # Package metadata
├── Dockerfile                     # Development environment
├── renv.lock                      # Package dependencies
└── Makefile                       # Convenient commands
```

**Key insight**: This is **exactly the same structure** as data analysis. The only difference is **what you put in R/** [@marwick2018].

## Step 2: Develop Functions with Tests (TDD Approach)

Following test-driven development [@wickham2011], write tests first, then implement functions.

### Initialize Testing

```bash
make r

R
> usethis::use_testthat()
> quit()

exit
```

### Example: Penguin Data Preparation Function

#### Step 2a: Write Test First

Create `tests/testthat/test-prepare_data.R`:

```r
# test-prepare_data.R
library(testthat)

test_that("prepare_penguin_data removes incomplete cases", {
  # Create test data with missing values
  test_data <- data.frame(
    species = c("Adelie", "Adelie", "Chinstrap"),
    bill_length_mm = c(39.1, NA, 46.5),
    bill_depth_mm = c(18.7, 17.8, 17.9),
    body_mass_g = c(3750, 3800, NA),
    sex = c("male", "female", "male")
  )

  result <- prepare_penguin_data(test_data)

  # Should keep only complete cases
  expect_equal(nrow(result), 1)
  expect_false(anyNA(result$bill_length_mm))
  expect_false(anyNA(result$body_mass_g))
})

test_that("prepare_penguin_data creates log transformation", {
  test_data <- data.frame(
    species = c("Adelie"),
    bill_length_mm = c(39.1),
    bill_depth_mm = c(18.7),
    body_mass_g = c(3750),
    sex = c("male")
  )

  result <- prepare_penguin_data(test_data)

  # Log variable should exist and be correct
  expect_true("log_body_mass_g" %in% names(result))
  expect_equal(result$log_body_mass_g, log(3750))
})

test_that("prepare_penguin_data validates input", {
  # Not a data frame
  expect_error(prepare_penguin_data("not a data frame"),
               "must be a data frame")

  # Missing required columns
  incomplete <- data.frame(x = 1, y = 2)
  expect_error(prepare_penguin_data(incomplete),
               "Missing required columns")
})
```

#### Step 2b: Run Test (Should Fail)

```bash
make docker-test

# Expected: Tests fail - function doesn't exist yet
```

#### Step 2c: Implement Function

Create `R/data_functions.R`:

```r
#' Prepare penguin data for analysis
#'
#' Removes incomplete cases for core morphological variables and creates
#' log-transformed body mass variable for allometric analyses.
#'
#' @param data Data frame containing penguin measurements. Must include columns:
#'   \code{bill_length_mm}, \code{bill_depth_mm}, \code{body_mass_g}, \code{sex}.
#'
#' @return Data frame with complete cases and added \code{log_body_mass_g} variable.
#'   Retains columns: species, island, bill measurements, body mass (original and log),
#'   sex, and year.
#'
#' @details
#' This function implements the data preparation protocol documented in
#' \code{analysis/scripts/01_prepare_data.R}. Missing values in any of the four
#' core variables (bill length, bill depth, body mass, sex) result in case removal.
#' The log transformation enables allometric scaling analyses [@schmidt-nielsen_scaling_1984].
#'
#' @examples
#' \dontrun{
#' library(palmerpenguins)
#' clean_data <- prepare_penguin_data(penguins)
#' summary(clean_data)
#' }
#'
#' @seealso \code{\link{fit_body_mass_model}} for allometric model fitting
#'
#' @references
#' Schmidt-Nielsen, K. (1984). *Scaling: Why Is Animal Size So Important?*
#' Cambridge University Press.
#'
#' @export
prepare_penguin_data <- function(data) {
  # Input validation following R package best practices (Wickham 2015)
  if (!is.data.frame(data)) {
    stop("Input 'data' must be a data frame", call. = FALSE)
  }

  required_cols <- c("bill_length_mm", "bill_depth_mm", "body_mass_g", "sex")
  missing_cols <- setdiff(required_cols, names(data))

  if (length(missing_cols) > 0) {
    stop(
      "Missing required columns: ",
      paste(missing_cols, collapse = ", "),
      call. = FALSE
    )
  }

  # Data preparation
  data %>%
    dplyr::filter(
      !is.na(bill_length_mm),
      !is.na(bill_depth_mm),
      !is.na(body_mass_g),
      !is.na(sex)
    ) %>%
    dplyr::mutate(log_body_mass_g = log(body_mass_g)) %>%
    dplyr::select(
      species, island,
      bill_length_mm, bill_depth_mm,
      body_mass_g, log_body_mass_g,
      sex, year
    )
}
```

#### Step 2d: Run Test (Should Pass)

```bash
make docker-test

# Expected: All tests pass
```

### Complete the Package with Additional Functions

Following the same TDD approach, add more functions:

- `fit_body_mass_model()` - Allometric scaling model
- `plot_bill_dimensions()` - Visualization
- `calculate_species_summary()` - Summary statistics

See **Data Analysis vignette** for examples.

## Step 3: Documentation with roxygen2

Comprehensive documentation is essential for package users [@wickham2015].

### Document All Exported Functions

Each function in `R/` needs roxygen2 documentation:

```r
#' @title - Brief description (1 line)
#' @description - Detailed description (multiple paragraphs OK)
#' @param - Document each parameter
#' @return - Describe return value
#' @details - Additional implementation details
#' @examples - Executable examples
#' @seealso - Related functions
#' @references - Citations for methods
#' @export - Make function available to users
```

### Generate Documentation

```bash
make r

R
> devtools::document()  # Generates man/*.Rd files from roxygen2 comments
> quit()

exit
```

### Verify Documentation

```bash
# Check all functions are documented
R CMD check

# Or use ZZCOLLAB convenience target
make check
```

## Step 4: Create Package Vignettes

Vignettes provide long-form documentation for package workflows [@wickham2015].

### Create Tutorial Vignette

```bash
make r

R
> usethis::use_vignette("penguin-analysis-tutorial")
> quit()

exit
```

Edit `vignettes/penguin-analysis-tutorial.Rmd`:

````markdown
---
title: "Penguin Morphology Analysis Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Penguin Morphology Analysis Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette demonstrates how to use the `penguintools` package to analyze
Palmer Penguins morphological data.

## Data Preparation

```{r}
library(penguintools)
library(palmerpenguins)

# Prepare data
clean_data <- prepare_penguin_data(penguins)
```

## Allometric Analysis

```{r}
# Fit scaling model
model <- fit_body_mass_model(clean_data)
summary(model)
```

## Visualization

```{r, fig.width=7, fig.height=5}
# Create bill dimensions plot
plot_bill_dimensions(clean_data)
```
````

### Build Vignettes

```bash
make r

R
> devtools::build_vignettes()
> quit()

exit
```

## Step 5: CRAN Compliance

CRAN has strict requirements for package submission [@wickham2015].

### Update DESCRIPTION File

```yaml
Package: penguintools
Title: Tools for Palmer Penguins Morphological Analysis
Version: 0.1.0
Authors@R:
    person("Your", "Name", , "your.email@example.com", role = c("aut", "cre"),
           comment = c(ORCID = "YOUR-ORCID-ID"))
Description: Provides data preparation, statistical modeling, and visualization
    tools for analyzing Palmer Penguins morphological data. Implements
    allometric scaling analyses following Schmidt-Nielsen (1984).
License: MIT + file LICENSE
Encoding: UTF-8
Roxygen: list(markdown = TRUE)
RoxygenNote: 7.2.3
Imports:
    dplyr (>= 1.0.0),
    ggplot2 (>= 3.3.0)
Suggests:
    testthat (>= 3.0.0),
    palmerpenguins,
    knitr,
    rmarkdown
VignetteBuilder: knitr
URL: https://github.com/yourusername/penguintools
BugReports: https://github.com/yourusername/penguintools/issues
```

### Add LICENSE File

```bash
make r

R
> usethis::use_mit_license("Your Name")
> quit()

exit
```

### Add NEWS.md

Create `NEWS.md` to document version history:

```markdown
# penguintools 0.1.0

* Initial CRAN submission
* Core functions for penguin morphology analysis:
  - `prepare_penguin_data()`: Data preparation
  - `fit_body_mass_model()`: Allometric scaling
  - `plot_bill_dimensions()`: Visualization
* Comprehensive test coverage (>95%)
* Tutorial vignette
```

### Add cran-comments.md

Create `cran-comments.md` for CRAN maintainers:

```markdown
## Test environments
* local: macOS (M1), R 4.3.1 (via Docker)
* GitHub Actions:
  - ubuntu-latest (release, devel, oldrel-1)
  - windows-latest (release)
  - macos-latest (release)

## R CMD check results

0 errors | 0 warnings | 0 notes

## Downstream dependencies

There are currently no downstream dependencies for this package.
```

## Step 6: Comprehensive Testing

Achieve >95% test coverage following best practices [@wickham2011].

### Test Coverage Report

```bash
make r

R
> library(covr)
> cov <- package_coverage()
> percent_coverage(cov)
# Target: >95%

> # Generate HTML report
> report(cov)

> quit()
exit
```

### Test Edge Cases

Ensure tests cover [@wickham2011]:
- **Normal cases**: Expected inputs produce expected outputs
- **Edge cases**: Empty data, single observation, extreme values
- **Error cases**: Invalid inputs produce informative errors
- **Boundary conditions**: Min/max values, NA handling

Example comprehensive test:

```r
test_that("prepare_penguin_data handles all edge cases", {
  # Normal case
  normal_data <- data.frame(
    species = c("Adelie", "Adelie"),
    bill_length_mm = c(39.1, 40.3),
    bill_depth_mm = c(18.7, 18.0),
    body_mass_g = c(3750, 3900),
    sex = c("male", "female")
  )
  result <- prepare_penguin_data(normal_data)
  expect_equal(nrow(result), 2)

  # Empty data
  empty <- normal_data[0, ]
  expect_equal(nrow(prepare_penguin_data(empty)), 0)

  # Single observation
  single <- normal_data[1, ]
  expect_equal(nrow(prepare_penguin_data(single)), 1)

  # All missing values
  all_na <- normal_data
  all_na$body_mass_g <- NA
  expect_equal(nrow(prepare_penguin_data(all_na)), 0)

  # Extreme values (should be kept - data-driven)
  extreme <- normal_data
  extreme$body_mass_g[1] <- 10000  # Very large penguin!
  expect_equal(nrow(prepare_penguin_data(extreme)), 2)

  # Invalid input type
  expect_error(prepare_penguin_data("not a data frame"))
  expect_error(prepare_penguin_data(list(x = 1, y = 2)))

  # Missing required columns
  incomplete <- data.frame(x = 1, y = 2)
  expect_error(prepare_penguin_data(incomplete), "Missing required columns")
})
```

## Step 7: Multi-Platform CI/CD

Test on all CRAN platforms using GitHub Actions.

### GitHub Actions Workflow

Create `.github/workflows/R-CMD-check.yml`:

```yaml
name: R-CMD-check

on:
  push:
    branches: [main, master]
  pull_request:
    branches: [main, master]

jobs:
  R-CMD-check:
    runs-on: ${{ matrix.config.os }}

    name: ${{ matrix.config.os }} (${{ matrix.config.r }})

    strategy:
      fail-fast: false
      matrix:
        config:
          - {os: macos-latest,   r: 'release'}
          - {os: windows-latest, r: 'release'}
          - {os: ubuntu-latest,  r: 'devel', http-user-agent: 'release'}
          - {os: ubuntu-latest,  r: 'release'}
          - {os: ubuntu-latest,  r: 'oldrel-1'}

    steps:
      - uses: actions/checkout@v4

      - uses: r-lib/actions/setup-r@v2
        with:
          r-version: ${{ matrix.config.r }}
          http-user-agent: ${{ matrix.config.http-user-agent }}
          use-public-rspm: true

      - uses: r-lib/actions/setup-r-dependencies@v2
        with:
          extra-packages: any::rcmdcheck
          needs: check

      - uses: r-lib/actions/check-r-package@v2
        with:
          upload-snapshots: true
```

### Test Coverage Workflow

Create `.github/workflows/test-coverage.yml`:

```yaml
name: test-coverage

on:
  push:
    branches: [main, master]
  pull_request:
    branches: [main, master]

jobs:
  test-coverage:
    runs-on: ubuntu-latest

    steps:
      - uses: actions/checkout@v4

      - uses: r-lib/actions/setup-r@v2
        with:
          use-public-rspm: true

      - uses: r-lib/actions/setup-r-dependencies@v2
        with:
          extra-packages: any::covr
          needs: coverage

      - name: Test coverage
        run: |
          covr::codecov(
            quiet = FALSE,
            clean = FALSE,
            install_path = file.path(Sys.getenv("RUNNER_TEMP"), "package")
          )
        shell: Rscript {0}

      - name: Show testthat output
        if: always()
        run: |
          ## --------------------------------------------------------------------
          find ${{ runner.temp }}/package -name 'testthat.Rout*' -exec cat '{}' \; || true
        shell: bash

      - name: Upload test results
        if: failure()
        uses: actions/upload-artifact@v3
        with:
          name: coverage-test-failures
          path: ${{ runner.temp }}/package
```

## Step 8: CRAN Submission

Final steps for CRAN submission [@wickham2015].

### Pre-Submission Checklist

```bash
# 1. Update version number in DESCRIPTION
# 2. Update NEWS.md with changes
# 3. Run comprehensive checks
make r

R
> # Check on multiple R versions
> devtools::check()  # Current R version

> # Check reverse dependencies (none for new package)
> devtools::revdep_check()

> # Spell check
> devtools::spell_check()

> # Check for CRAN policy compliance
> rhub::check_for_cran()

> # Windows check
> devtools::check_win_devel()
> devtools::check_win_release()

> quit()
exit

# 4. Verify all CI checks pass on GitHub
# 5. Update cran-comments.md with test results
```

### Submit to CRAN

```bash
make r

R
> # Final check
> devtools::check()

> # Submit to CRAN
> devtools::release()

> # Follow prompts - answer questions carefully
> # CRAN maintainers will review and provide feedback

> quit()
exit
```

### Post-Submission

After CRAN acceptance:

```bash
# Tag release
git tag -a v0.1.0 -m "CRAN release 0.1.0"
git push origin v0.1.0

# Create GitHub release
gh release create v0.1.0 --title "v0.1.0: Initial CRAN Release" --notes "See NEWS.md for details"
```

## Step 9: Package Website with pkgdown

Create professional package documentation website [@wickham2015].

### Setup pkgdown

```bash
make r

R
> usethis::use_pkgdown()
> pkgdown::build_site()
> quit()

exit
```

### Deploy to GitHub Pages

Create `.github/workflows/pkgdown.yml`:

```yaml
name: pkgdown

on:
  push:
    branches: [main, master]
  pull_request:
    branches: [main, master]
  release:
    types: [published]
  workflow_dispatch:

jobs:
  pkgdown:
    runs-on: ubuntu-latest
    permissions:
      contents: write
    steps:
      - uses: actions/checkout@v4

      - uses: r-lib/actions/setup-pandoc@v2

      - uses: r-lib/actions/setup-r@v2
        with:
          use-public-rspm: true

      - uses: r-lib/actions/setup-r-dependencies@v2
        with:
          extra-packages: any::pkgdown, local::.
          needs: website

      - name: Build site
        run: pkgdown::build_site_github_pages(new_process = FALSE, install = FALSE)
        shell: Rscript {0}

      - name: Deploy to GitHub pages
        if: github.event_name != 'pull_request'
        uses: JamesIves/github-pages-deploy-action@v4.4.1
        with:
          clean: false
          branch: gh-pages
          folder: docs
```

Website will be available at: `https://yourusername.github.io/penguintools/`

# Advanced Package Development

## Using Example Data

Include example datasets in your package:

```bash
make r

R
> # Create data object
> library(palmerpenguins)
> penguins_sample <- head(penguins, 20)
>
> # Save as package data
> usethis::use_data(penguins_sample)
>
> # Document the dataset
> usethis::use_r("data")

> quit()
exit
```

Document in `R/data.R`:

```r
#' Sample of Palmer Penguins data
#'
#' A subset of 20 observations from the Palmer Penguins dataset,
#' included for examples and testing.
#'
#' @format A data frame with 20 rows and 8 variables:
#' \describe{
#'   \item{species}{Penguin species (Adelie, Chinstrap, Gentoo)}
#'   \item{island}{Island in Palmer Archipelago}
#'   \item{bill_length_mm}{Bill length (millimeters)}
#'   \item{bill_depth_mm}{Bill depth (millimeters)}
#'   \item{flipper_length_mm}{Flipper length (millimeters)}
#'   \item{body_mass_g}{Body mass (grams)}
#'   \item{sex}{Penguin sex}
#'   \item{year}{Study year}
#' }
#' @source Palmer Station LTER \url{https://pallter.marine.rutgers.edu/}
#' @references
#' Horst AM, Hill AP, Gorman KB (2020). palmerpenguins: Palmer Archipelago
#' (Antarctica) penguin data. R package version 0.1.0.
"penguins_sample"
```

## S3 Methods

Implement S3 methods for clean interfaces:

```r
#' Print method for penguin model results
#'
#' @param x Object of class "penguin_model"
#' @param ... Additional arguments (unused)
#' @export
print.penguin_model <- function(x, ...) {
  cat("Penguin Allometric Model\n")
  cat("========================\n")
  cat("Formula:", deparse(x$formula), "\n")
  cat("Observations:", nrow(x$data), "\n")
  cat("R-squared:", round(x$r.squared, 3), "\n")
  invisible(x)
}
```

## Rcpp Integration

For performance-critical functions, integrate C++:

```bash
make r

R
> usethis::use_rcpp()
> quit()

exit
```

Create `src/fast_functions.cpp`:

```cpp
#include <Rcpp.h>
using namespace Rcpp;

//' Fast body mass standardization
//'
//' @param body_mass Numeric vector of body mass values
//' @return Standardized body mass (mean=0, sd=1)
//' @export
// [[Rcpp::export]]
NumericVector standardize_mass_cpp(NumericVector body_mass) {
  double mean = Rcpp::mean(body_mass);
  double sd = Rcpp::sd(body_mass);
  return (body_mass - mean) / sd;
}
```

# Best Practices Summary

Following R package development best practices [@wickham2015; @wickham2011]:

## Code Quality
✅ **Comprehensive documentation** - All functions have roxygen2 docs
✅ **Informative examples** - Every function has working examples
✅ **Input validation** - Check parameters, provide clear error messages
✅ **Consistent style** - Follow tidyverse style guide

## Testing
✅ **High test coverage** - >95% code coverage target
✅ **Edge case testing** - Empty data, NAs, extreme values
✅ **Error testing** - Invalid inputs produce informative errors
✅ **Regression tests** - Tests prevent code breakage

## Documentation
✅ **README with examples** - Clear introduction and usage
✅ **Vignettes** - Long-form tutorials for workflows
✅ **NEWS file** - Version history with changes
✅ **pkgdown website** - Professional online documentation

## CRAN Compliance
✅ **LICENSE file** - Clear open source license
✅ **DESCRIPTION complete** - All metadata filled
✅ **Multi-platform testing** - Windows, macOS, Linux
✅ **No check errors** - Clean R CMD check on all platforms

## Version Control
✅ **Git history** - Clear commit messages
✅ **GitHub integration** - Issues, pull requests, releases
✅ **CI/CD automation** - Tests run on every commit
✅ **Semantic versioning** - Clear version numbering

# Progression From Analysis to Package

## Starting from Existing Analysis

If you already have analysis code, the progression is natural [@marwick2018]:

### Month 0: Analysis Phase
```
my-project/
├── analysis/
│   ├── data/
│   └── scripts/
│       ├── 01_prepare.R
│       ├── 02_analyze.R
│       └── 03_visualize.R
```

**No package yet** - just analysis scripts.

### Month 1: Extract Reusable Functions
```
my-project/
├── analysis/
│   ├── data/
│   └── scripts/
│       ├── 01_prepare.R      # Now sources R/data_functions.R
│       ├── 02_analyze.R      # Now sources R/analysis_functions.R
│       └── 03_visualize.R    # Now sources R/plot_functions.R
├── R/
│   ├── data_functions.R      # ADD: Extracted from scripts
│   ├── analysis_functions.R  # ADD: Extracted from scripts
│   └── plot_functions.R      # ADD: Extracted from scripts
```

**Still not a package** - just better organized. Functions are in `R/` but no package infrastructure yet.

### Month 2: Add Tests
```
my-project/
├── analysis/
├── R/
└── tests/                    # ADD: Unit tests
    └── testthat/
        ├── test-data_functions.R
        ├── test-analysis_functions.R
        └── test-plot_functions.R
```

**Still not a package** - but now with quality assurance.

### Month 3: Add Package Documentation
```
my-project/
├── analysis/
├── R/
├── tests/
├── man/                      # ADD: Generated docs
│   ├── prepare_data.Rd
│   └── ...
├── vignettes/                # ADD: Tutorials
│   └── tutorial.Rmd
└── NEWS.md                   # ADD: Version history
```

**Now it's a package!** Ready for distribution.

### Month 6: CRAN Submission
```
my-project/
├── analysis/
├── R/
├── tests/
├── man/
├── vignettes/
├── NEWS.md
├── cran-comments.md         # ADD: CRAN submission notes
└── LICENSE                   # ADD: Formal license
```

**CRAN-ready package** - submit and share with world.

## The Key: No Migration Required

At every stage, you **never rebuild the project**. Just **add what you need, when you need it** [@marwick2018].

## Sharing as Blog Post

Want to share your package development journey or analysis as a blog post? ZZCOLLAB projects integrate with Quarto blogs:

```bash
# Run the blog setup script
modules/setup_symlinks.sh

# See vignette: "Blog Post Development with ZZCOLLAB"
```

# References

Marwick, B., Boettiger, C., & Mullen, L. (2018). Packaging Data Analytical Work Reproducibly Using R (and Friends). *The American Statistician*, 72(1), 80-88.

Wickham, H. (2015). *R Packages: Organize, Test, Document, and Share Your Code*. O'Reilly Media.

Wickham, H. (2011). testthat: Get Started with Testing. *The R Journal*, 3(1), 5-10.

Wilson, G., Bryan, J., Cranston, K., Kitzes, J., Nederbragt, L., & Teal, T. K. (2017). Good enough practices in scientific computing. *PLOS Computational Biology*, 13(6), e1005510.

Stodden, V., Seiler, J., & Ma, Z. (2018). An empirical analysis of journal policy effectiveness for computational reproducibility. *Proceedings of the National Academy of Sciences*, 115(11), 2584-2589.

Schmidt-Nielsen, K. (1984). *Scaling: Why Is Animal Size So Important?* Cambridge University Press.

# Additional Resources

- Wickham's R Packages: https://r-pkgs.org/
- CRAN Repository Policy: https://cran.r-project.org/web/packages/policies.html
- rOpenSci Packaging Guide: https://devguide.ropensci.org/
- Tidyverse Style Guide: https://style.tidyverse.org/
- ZZCOLLAB Testing Guide: `docs/TESTING_GUIDE.md`
- ZZCOLLAB Development Guide: `docs/DEVELOPMENT.md`
