---
title: "R Graphics in Kitty Terminal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{R Graphics in Kitty Terminal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette describes how to display R graphics when using the kitty terminal
emulator with Docker-based R environments. Unlike X11 forwarding, kitty's native
graphics protocol displays images directly in the terminal without requiring
XQuartz or X11 libraries.

### Why Kitty Graphics?

| Approach | Pros | Cons |
|----------|------|------|
| **Kitty native** | No X11 setup, lighter Docker image, works over SSH | Plots are static (save-then-view) |
| **X11 forwarding** | Interactive R graphics devices | Requires XQuartz, heavier image, latency |
| **httpgd** | Interactive, works anywhere with browser | Requires port forwarding |

## Prerequisites

- [Kitty terminal](https://sw.kovidgoyal.net/kitty/) installed on your host
- Docker with a ZZCOLLAB profile (e.g., `ubuntu_standard_analysis_vim`)
- No X11 or XQuartz required

## Method 1: Save and Display with icat

The simplest approach: save plots to files and display them using kitty's
`icat` kitten.

### Basic Workflow

```{r}
library(ggplot2)

# Create a plot
p <- ggplot(mtcars, aes(wt, mpg)) +

geom_point() +
theme_minimal()

# Save to file
ggsave("plot.png", p, width = 8, height = 6, dpi = 150)
```
```bash
# Display in kitty (run from terminal, not R)
kitty +kitten icat plot.png
```

### Helper Function

Add this to your `.Rprofile` or project's R scripts to streamline the workflow:

```{r}
#' Display a plot in kitty terminal
#'
#' Saves the plot to a temporary file and displays it using kitty's icat
#'
#' @param plot A ggplot2 object or NULL to use last_plot()
#' @param width Plot width in inches
#' @param height Plot height in inches
#' @param dpi Resolution in dots per inch
#'
#' @examples
#' library(ggplot2)
#' ggplot(mtcars, aes(wt, mpg)) + geom_point()
#' kitty_plot()
kitty_plot <- function(plot = NULL, width = 8, height = 6, dpi = 150) {
if (is.null(plot)) {
    plot <- ggplot2::last_plot()
}
if (is.null(plot)) {
    stop("No plot to display")
}

tmp <- tempfile(fileext = ".png")
ggplot2::ggsave(tmp, plot, width = width, height = height, dpi = dpi)

# Display using kitty icat
system2("kitty", c("+kitten", "icat", tmp))

invisible(plot)
}
```

Usage:

```{r}
library(ggplot2)

# Create and display in one step
ggplot(mtcars, aes(wt, mpg)) +
geom_point() +
theme_minimal()

kitty_plot()

# Or with custom dimensions
kitty_plot(width = 10, height = 8)
```

### Base R Graphics

For base R plots, save explicitly:

```{r}
#' Display base R plot in kitty
#'
#' @param expr An expression that creates a plot
#' @param width Width in inches
#' @param height Height in inches
kitty_base <- function(expr, width = 8, height = 6) {
tmp <- tempfile(fileext = ".png")
png(tmp, width = width, height = height, units = "in", res = 150)
eval(substitute(expr))
dev.off()
system2("kitty", c("+kitten", "icat", tmp))
}

# Usage
kitty_base(plot(mtcars$wt, mtcars$mpg))
kitty_base(hist(rnorm(1000)))
```

## Method 2: httpgd for Interactive Graphics

The `httpgd` package provides a web-based graphics device that works with any
browser. This gives you interactive, zoomable plots.

### Setup

```{r}
install.packages("httpgd")
```

### Usage

```{r}
library(httpgd)

# Start the graphics server
hgd()
#> httpgd server running at http://127.0.0.1:8888/live

# Now create plots - they appear in your browser
library(ggplot2)
ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +
geom_point(size = 3) +
theme_minimal()
```
Open the URL in your browser to see interactive plots.

### Docker Port Forwarding

When running R in Docker, forward the httpgd port:

```bash
# Start container with port forwarding
docker run -it -p 8888:8888 -v $(pwd):/home/rstudio/project myproject
```
```{r}
# Inside R
library(httpgd)
hgd(host = "0.0.0.0", port = 8888)

# Create plots - view at http://localhost:8888/live
```

### Makefile Target

Add to your Makefile for convenience:

```makefile
# Start R with httpgd port forwarded
r-httpgd:
	docker run --platform linux/amd64 --rm -it \
		-p 8888:8888 \
		-v $$(pwd):/home/rstudio/project \
		$(PACKAGE_NAME)
```

## Method 3: Combining Both Approaches

For maximum flexibility, use httpgd for interactive exploration and kitty_plot
for quick static views:

```{r}
library(ggplot2)
library(httpgd)

# Start httpgd for interactive work
hgd()

# Explore interactively
p <- ggplot(diamonds, aes(carat, price, color = cut)) +
geom_point(alpha = 0.1) +
scale_y_log10() +
theme_minimal()
p

# Quick static view in terminal
kitty_plot(p)
```

## Complete .Rprofile Setup

Add these functions to your `.Rprofile` for automatic availability:

```{r}
# Kitty graphics helpers (add to .Rprofile)
if (interactive() && Sys.getenv("TERM") == "xterm-kitty") {
kitty_plot <- function(plot = NULL, width = 8, height = 6, dpi = 150) {
    if (is.null(plot)) plot <- ggplot2::last_plot()
    if (is.null(plot)) stop("No plot to display")
    tmp <- tempfile(fileext = ".png")
    ggplot2::ggsave(tmp, plot, width = width, height = height, dpi = dpi)
    system2("kitty", c("+kitten", "icat", tmp))
    invisible(plot)
}

kitty_base <- function(expr, width = 8, height = 6) {
    tmp <- tempfile(fileext = ".png")
    png(tmp, width = width, height = height, units = "in", res = 150)
    eval(substitute(expr), envir = parent.frame())
    dev.off()
    system2("kitty", c("+kitten", "icat", tmp))
}

message("Kitty graphics helpers loaded: kitty_plot(), kitty_base()")
}
```

## Comparison with X11

### When to Use Kitty Native Graphics

- Working over SSH (X11 forwarding is slow)
- Lighter Docker images preferred
- Static plots are sufficient
- macOS without XQuartz installed

### When to Use X11

- Need interactive base R graphics (locator, identify)
- Real-time plot updates required
- Using packages that require X11 (some Shiny features)
- Already have XQuartz configured

### Image Size Comparison

| Profile | Size | Graphics Method |
|---------|------|-----------------|
| `ubuntu_standard_analysis_vim` | ~1.8GB | kitty/httpgd |
| `ubuntu_x11_analysis_vim` | ~2.2GB | X11 forwarding |

## Troubleshooting

### icat Command Not Found

Ensure you're running kitty (not iTerm2, Terminal.app, etc.):

```bash
echo $TERM
# Should output: xterm-kitty
```

### Plots Not Displaying

Check that kitty is the active terminal:

```bash
kitty +kitten icat --detect-support
```

### httpgd Port Already in Use

```{r}
# Use a different port
hgd(port = 9999)
```

### Docker Container Cannot Reach Host

For httpgd, bind to all interfaces:

```{r}
hgd(host = "0.0.0.0", port = 8888)
```

## See Also

- [Kitty Graphics Protocol](https://sw.kovidgoyal.net/kitty/graphics-protocol/)
- [httpgd Package](https://nx10.github.io/httpgd/)
- [ZZCOLLAB X11 Plotting Workflow](workflow-x11-plotting.html)
