---
title: "Solo Developer Workflow Guide"
author: "ZZCOLLAB Team"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Solo Developer Workflow Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# ZZCOLLAB Solo Developer Workflow Guide

This vignette demonstrates the complete workflow for solo developers using ZZCOLLAB to create reproducible research environments with professional-grade practices.

## Initial Setup (One-Time)

### 1. Install ZZCOLLAB System

```bash
# Clone and install zzcollab
git clone https://github.com/rgt47/zzcollab.git
cd zzcollab && ./install.sh

# Verify installation  
zzcollab --help && which zzcollab
```

### 2. Configuration Setup (Recommended)

ZZCOLLAB includes a configuration system to eliminate repetitive typing and set project defaults. **Set this up once and simplify all subsequent commands.**

```bash
# Initialize configuration file
zzcollab --config init

# Set your defaults (customize as needed)
zzcollab --config set team-name "rgt47"              # Your Docker Hub account
zzcollab --config set github-account "rgt47"        # Your GitHub username  
zzcollab --config set build-mode "standard"         # fast, standard, comprehensive
zzcollab --config set dotfiles-dir "~/dotfiles"     # Path to your dotfiles

# View your configuration
zzcollab --config list
```

## Project Creation

### Choose Your Development Environment

**Quick Start (Recommended):**

```bash
# Creates optimal variants automatically + GitHub repo
zzcollab -i -p myproject --github    # Creates: minimal + analysis variants
```

**Interactive Variant Selection (Power Users):**

```bash
mkdir myproject && cd myproject
zzcollab -i -p myproject             # Creates project + config.yaml
./add_variant.sh                     # Browse comprehensive variant library

# Interactive menu shows 14 variants:
# 📦 STANDARD RESEARCH ENVIRONMENTS
#  1) minimal          ~800MB  - Essential R packages only
#  2) analysis         ~1.2GB  - Tidyverse + data analysis tools  
#  3) modeling         ~1.5GB  - Machine learning with tidymodels
#  4) publishing       ~3GB    - LaTeX, Quarto, bookdown, blogdown
#  5) shiny            ~1.8GB  - Interactive web applications
#  6) shiny_verse      ~3.5GB  - Shiny with tidyverse + publishing
#
# 🔬 SPECIALIZED DOMAINS
#  7) bioinformatics   ~2GB    - Bioconductor genomics packages
#  8) geospatial       ~2.5GB  - sf, terra, leaflet mapping tools
#
# 🏔️ LIGHTWEIGHT ALPINE VARIANTS  
#  9) alpine_minimal   ~200MB  - Ultra-lightweight for CI/CD
# 10) alpine_analysis  ~400MB  - Essential analysis in tiny container
# 11) hpc_alpine       ~600MB  - High-performance parallel processing
#
# 🧪 R-HUB TESTING ENVIRONMENTS
# 12) rhub_ubuntu      ~1GB    - CRAN-compatible package testing
# 13) rhub_fedora      ~1.2GB  - Test against R-devel
# 14) rhub_windows     ~1.5GB  - Windows compatibility testing

# Select variants that match your workflow, then:
zzcollab --variants-config config.yaml --github
```

**Legacy Approach (Limited to 3 variants):**

```bash
# Traditional: Limited to shell/rstudio/verse only
zzcollab -i -p myproject -B rstudio --github    # RStudio interface
zzcollab -i -p myproject -B all --github        # All 3 legacy variants
```

### Recommended Variant Combinations by Use Case

- **Data Analysts**: `analysis` + `publishing` (tidyverse + reporting)
- **Bioinformaticians**: `bioinformatics` + `alpine_minimal` (research + CI/CD)
- **Package Developers**: `minimal` + `rhub_ubuntu` (development + testing)
- **Web App Developers**: `shiny_verse` + `alpine_minimal` (apps + deployment)
- **Academic Researchers**: `modeling` + `publishing` (analysis + manuscripts)

## Daily Development Workflow

### 1. Start Development Environment

```bash
cd myproject

# Choose your interface based on selected variants:
make docker-zsh         # Shell interface (works with any variant)
make docker-rstudio     # RStudio Server at localhost:8787 (if rstudio/shiny variants)
make docker-r           # R console only
make docker-verse       # Publishing workflow with LaTeX (if publishing variant)
```

### 2. Iterative Development (Inside Container)

**Working in the container:**

```bash
# You're now inside the Docker container with all packages pre-installed
# Your project directory is mounted at /home/analyst/project

# R package development
R
devtools::load_all()           # Load your package functions
devtools::test()              # Run tests
devtools::document()          # Generate documentation
quit()

# Analysis scripts
vim scripts/01_data_analysis.R    # Create/edit analysis
R --vanilla < scripts/01_data_analysis.R  # Run script

# Create new functions
vim R/my_functions.R          # Add R functions
vim tests/testthat/test-my_functions.R  # Write tests

# Install additional packages (will be tracked automatically)
R
install.packages("newpackage")
quit()

# Work on reports
vim analysis/report.Rmd       # Edit R Markdown report
R
rmarkdown::render("analysis/report.Rmd")  # Generate report
quit()

# Git workflow (from inside container)
git status                    # Check changes
git add .                    # Stage changes
git diff --cached            # Review staged changes
```

### 3. Exit Container and Commit

```bash
# Exit the development container
exit

# You're now back on your host system
# Validate dependencies and run final tests
make docker-check-renv-fix    # Auto-fix any dependency issues
make docker-test             # Run all tests in clean environment
make docker-render           # Ensure reports render correctly

# Git workflow - commit and push
git status                   # Check what changed
git diff                    # Review changes

git add .
git commit -m "Add data analysis with visualization

- Implement customer segmentation analysis
- Add clustering functions with tests
- Generate summary report with plots
- All tests passing, dependencies validated"

git push origin main        # Push to GitHub
```

### 4. What Happens Automatically

When you push changes to GitHub:

1. **GitHub Actions automatically**:
   - ✅ Runs R package validation
   - ✅ Executes all tests
   - ✅ Renders analysis reports
   - ✅ Detects if new packages were added

2. **If new packages detected**:
   - ✅ Rebuilds Docker image with new packages
   - ✅ Pushes updated image to Docker Hub
   - ✅ Next time you run `make docker-zsh`, you get the updated environment

### 5. Continue Development Cycle

```bash
# Start next iteration
make docker-zsh              # Continue with updated environment
# ... more development inside container ...
exit
# ... commit and push changes ...
```

## Advanced Development Patterns

### Working with Different Variants

```bash
# Switch between different environments for different tasks
make docker-zsh              # Use analysis variant for data exploration
make docker-rstudio          # Use RStudio for interactive development
make docker-verse            # Use publishing variant for report writing

# Each environment has specialized packages for its purpose
```

### Testing and Validation Workflow

```bash
# Before committing, validate your work:
make docker-test             # Run all automated tests
make docker-check            # R CMD check validation
make docker-render           # Ensure all reports render

# Fix any issues before committing
make docker-zsh
# ... fix issues inside container ...
exit
```

### Package Development Focus

```bash
# Inside container - R package development workflow
R
devtools::check()            # Full package check
devtools::build()            # Build package
devtools::install()          # Install your package
usethis::use_test("myfunction")  # Create test file
quit()

# Document and check
make docker-document         # Generate documentation
make docker-check           # Full package validation
```

## Practical Example: Penguin Bill Analysis

Let's walk through a complete example using the Palmer penguins dataset to demonstrate the iterative development workflow.

### Step 1: Create Project and Initial Analysis

```bash
# Set up the project
zzcollab -i -p penguin-analysis --github

# Start development environment
cd penguin-analysis
make docker-zsh
```

**Inside the container - Create initial analysis script:**

```bash
# Create the analysis script
vim scripts/01_penguin_exploration.R
```

**Contents of `scripts/01_penguin_exploration.R`:**

```r
#' Penguin Bill Analysis
#' Explore relationship between bill depth and log of bill length

# Load required packages
library(palmerpenguins)
library(ggplot2)
library(dplyr)

#' Create scatter plot of bill depth vs log(bill length)
#' @return ggplot object
create_bill_plot <- function() {
  penguins %>%
    filter(!is.na(bill_length_mm), !is.na(bill_depth_mm)) %>%
    ggplot(aes(x = log(bill_length_mm), y = bill_depth_mm)) +
    geom_point(aes(color = species), alpha = 0.7, size = 2) +
    labs(
      title = "Penguin Bill Depth vs Log(Bill Length)",
      x = "Log(Bill Length) (mm)",
      y = "Bill Depth (mm)",
      color = "Species"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
}

# Create and display the plot
bill_plot <- create_bill_plot()
print(bill_plot)

# Save the plot
ggsave("figures/bill_analysis.png", bill_plot, width = 8, height = 6, dpi = 300)

cat("Analysis complete! Plot saved to figures/bill_analysis.png\n")
```

**Create the function file:**

```bash
# Create R function file
vim R/penguin_functions.R
```

**Contents of `R/penguin_functions.R`:**

```r
#' Create scatter plot of bill depth vs log(bill length)
#' 
#' @param data Data frame containing penguin data (default: palmerpenguins::penguins)
#' @return ggplot object
#' @export
#' @examples
#' plot <- create_bill_plot()
#' print(plot)
create_bill_plot <- function(data = palmerpenguins::penguins) {
  if (!requireNamespace("ggplot2", quietly = TRUE)) {
    stop("ggplot2 package is required")
  }
  if (!requireNamespace("dplyr", quietly = TRUE)) {
    stop("dplyr package is required")
  }
  
  data %>%
    dplyr::filter(!is.na(bill_length_mm), !is.na(bill_depth_mm)) %>%
    ggplot2::ggplot(ggplot2::aes(x = log(bill_length_mm), y = bill_depth_mm)) +
    ggplot2::geom_point(ggplot2::aes(color = species), alpha = 0.7, size = 2) +
    ggplot2::labs(
      title = "Penguin Bill Depth vs Log(Bill Length)",
      x = "Log(Bill Length) (mm)",
      y = "Bill Depth (mm)",
      color = "Species"
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(legend.position = "bottom")
}
```

**Create tests for the function:**

```bash
# Create test file
vim tests/testthat/test-penguin_functions.R
```

**Contents of `tests/testthat/test-penguin_functions.R`:**

```r
test_that("create_bill_plot works correctly", {
  # Test with default data
  plot <- create_bill_plot()
  
  # Check that it returns a ggplot object
  expect_s3_class(plot, "ggplot")
  
  # Check plot components
  expect_equal(plot$labels$title, "Penguin Bill Depth vs Log(Bill Length)")
  expect_equal(plot$labels$x, "Log(Bill Length) (mm)")
  expect_equal(plot$labels$y, "Bill Depth (mm)")
  expect_equal(plot$labels$colour, "Species")
})

test_that("create_bill_plot handles custom data", {
  # Create test data
  test_data <- data.frame(
    bill_length_mm = c(40, 45, 50),
    bill_depth_mm = c(18, 20, 22),
    species = c("A", "B", "C")
  )
  
  plot <- create_bill_plot(test_data)
  expect_s3_class(plot, "ggplot")
})

test_that("create_bill_plot handles missing values", {
  # Create test data with NA values
  test_data <- data.frame(
    bill_length_mm = c(40, NA, 50),
    bill_depth_mm = c(18, 20, NA),
    species = c("A", "B", "C")
  )
  
  plot <- create_bill_plot(test_data)
  expect_s3_class(plot, "ggplot")
  
  # Should have only 1 point after filtering NAs
  expect_equal(nrow(plot$data), 1)
})
```

**Test and run the analysis:**

```bash
# Install required packages
R
install.packages(c("palmerpenguins", "ggplot2", "dplyr"))
quit()

# Test the function
R
devtools::load_all()
devtools::test()
quit()

# Run the analysis script
mkdir -p figures
R --vanilla < scripts/01_penguin_exploration.R

# Check the git status
git status
git add .
git diff --cached
```

### Step 2: Exit Container and First Commit

```bash
# Exit the development container
exit

# Validate dependencies and test
make docker-check-renv-fix    # Auto-add new packages to renv.lock
make docker-test             # Run tests in clean environment

# First commit and push
git add .
git commit -m "Add initial penguin bill analysis

- Create scatter plot of bill depth vs log(bill length)
- Add create_bill_plot() function with comprehensive tests
- Generate publication-quality figure
- All tests passing, dependencies tracked"

git push origin main
```

### Step 3: Continue Analysis - Add Regression Line

After the first push, continue with enhanced analysis:

```bash
# Start development environment again
make docker-zsh

# Update the analysis script
vim scripts/01_penguin_exploration.R
```

**Updated `scripts/01_penguin_exploration.R`:**

```r
#' Penguin Bill Analysis - Enhanced with Regression
#' Explore relationship between bill depth and log of bill length

# Load required packages
library(palmerpenguins)
library(ggplot2)
library(dplyr)
library(broom)

#' Create enhanced scatter plot with regression line
#' @return ggplot object
create_enhanced_bill_plot <- function() {
  penguins %>%
    filter(!is.na(bill_length_mm), !is.na(bill_depth_mm)) %>%
    ggplot(aes(x = log(bill_length_mm), y = bill_depth_mm)) +
    geom_point(aes(color = species), alpha = 0.7, size = 2) +
    geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
    labs(
      title = "Penguin Bill Depth vs Log(Bill Length) with Regression Line",
      x = "Log(Bill Length) (mm)",
      y = "Bill Depth (mm)",
      color = "Species",
      caption = "Dashed line shows linear regression fit with 95% confidence interval"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
}

#' Fit linear model for bill depth vs log(bill length)
#' @return list with model object and summary statistics
fit_bill_model <- function() {
  clean_data <- penguins %>%
    filter(!is.na(bill_length_mm), !is.na(bill_depth_mm)) %>%
    mutate(log_bill_length = log(bill_length_mm))
  
  model <- lm(bill_depth_mm ~ log_bill_length, data = clean_data)
  
  list(
    model = model,
    summary = summary(model),
    r_squared = summary(model)$r.squared,
    coefficients = tidy(model)
  )
}

# Create enhanced plot
enhanced_plot <- create_enhanced_bill_plot()
print(enhanced_plot)

# Fit regression model
model_results <- fit_bill_model()
cat("\nRegression Results:\n")
cat("R-squared:", round(model_results$r_squared, 3), "\n")
print(model_results$coefficients)

# Save outputs
ggsave("figures/bill_analysis_with_regression.png", enhanced_plot, 
       width = 8, height = 6, dpi = 300)

# Save model results
saveRDS(model_results, "results/bill_model.rds")

cat("\nEnhanced analysis complete!\n")
cat("Plot: figures/bill_analysis_with_regression.png\n")
cat("Model: results/bill_model.rds\n")
```

**Update the function file:**

```bash
vim R/penguin_functions.R
```

**Add to `R/penguin_functions.R`:**

```r
#' Create enhanced scatter plot with regression line
#' 
#' @param data Data frame containing penguin data (default: palmerpenguins::penguins)
#' @return ggplot object
#' @export
create_enhanced_bill_plot <- function(data = palmerpenguins::penguins) {
  if (!requireNamespace("ggplot2", quietly = TRUE)) {
    stop("ggplot2 package is required")
  }
  if (!requireNamespace("dplyr", quietly = TRUE)) {
    stop("dplyr package is required")
  }
  
  data %>%
    dplyr::filter(!is.na(bill_length_mm), !is.na(bill_depth_mm)) %>%
    ggplot2::ggplot(ggplot2::aes(x = log(bill_length_mm), y = bill_depth_mm)) +
    ggplot2::geom_point(ggplot2::aes(color = species), alpha = 0.7, size = 2) +
    ggplot2::geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
    ggplot2::labs(
      title = "Penguin Bill Depth vs Log(Bill Length) with Regression Line",
      x = "Log(Bill Length) (mm)",
      y = "Bill Depth (mm)",
      color = "Species",
      caption = "Dashed line shows linear regression fit with 95% confidence interval"
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(legend.position = "bottom")
}

#' Fit linear model for bill depth vs log(bill length)
#' 
#' @param data Data frame containing penguin data (default: palmerpenguins::penguins)
#' @return list with model object and summary statistics
#' @export
fit_bill_model <- function(data = palmerpenguins::penguins) {
  if (!requireNamespace("dplyr", quietly = TRUE)) {
    stop("dplyr package is required")
  }
  if (!requireNamespace("broom", quietly = TRUE)) {
    stop("broom package is required")
  }
  
  clean_data <- data %>%
    dplyr::filter(!is.na(bill_length_mm), !is.na(bill_depth_mm)) %>%
    dplyr::mutate(log_bill_length = log(bill_length_mm))
  
  model <- lm(bill_depth_mm ~ log_bill_length, data = clean_data)
  
  list(
    model = model,
    summary = summary(model),
    r_squared = summary(model)$r.squared,
    coefficients = broom::tidy(model)
  )
}
```

**Add tests for new functions:**

```bash
vim tests/testthat/test-penguin_functions.R
```

**Add to test file:**

```r
test_that("create_enhanced_bill_plot works correctly", {
  plot <- create_enhanced_bill_plot()
  
  expect_s3_class(plot, "ggplot")
  expect_equal(plot$labels$title, 
               "Penguin Bill Depth vs Log(Bill Length) with Regression Line")
  expect_true(grepl("regression", plot$labels$caption, ignore.case = TRUE))
})

test_that("fit_bill_model returns correct structure", {
  model_results <- fit_bill_model()
  
  expect_type(model_results, "list")
  expect_true("model" %in% names(model_results))
  expect_true("summary" %in% names(model_results))
  expect_true("r_squared" %in% names(model_results))
  expect_true("coefficients" %in% names(model_results))
  
  expect_s3_class(model_results$model, "lm")
  expect_type(model_results$r_squared, "double")
  expect_s3_class(model_results$coefficients, "data.frame")
})

test_that("fit_bill_model handles custom data", {
  test_data <- data.frame(
    bill_length_mm = c(40, 45, 50, 55),
    bill_depth_mm = c(18, 19, 20, 21),
    species = c("A", "B", "C", "A")
  )
  
  model_results <- fit_bill_model(test_data)
  expect_s3_class(model_results$model, "lm")
  expect_true(model_results$r_squared >= 0 && model_results$r_squared <= 1)
})
```

**Test and run the enhanced analysis:**

```bash
# Install new package
R
install.packages("broom")
quit()

# Test the new functions
R
devtools::load_all()
devtools::test()
quit()

# Create results directory and run enhanced analysis
mkdir -p results
R --vanilla < scripts/01_penguin_exploration.R

# Check what changed
git status
git diff
```

### Step 4: Second Commit with Enhancement

```bash
# Exit container
exit

# Validate enhanced analysis
make docker-check-renv-fix    # Track new broom package
make docker-test             # Ensure all tests pass

# Commit the enhancement
git add .
git commit -m "Add regression analysis to penguin bill study

- Add linear regression line to scatter plot
- Implement fit_bill_model() function with model diagnostics
- Include R-squared and coefficient estimates
- Add comprehensive tests for regression functionality  
- Save model results for reproducibility
- All tests passing, broom package added to dependencies"

git push origin main
```

### What This Example Demonstrates

1. **Complete workflow**: From initial analysis to enhanced version
2. **Professional practices**: Functions, tests, documentation
3. **Iterative development**: Build on previous work incrementally  
4. **Dependency tracking**: Automatic renv.lock updates
5. **Reproducible outputs**: Saved plots and model objects
6. **Quality assurance**: Tests validate function behavior
7. **Version control**: Clear commit messages with detailed changes

This example shows how ZZCOLLAB supports **professional data science workflows** with **minimal overhead**! 🐧📊

## Benefits of Solo Workflow

- ✅ **Reproducible**: Every development session uses identical environment
- ✅ **Isolated**: No conflicts with your host system R installation
- ✅ **Collaborative-ready**: Easy to share exact environment with others
- ✅ **Professional**: Automated testing, validation, and CI/CD
- ✅ **Flexible**: 14+ variants for different research domains
- ✅ **Lightweight options**: Alpine variants ~200MB vs standard ~1GB+
- ✅ **Automatic dependency tracking**: Never lose track of required packages
- ✅ **Version controlled**: Complete project history including environment

## Solo to Team Transition

If you later want to collaborate:

1. **Your project is already team-ready** - others can join with:
   ```bash
   git clone https://github.com/yourname/myproject.git
   cd myproject
   zzcollab -t yourname -p myproject -I analysis  # Join with analysis variant
   ```

2. **No migration needed** - the same Docker images and workflow work for teams

This workflow provides **enterprise-grade reproducibility** with **solo developer simplicity**! 🚀

## Configuration System Benefits

With configuration set up, all commands become simpler:

```bash
# Traditional verbose approach:
zzcollab -i -t rgt47 -p myproject -B analysis -S -d ~/dotfiles --github

# Config-simplified approach (identical result):
zzcollab -i -p myproject -B analysis --github

# Modern variant approach (uses config.yaml):
zzcollab -i -p myproject --github    # Creates default variants automatically
```

The configuration system eliminates repetitive typing while maintaining full flexibility for custom workflows.

## Using the R Interface

```{r, eval = FALSE}
# Load the zzcollab R package
library(zzcollab)

# Configuration-based approach (recommended)
init_config()
set_config("team_name", "myteam")
set_config("build_mode", "standard")

# Create and manage projects from R
init_project(project_name = "penguin-analysis")
status()  # Check project status
```

## Next Steps

- **Learn team collaboration**: See `vignette("workflow-team", "zzcollab")` for multi-developer workflows
- **R package development**: Explore the built-in R package structure and testing framework
- **Advanced variants**: Try specialized variants for bioinformatics, geospatial analysis, or web applications
- **CI/CD integration**: Leverage the automated testing and Docker image management