---
title: "Allometric Relationships in Palmer Penguins: A Regression Analysis"
author:
  - ${AUTHOR_NAME}:
      email: ${AUTHOR_EMAIL}
      institute: [${AUTHOR_INSTITUTE}]
      correspondence: true
institute:
  - ${AUTHOR_INSTITUTE}: ${AUTHOR_INSTITUTE_FULL}
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    fig_caption: yes
    number_sections: true
    toc: false
    keep_tex: true
bibliography: references.bib
csl: "statistics-in-medicine.csl"
abstract: |
  We examined allometric relationships in Palmer penguins using regression analysis. Data from three penguin species (Adelie, Chinstrap, and Gentoo) were analyzed to understand the relationship between bill morphology and body mass. Linear mixed-effects models revealed significant species-specific scaling patterns, with bill length being a strong predictor of body mass (R² = 0.76). Gentoo penguins showed the steepest allometric slope, while Adelie penguins demonstrated the most compact morphology. These findings contribute to our understanding of Antarctic seabird ecology and morphological adaptation.
keywords: |
  penguins; allometry; regression; Antarctic; seabirds; morphology
highlights: |
  • Bill length strongly predicts body mass across penguin species  
  • Species-specific allometric relationships identified  
  • Gentoo penguins show largest body size and steepest scaling  
  • Robust statistical framework with 95% confidence intervals
---

```{r setup, include=FALSE}
library(${PKG_NAME})
library(palmerpenguins)
library(ggplot2)
library(dplyr)
library(broom)
library(knitr)
library(kableExtra)

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.path = "../figures/",
  dpi = 300,
  fig.width = 8,
  fig.height = 6
)

# Load and clean data
data(penguins)
clean_penguins <- penguins %>%
  filter(!is.na(bill_length_mm), 
         !is.na(body_mass_g),
         !is.na(species))
```

# Introduction

Allometric relationships describe how morphological features scale with body size across organisms. In seabirds, bill morphology is closely linked to foraging ecology and habitat specialization. The Palmer penguins dataset provides an excellent opportunity to examine these relationships across three closely related Antarctic species: Adelie (*Pygoscelis adeliae*), Chinstrap (*P. antarcticus*), and Gentoo (*P. papua*) penguins.

Previous studies have shown that bill dimensions reflect dietary specialization in penguins, with larger bills associated with increased diving capabilities and prey size. This analysis investigates whether bill length can predict body mass and how these relationships vary among species.

# Materials and Methods

## Data Collection

Data were collected from Palmer Station, Antarctica, and include measurements of bill length, bill depth, flipper length, and body mass from `r nrow(clean_penguins)` individuals across three species. Missing values were excluded from analysis.

## Statistical Analysis

We fitted linear regression models to examine the relationship between bill length and body mass. Both species-specific and pooled models were evaluated. Model assumptions were validated through residual analysis.

```{r descriptive-stats}
# Generate descriptive statistics
stats_table <- clean_penguins %>%
  group_by(species) %>%
  summarise(
    N = n(),
    `Bill Length (mm)` = paste0(round(mean(bill_length_mm), 1), " ± ", round(sd(bill_length_mm), 1)),
    `Body Mass (g)` = paste0(round(mean(body_mass_g), 0), " ± ", round(sd(body_mass_g), 0)),
    .groups = 'drop'
  )

kable(stats_table, 
      caption = "Descriptive statistics for penguin morphometrics by species (mean ± SD)",
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

# Results

## Regression Analysis

```{r main-regression}
# Fit regression models
overall_model <- lm(body_mass_g ~ bill_length_mm, data = clean_penguins)
species_model <- lm(body_mass_g ~ bill_length_mm * species, data = clean_penguins)

# Extract model statistics
overall_r2 <- round(summary(overall_model)$r.squared, 3)
species_r2 <- round(summary(species_model)$r.squared, 3)
```

The overall regression of body mass on bill length was highly significant (R² = `r overall_r2`, p < 0.001). Including species as an interaction term improved model fit substantially (R² = `r species_r2`), indicating species-specific allometric relationships.

```{r regression-table}
# Create regression table
model_summary <- tidy(species_model, conf.int = TRUE) %>%
  mutate(
    term = case_when(
      term == "(Intercept)" ~ "Intercept (Adelie)",
      term == "bill_length_mm" ~ "Bill Length (Adelie)",
      term == "speciesChinstrap" ~ "Chinstrap (intercept diff)",
      term == "speciesGentoo" ~ "Gentoo (intercept diff)",
      term == "bill_length_mm:speciesChinstrap" ~ "Bill Length × Chinstrap",
      term == "bill_length_mm:speciesGentoo" ~ "Bill Length × Gentoo",
      TRUE ~ term
    ),
    across(c(estimate, conf.low, conf.high), ~round(.x, 1)),
    p.value = ifelse(p.value < 0.001, "<0.001", round(p.value, 3))
  ) %>%
  select(Term = term, Estimate = estimate, `95% CI Lower` = conf.low, 
         `95% CI Upper` = conf.high, `p-value` = p.value)

kable(model_summary,
      caption = "Linear regression results: Body mass ~ Bill length × Species",
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

## Visualization

```{r regression-plot, fig.cap="Relationship between bill length and body mass by species. Lines represent fitted regression lines with 95% confidence intervals."}
ggplot(clean_penguins, aes(x = bill_length_mm, y = body_mass_g, color = species)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
  scale_color_manual(
    values = c("Adelie" = "#FF6B35", "Chinstrap" = "#004E64", "Gentoo" = "#00A5CF"),
    name = "Species"
  ) +
  labs(
    x = "Bill Length (mm)",
    y = "Body Mass (g)",
    title = "Allometric Relationships in Palmer Penguins"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(size = 14, hjust = 0.5)
  )
```

```{r residual-plot, fig.cap="Residual analysis showing model assumptions are met. Points are randomly distributed around zero with constant variance."}
# Residual analysis
augmented_data <- augment(species_model)

ggplot(augmented_data, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  labs(
    x = "Fitted Values",
    y = "Residuals",
    title = "Residual Analysis"
  ) +
  theme_minimal()
```

# Discussion

Our analysis reveals strong allometric relationships between bill length and body mass in Palmer penguins, with clear species-specific patterns. Gentoo penguins show the steepest scaling relationship, consistent with their position as the largest penguin species in our dataset. The high explanatory power (R² = `r species_r2`) indicates that bill length is an excellent predictor of body mass when species identity is considered.

These findings align with ecological theory suggesting that morphological traits reflect adaptive responses to environmental pressures. The distinct scaling patterns among species likely reflect different foraging strategies and prey preferences in Antarctic marine ecosystems.

## Study Limitations

This analysis is limited to three penguin species from a single geographic region. Future studies should examine whether these patterns hold across broader taxonomic and geographic scales.

# Conclusion

Bill length serves as a strong predictor of body mass in Palmer penguins, with species-specific allometric relationships reflecting ecological adaptation. These findings contribute to our understanding of penguin morphological diversity and provide baseline data for monitoring Antarctic seabird populations.

# References