---
title: "Using zzcollab from R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using zzcollab from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Don't evaluate code chunks by default
)
```

# Introduction

The `zzcollab` package provides R interfaces to the zzcollab Docker-based research collaboration framework. This vignette demonstrates how to use the R functions to create, manage, and collaborate on research projects.

## Installation

First, install the zzcollab command-line tool and then load the R package:

```{r setup}
library(zzcollab)
```

## Quick Start

### For Team Leaders: Initialize a New Project

```{r init-project}
# Initialize a new research project with team collaboration
init_project(
  team_name = "myteam",
  project_name = "myproject",
  build_mode = "standard",
  dotfiles_path = "~/dotfiles"
)
```

### For Team Members: Join an Existing Project

```{r join-project}
# Join an existing project
join_project(
  team_name = "myteam",
  project_name = "myproject",
  interface = "shell",
  build_mode = "standard"
)
```

### For Individual Use: Setup a Project

```{r setup-project}
# Setup a project in the current directory
setup_project(
  build_mode = "standard",
  dotfiles_path = "~/dotfiles"
)
```

## Build Modes

The zzcollab framework supports three build modes:

- **`"fast"`**: Minimal Docker + minimal packages (fastest builds)
- **`"standard"`**: Standard Docker + standard packages (balanced, default)
- **`"comprehensive"`**: Extended Docker + full packages (kitchen sink)

```{r build-modes}
# Fast mode - minimal setup for quick development
init_project("team", "project", build_mode = "fast")

# Standard mode - balanced approach (default)
init_project("team", "project", build_mode = "standard")

# Comprehensive mode - full featured environment
init_project("team", "project", build_mode = "comprehensive")
```

## Docker Container Management

### Check Container Status

```{r status}
# Check running zzcollab containers
status()
```

### Rebuild Docker Images

```{r rebuild}
# Rebuild Docker image after dependency changes
rebuild()
```

### List Team Images

```{r team-images}
# List available team Docker images
team_images()
```

## Package Management

### Add New Packages

```{r add-package}
# Add packages to the project
add_package(c("dplyr", "ggplot2"))
```

### Sync Environment

```{r sync-env}
# Sync environment with renv.lock
sync_env()
```

## Analysis and Reporting

### Run Analysis Scripts

```{r run-script}
# Execute R script in container
run_script("analysis/my_analysis.R")
```

### Render Reports

```{r render-report}
# Render analysis reports
render_report("analysis/report.Rmd")
```

### Validate Reproducibility

```{r validate-repro}
# Check if environment is reproducible
validate_repro()
```

## Git Integration

### Check Git Status

```{r git-status}
# Check git status
git_status()
```

### Create Feature Branch

```{r create-branch}
# Create and switch to feature branch
create_branch("feature/new-analysis")
```

### Commit Changes

```{r git-commit}
# Commit changes
git_commit("Add new analysis results")
```

### Push Changes

```{r git-push}
# Push to GitHub
git_push()
```

### Create Pull Request

```{r create-pr}
# Create pull request
create_pr(
  title = "Add new analysis results",
  body = "This PR adds the results from our latest analysis."
)
```

## Getting Help

### Get Help Information

```{r help}
# Get general help
zzcollab_help()

# Get initialization help
zzcollab_help(init_help = TRUE)
```

### Get Next Steps

```{r next-steps}
# Get next steps guidance
zzcollab_next_steps()
```

## Complete Workflow Example

Here's a complete example of a typical zzcollab workflow:

```{r complete-workflow}
# 1. Initialize project (team leader)
init_project(
  team_name = "datascience",
  project_name = "covid-analysis",
  build_mode = "standard",
  dotfiles_path = "~/dotfiles"
)

# 2. Add required packages
add_package(c("tidyverse", "lubridate", "plotly"))

# 3. Create feature branch
create_branch("feature/exploratory-analysis")

# 4. Run analysis
run_script("scripts/exploratory_analysis.R")

# 5. Render report
render_report("analysis/covid_report.Rmd")

# 6. Validate reproducibility
if (validate_repro()) {
  message("✅ Environment is reproducible")
} else {
  message("❌ Environment needs attention")
}

# 7. Commit and push
git_commit("Add COVID-19 exploratory analysis")
git_push()

# 8. Create pull request
create_pr(
  title = "Add COVID-19 Exploratory Analysis",
  body = "This PR adds exploratory analysis of COVID-19 data with visualizations."
)
```

## Advanced Usage

### Custom Base Images

```{r custom-base}
# Use custom base image
setup_project(
  base_image = "myteam/myproject-base:latest",
  build_mode = "fast"
)
```

### Team Collaboration Workflow

```{r team-workflow}
# Team leader initializes
init_project("datascience", "covid-analysis", build_mode = "standard")

# Team members join
join_project("datascience", "covid-analysis", interface = "rstudio")

# Everyone can now work with the same environment
status()  # Check container status
sync_env()  # Sync with latest dependencies
```

## Best Practices

1. **Use consistent build modes** across your team
2. **Always validate reproducibility** before committing
3. **Use feature branches** for new analyses
4. **Keep your environment synchronized** with `sync_env()`
5. **Document your workflow** in your analysis scripts

## Troubleshooting

### Common Issues

If you encounter issues:

1. Check that zzcollab is installed and in your PATH
2. Verify Docker is running
3. Ensure you're in a zzcollab project directory
4. Check that all dependencies are synchronized

```{r troubleshooting}
# Check if zzcollab is available
try(zzcollab_help())

# Check Docker status
status()

# Validate environment
validate_repro()
```

For more detailed information, see the zzcollab documentation and GitHub repository.