---
title: "Palmer Penguins Bill Dimensions Analysis"
author: "Reproducible Research Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load package functions from R/ directory
devtools::load_all(here::here())
```

## Introduction

Analysis of bill length and bill depth measurements across three penguin species
from the Palmer Station Antarctica LTER.

This report demonstrates the complete data workflow:
- Raw data → Processing → Analysis → Visualization

## Data Processing

```{r data-processing}
library(dplyr)
library(ggplot2)

# Load the processed data (function from R/data_utils.R)
penguins <- load_penguin_data()

# Preview data
head(penguins)
```

## Summary Statistics

```{r summary-stats}
# Calculate summary by species (function from R/data_utils.R)
summary_stats <- summarize_by_species(penguins)
knitr::kable(summary_stats, digits = 2, caption = "Bill dimensions by species")
```

## Visualization

### Bill Dimensions by Species

```{r bill-plot, fig.width=8, fig.height=6}
# Generate plot using function from R/bill_analysis.R
bill_plot <- create_bill_plot(penguins)
print(bill_plot)

# Save plot
dir.create(here::here("analysis/figures"), showWarnings = FALSE, recursive = TRUE)
ggplot2::ggsave(here::here("analysis/figures/bill_scatter.png"),
                bill_plot, width = 8, height = 6)
```

## Summary

The analysis reveals distinct bill dimension patterns across penguin species:
- Data processed from raw measurements (Step 3)
- Functions in R/ provide reusable analysis components
- Unit tests ensure correctness
